---
title: "Text Mining Final Project"
format: html
editor: visual
---


## Introduction

**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**
**Está en el google doc, incluirla al final**

## Setup, data loading, and preprocessing

### Packages

We begin by loading all R packages required throughout this project.

```{r}
#| message: false
#| warning: false

# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS # REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS

# Core data wrangling, visualization, and string handling
library(tidyverse)
library(tidytext)
library(textdata)
library(sentimentr)
library(topicmodels)
library(quanteda)
library(quanteda.textplots)
library(textstem)
library(igraph)
library(ggraph)
library(widyr)
library(reshape2)
library(RColorBrewer)
library(wordcloud)
library(explore)
```

### Loading the data

Our corpus was downloaded from [OpenAlex](https://openalex.org/). We created a custom R script (`openalex_download.R`, included in this project folder, along with documentation) to bulk-download article data from our target journals.

Having run the script, we load the `.RData` file, containing all data structures exactly as they were returned by the OpenAlex API.

```{r}
# Load the RData file. This creates the object `articles_raw` in the environment
load("data/articles_sample.RData")
```

### Initial Inspection

We begin by examining the structure of the raw dataset: variable types, missing values, and basic summary statistics.
```{r}
#| label: inspect-structure

# Overview of all variables: types, missing values, and distributions
describe(articles_raw)
```

Not all records include abstracts — OpenAlex coverage varies by journal and year. Since our entire analysis is based on abstract text, we remove records without abstracts and restrict the corpus to our study period (2016–2025).
```{r}
#| label: filter-corpus

# Filter: keep only records with an abstract, within our study period
articles <- articles_raw |> 
  filter(!is.na(abstract)) |>
  filter(publication_year >= 2016, publication_year <= 2025)

# How many articles remain after filtering
nrow(articles)
```
```{r}
#| label: corpus-by-journal-year

# Articles per journal and year — the basic shape of our corpus
articles |>
  count(source_display_name, publication_year) |>
  pivot_wider(names_from = source_display_name, values_from = n, values_fill = 0)
```
```{r}
#| label: corpus-plot

# Visualize the distribution of abstracts over time, by journal
articles |>
  count(source_display_name, publication_year) |>
  ggplot(aes(x = publication_year, y = n, fill = source_display_name)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of abstracts per year and journal",
    x = "Publication year",
    y = "Number of abstracts",
    fill = "Journal"
  ) +
  theme_minimal()
```

The corpus is now loaded, filtered, and ready for preprocessing.

### Preprocessing

HERE WE NEED TO DO A PREPROCESSING PHASE, COMMON FOR ALL THE PARTS. OR WE CAN DO PREPROCESSING FOR EACH PART. BUT MAYBE IT IS BETTER TO DO SOMETHING COMMON FOR ALL.

## Section 3. Topic analysis [RQ1 + RQ2] 

## Section 4. WRITING STYLE AND THE RISE OF AI [RQ3] 

## Section 5. SENTIMENT TOWARDS AI IN SOCIAL SCIENCE ABSTRACTS [RQ4]

## Section 6. IF WE CAN [RQ5]

## Section 7. Conclussions and limitations

## Section 8. References