---
title: "Text Mining Final Project"
format: html
editor: visual
---

## Introduction

**Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final**

## Setup, data loading, and preprocessing [SI NO HACEMOS EL PREPROCESSING AQUÍ, QUITARLO DEL TÍTULO]

### Packages

We begin by loading all R packages required throughout this project.

```{r}
#| message: false
#| warning: false

# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS # REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS

library(tidyverse)
library(tidytext)
library(textdata)
library(sentimentr)
library(topicmodels)
library(quanteda)
library(quanteda.textplots)
library(textstem)
library(igraph)
library(ggraph)
library(widyr)
library(reshape2)
library(RColorBrewer)
library(wordcloud)
library(explore)
library(patchwork)
```

### Loading the data

Our corpus was downloaded from [OpenAlex](https://openalex.org/). We created a custom R script (`openalex_download.R`, included in this project folder, along with documentation) to bulk-download article data from our target journals.

Having run the script, we load the `.RData` file, containing all data structures exactly as they were returned by the OpenAlex API.

```{r}
# Load the RData file. This creates the object `articles_raw` in the environment
load("data/articles.RData")
```

### Initial Inspection

We begin by examining the structure of the raw dataset: variable types, missing values, and basic summary statistics.

```{r}
# Overview of all variables: types, missing values, and distributions
describe(articles_raw)
```

Not all records include abstracts (19.6% of the observations do not include it).Since our entire analysis is based on abstract text, we remove records without abstracts and restrict the corpus to our study period (2016–2025).

```{r}
# Filter: keep only records with an abstract, within our study period
articles <- articles_raw |> 
  filter(!is.na(abstract)) |>
  filter(publication_year >= 2016, publication_year <= 2025)

# Let's observe our new filtered data:
describe(articles)
```

Our new data passed from 12589 to 9484 articles. We have 44 columns from which we will later select only the relevant ones for our analysis.

We explore the distribution of articles by journal and year.

```{r}
# Articles per journal and year
articles |>
  count(source_display_name, publication_year) |>
  arrange(publication_year) |>
  pivot_wider(names_from = source_display_name, values_from = n, values_fill = 0)
```

```{r}
# # Visualize the distribution of abstracts over time, by journal
# articles |>
#   count(source_display_name, publication_year) |>
#   ggplot(aes(x = publication_year, y = n, fill = source_display_name)) +
#   geom_col(position = "dodge") +
#   labs(
#     title = "Number of abstracts per year and journal",
#     x = "Publication year",
#     y = "Number of abstracts",
#     fill = "Journal"
#   ) +
#   theme_minimal()
```

```{r}
# Merge Palgrave Communications into HSSC — same journal, renamed June 2020
# Full names with abbreviations in parentheses for clarity
articles_merged <- articles |>
  mutate(source_display_name = if_else(
    source_display_name == "Palgrave Communications",
    "Humanities and Social Sciences Communications",
    source_display_name
  )) |>
  mutate(journal_label = case_match(
    source_display_name,
    "American Political Science Review"             ~ "American Political Science Review (APSR)",
    "American Sociological Review"                  ~ "American Sociological Review (ASR)",
    "European Journal of Political Research"        ~ "European Journal of Political Research (EJPR)",
    "European Sociological Review"                  ~ "European Sociological Review (ESR)",
    "Nature Human Behaviour"                        ~ "Nature Human Behaviour (NHB)",
    "Humanities and Social Sciences Communications" ~ "Humanities and Social Sciences Communications (HSSC)"
  ))

# Color palette — keys must match journal_label values exactly
journal_colors <- c(
  "American Political Science Review (APSR)"      = "#2C3E6B",
  "American Sociological Review (ASR)"            = "#C0392B",
  "European Journal of Political Research (EJPR)" = "#8E44AD",
  "European Sociological Review (ESR)"            = "#E67E22",
  "Nature Human Behaviour (NHB)"                  = "#27AE60",
  "Humanities and Social Sciences Communications (HSSC)" = "#2980B9"
)

# Plot 1: one facet per journal, bars by year
# Each panel has its own y-axis scale (scales = "free_y") so within-journal
# trends are visible even for journals with very different volumes
p_facets <- articles_merged |>
  count(journal_label, publication_year) |>
  ggplot(aes(x = publication_year, y = n, fill = journal_label)) +
  geom_col(width = 0.7) +
  facet_wrap(~ journal_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(breaks = seq(2016, 2025, by = 3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Publication year", y = "Number of abstracts") +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text       = element_text(size = 9, face = "bold", color = "#1a1a2e"),
    axis.text.x      = element_text(size = 7, color = "#4a4a6a"),
    axis.text.y      = element_text(size = 7, color = "#4a4a6a"),
    axis.title       = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e8e8f0", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", color = NA)
  )

# Plot 2: thin stacked horizontal bar showing corpus composition
# No labels — colors alone carry the journal identity, matched to the facets above
p_bar <- articles_merged |>
  count(journal_label) |>
  mutate(
    pct = n / sum(n),
    journal_label = fct_reorder(journal_label, n)
  ) |>
  ggplot(aes(x = pct, y = 1, fill = journal_label)) +
  geom_col(orientation = "y", width = 0.6) +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(x = "Share of total abstracts", y = NULL) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.y     = element_blank(),
    axis.text.x     = element_text(size = 7, color = "#4a4a6a"),
    axis.title.x    = element_text(size = 9, color = "#4a4a6a", margin = margin(t = 6)),
    panel.grid      = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Stack the two plots vertically with patchwork
# heights = c(10, 1) keeps the proportion bar slim relative to the facets
p_facets / p_bar +
  plot_layout(heights = c(10, 1)) +
  plot_annotation(
    title    = "Number of abstracts per year and journal",
    subtitle = "Palgrave Communications merged with HSSC (same journal, renamed June 2020)",
    theme    = theme(
      plot.title      = element_text(size = 14, face = "bold", color = "#1a1a2e",
                                     family = "serif", margin = margin(b = 4)),
      plot.subtitle   = element_text(size = 9, color = "#4a4a6a",
                                     family = "serif", margin = margin(b = 12)),
      plot.caption    = element_text(size = 7, color = "#7a7a9a",
                                     family = "serif", hjust = 0,
                                     margin = margin(t = 10)),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )
```

The six journals in our corpus do not have a balanced share of articles: HSSC represents over 60% of the corpus, while there is relative balanced among the other five journals. Another trend to take into account is related to the temporal distribution: while the first four journals show a relatively balanced distribution of articles by year, the bottom two ---the journals in the *Nature portfolio*--- have exponentially increased the amount of publications in recent years. While this is something we must take into account, we will not exclude any journal from the corpus just for the sake of balance, since they are all reputed and influential journals, excluding one (for example HSSC), would affect the representativeness of the corpus. This is a trade off we are willing to accept, even if many articles are published in the same journal, they are still articles being published in top social science journals, and we should take them into account when answering our research questions. **\[REVISAR ESTE PÁRRAFO; ME PARECE REPETITVO\]**

## Section 3. Topic analysis \[RQ1 + RQ2\]

### Preprocessing

#### Tokenization

We start by keeping the relevant columns: id, publication year, journal, and abstract.

```{r}
# We tokenize at the word level, keeping key metadata columns as identifiers:
tidy_abstracts_raw <- articles |>
  select(id, publication_year, source_display_name, abstract)
```

Then tokenize at the word level:

```{r}
# Tokenize at the word level. Each row will be one token:
tidy_abstracts_raw <- tidy_abstracts_raw |>
  unnest_tokens(
    output = word,
    input  = abstract,
    token  = "words"   # default: lowercase, strips punctuation
  )

# To check that it worked:
tidy_abstracts_raw
```

After tokenizing words, we need to remove all those stopwords that are not useful for analysis, since they have very little informational value.

#### Filtering standard and domain-specific stopwords

We use `stop_words`, a dataframe that includes stopwords from three lexicons (SMART, Snowball, Onix), provided by library `tidy_text` as our standard stopword filter.

```{r}
# Filter stopwords using tidytext `stop_words`
tidy_abstracts <- tidy_abstracts_raw |>
  anti_join(stop_words, by = "word")

tidy_abstracts
```

Academic abstracts also contain high-frequency words that carry no topical meaning and would pollute TF-IDF rankings and LDA topics if left in.

Using Claude **[QUÉ VERSIÓN]** to identify and create a list of domain-specif stopwords ---we attached a .csv document of our data after filtering, asking to create a list of little informational value words that should be removed---. We manually reviewed this list, removing words we thought had informational value and adding some other little informative verbs and words such as "abstract", "enables", "responses", "based", "effect", "effects", which were not included in the list. We also added "palgrave", which we found to have a high influence in some of the metrics if not excluded.

We define two tiers for filtering: *[IGUAL QUITABA ESTO; NO ENTIENDO EN QUÉ ESTAMOS HACIENDO DOS TIERS]*
(a) Generic academic boilerplate: words common to all scientific writing
(b) Method/corpus noise: reporting verbs, general adjectives and adverbs without sentiment component, numbers as words, single characters, statistics and document-structure wording, and encoding artefacts

```{r}
custom_stopwords <- tibble(word = c(

  # Academic words: what the paper is / what it does, not what it's about
  "study", "studies",
  "paper", "abstract", "article", "articles",
  "research", "researchers", "scholars",
  "findings", "finding",
  "objective", "purpose", "aims", "aim", "aimed",
  "literature", "academic", "scientific", "interdisciplinary",
  "published", "corpus",

  # Methodological words: how it was done, not the topic
  "results", "result",
  "analysis", "analyses", "analyzed", "analyze", "analysed", "analyzing",
  "methods", "method", "methodology",
  "approach", "approaches",
  "framework", "frameworks",
  "review", "reviews", "reviewed",
  "data", "dataset", "collected", "collection",
  "survey", "questionnaire",
  "interviews", "respondents", "participants",
  "experiment", "experimental",
  "quantitative", "qualitative",
  "empirical", "theoretical",
  "systematic", "mixed",
  "conducted", "designed", "selected",
  "modeling", "mediating", "moderating",
  "panel", "equation", "indicators",
  "assessment", "evaluation", "index",
  "association", "comparative", "temporal",
  "sem", "ci",                              # statistical abbreviations
  "thematic", "conceptual",
  "techniques", "categories",
  "hypothesis",

  # Reporting verbs: "This paper X...", carry no topical signal
  "reveal", "revealed", "reveals",
  "show", "showed", "shown",
  "suggest", "suggested", "suggests", "suggesting",
  "indicate", "indicated", "indicates",
  "demonstrate", "demonstrated", "demonstrates",
  "highlight", "highlighted", "highlights", "highlighting",
  "argue", "argues",
  "examine", "examined", "examines", "examining",
  "explore", "explored", "explores", "exploring",
  "enables", "responses",
  "investigate", "investigated", "investigates",
  "discuss", "discussed", "discusses",
  "propose", "proposed", "proposes",
  "develop", "developed", "develops",
  "apply", "applied", "applies",
  "identify", "identified", "identifies",
  "include", "included", "includes",
  "consider", "considered", "considers",
  "focus", "focused", "focuses", "focusing",
  "address", "addressed", "addresses", "addressing",
  "assess", "assessed", "assesses",
  "measure", "measured", "measures",
  "report", "reported", "reports",
  "find", "found",
  "compare", "compared", "compares",
  "predict", "predicted", "predicts",
  "observe", "observed", "observes",
  "provide", "provided", "provides", "providing",
  "offer", "offers", "offering",
  "contribute", "contributes", "contributing",
  "employ", "employed", "employs", "employing",
  "utilize", "utilized", "utilizes", "utilizing",
  "draw", "draws", "drawing", "drawn",
  "determine", "determined", "determines",
  "serve", "serves",
  "explain", "explained", "explains",
  "understand",
  "remains", "remain",
  "highlighting", "drawing", "focusing", "addressing",
  "plays", "driven", "shaped", "shaping",
  "influenced", "influencing", "influences",
  "analyzes", "analysed",
  "contributes", "based",

  # Generic structural nouns (no topical value) 
  "role", "roles",
  "relationship", "relationships",
  "understanding",
  "implications", "implication",
  "outcomes", "outcome",
  "perspective", "perspectives",
  "factors", "factor",
  "patterns", "pattern",
  "context", "contexts",
  "aspects", "aspect",
  "type", "types",
  "form", "forms",
  "series", "body",
  "subject", "account",
  "extent", "basis",
  "complexity",
  "phenomenon",
  "topic", "topics",
  "field", "fields",
  "term", "terms",
  "characteristics",
  "scale",
  "degree",
  "elements", "element",
  "dimensions", "dimension",
  "mechanisms", "mechanism",
  "structure", "structures",
  "process", "processes",
  "concept", "concepts",
  "theories", "theory",      
  "themes",
  "relations",
  "functions",
  "responses",
  "words",
  "texts",
  "sources",
  "strategies",              
  "approaches",
  "activities",
  "tasks", "task",
  "questions", "question",
  "settings",
  "core",
  "scope",
  "views",
  "purpose", "effect", "effects",

  # Generic adjectives and adverbs
  # NOTE: adjectives with sentiment value (innovative, robust, limited,
  # critical, comprehensive, substantial, practical, crucial, essential,
  # valuable, major, effective, strong, diverse) are deliberately EXCLUDED
  # from this list so they are preserved for sentiment analysis.
  "specific", "specifically",
  "central", "primary", "main",
  "existing", "current", "recent",
  "broader",
  "initial", "established",
  "independent", "integrated",
  "increasingly", "primarily",
  "significant", "significantly",
  "including",
  "similar",
  "multiple",
  "total",
  "external", "internal",
  "single",
  "common",
  "distinct",
  "relevant",
  "real",
  "short",
  "direct",
  "related",
  "selected",
  "designed",
  "conducted",
  "technical",
  "growing",
  "past",
  "contemporary",
  "key",                    
  "top",
  "secondary",
  "structured",
  "comparative",
  "temporal",
  "underlying",
  "systematic",
  "frequency",

  # Connectors that survive standard stopword removal 
  "additionally", "furthermore", "however", "thus",
  "therefore", "whereas", "despite", "although",
  "i.e", "e.g", "ie", "eg", "et", "al",
  "due",
  "addition",
  "contrast",
  "finally",
  "notably",
  "broadly",

  # Numbers and years tokenized as words [yo lo quitaría]
  # "1", "2", "3", "4", "5", "10", "19", "95",
  # "2016", "2017", "2018", "2019", "2020",
  # "2021", "2022", "2023", "2024", "2025",

  # Encoding artefacts and OpenAlex noise 
  "â", "ã", "ï", "œ",
  "pre", "multi",             # prefix fragments tokenized as standalone words
  
  # Other:
  "palgrave"
))

tidy_abstracts <- tidy_abstracts |>
  anti_join(custom_stopwords, by = "word")

tidy_abstracts
```

This is the prompt we used to get this list:

*You are a text mining expert working on an academic research project in social sciences.*

*I am sharing a CSV file containing word frequencies from a corpus of social science journal abstracts, after standard English stopwords have already been removed using the tidytext stop_words lexicon in R.*

*Your task is to identify and generate a list of domain-specific stopwords that should be removed from this corpus before topic modelling (LDA) and TF-IDF analysis. These are words that appear at high frequency but carry no thematic or discriminative value for social science research —such as academic boilerplate verbs ("suggest", "examine", "findings"), generic methodological terms ("survey", "quantitative", "framework"), and encoding artefacts.*

*Critical constraint: do NOT include in the stopword list any word that carries sentiment meaning and could be useful for sentiment analysis. This includes words with positive or negative valence in standard sentiment lexicons (AFINN, Bing, NRC), such as evaluative adjectives ("innovative", "robust", "limited"), directional verbs ("improve", "achieve"), and explicit sentiment adverbs ("positively", "negatively"). These words must be preserved in the corpus.*

*Return a clean, categorised R tibble with the final list of domain-specific stopwords to remove, with comments explaining each category.*

There are still some words we do not need for the analysis: pure numbers, tokens under 3 characters, any remaining numeric strings. So we filter them:

[NOTA DE MIGUEL] **YO DIRÍA QUE NO DEBEMOS EXCLUIR WORDS UNDER 3 CHARACTERS, NEITHER PURE NUMBERS, ¿COULDN'T THOSE HAVE MEANING? POR EJEMPLO: "SEX", "BOT", "JOB", "IQ", "EU", "US", "VR", "1945", "1978", "2020", SE ME OCURREN A BOTE PRONTO.**

```{r}
# tidy_abstracts <- tidy_abstracts |>
#   filter(
#     !str_detect(word, "^[0-9]+$"),   # pure digits
#     !str_detect(word, "^[0-9]"),     # starts with a digit (e.g. "2023")
#     str_length(word) >= 3            # drop "ai", "er", etc. (will keep "ai" intentionally — see below)
#   )
```

The 3-character filter above drops "ai", but "ai" is substantively crucial for our RQ2. We add it back by filtering \>= 2 chars only for that token.

Simpler approach: lower the threshold to 2 but re-apply the custom list. Here we just relax to nchar \>= 2, since our custom list already handles noise.

```{r}
# tidy_abstracts <- tidy_abstracts_raw |>
#   anti_join(stop_words, by = "word") |>
#   anti_join(custom_stopwords, by = "word") |>
#   filter(
#     !str_detect(word, "^[0-9]"),
#     str_length(word) >= 2
#   )
```

#### Summary and visualizations of the clean corpus

```{r}
# Total tokens after preprocessing
nrow(tidy_abstracts)

# Unique vocabulary size
n_distinct(tidy_abstracts$word)

# Tokens per journal
tidy_abstracts |>
  count(source_display_name, name = "n_tokens")

# Top 20 most frequent words overall — sanity check
tidy_abstracts |>
  count(word, sort = TRUE) |>
  slice_head(n = 20)
```

```{r}
# Visualize top words
tidy_abstracts |>
  count(word, sort = TRUE) |>
  slice_head(n = 25) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word)) +
  geom_col(fill = "#2C3E6B", width = 0.7) +
  geom_text(aes(label = scales::comma(n)), 
            hjust = -0.15, size = 3, color = "#2C3E6B", family = "serif") +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = scales::comma
  ) +
  labs(
    title = "Most frequent terms after preprocessing",
    subtitle = "Top 25 tokens across all abstracts (2016–2025)",
    x = "Token frequency",
    y = NULL,
    caption = "Source: Nature Human Behaviour & HSSC — OpenAlex corpus"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title       = element_text(size = 14, face = "bold", color = "#1a1a2e",
                                    margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 10, color = "#4a4a6a",
                                    margin = margin(b = 12)),
    plot.caption     = element_text(size = 7, color = "#7a7a9a", hjust = 0,
                                    margin = margin(t = 10)),
    axis.text.y      = element_text(size = 10, color = "#1a1a2e"),
    axis.text.x      = element_text(size = 8,  color = "#4a4a6a"),
    axis.title.x     = element_text(size = 9,  color = "#4a4a6a",
                                    margin = margin(t = 8)),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#e8e8f0", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.background    = element_rect(fill = "white", color = NA),
    plot.margin        = margin(20, 25, 15, 15)
  )
```

The top 25 terms paint a coherent picture of the corpus. The dominance of *social*, *development*, *impact*, and *economic* confirms these journals prioritise research on societal outcomes and economic processes. The cluster of *health*, *policy*, and *public* points to a significant amount of public health and governance research. The public health topic may come from the pandemic years (2020-2021) that covers our time period of research.

The prominence of *China* and *Chinese* (both in the top 15) is notable and likely reflects the journals' broad international scope combined with the scale of Chinese academic output over the study period. [**NO SÉ SI DIRÍA ESTO; QUIZÁ DIRÍA QUE REFLEJA EL HECHO DE QUE CHINA SE ESTÁ CONVIERTIENDO EN UNO DE LOS PRINICPALES ACTORES A NIVEL MUNDIAL*]

The *education*–*learning*–*students* cluster suggests that educational research is one of the most represented subfields, while *digital*, *information*, and *media* signal a sustained interest in technology and communication topics throughout the decade, which has relevance for our research questions about AI.

Finally, the presence of positive and influence among the top 25 reveals that sentiment-bearing vocabulary is surviving preprocessing as intended.


### Analysis [RQ1 + RQ2]

In order to address RQ1 and RQ2, we will use three complementary techniques: TF-IDF to identify distinctive vocabulary by year (and also for the period before and after the public release of ChatGPT); word frequency analysis using a custom set of AI-related words to measure the growth of AI-related terminology by year; and LDA topic modelling.

#### Distinctive vocabulary by year and period (TF-IDF)

We use TF-IDF to identify the most distinctive words of each year and of each period (2016–2022, before Chat GPT, and 2023–2025, after it's public release), treating each time unit as a document.

##### TF-IDF by year

```{r}
# Count how many times each word appears in each year
words_by_year <- tidy_abstracts |>
  count(publication_year, word, sort = TRUE)

# Compute TF-IDF treating each year as a "document"
tfidf_by_year <- words_by_year |>
  bind_tf_idf(word, publication_year, n) |>
  arrange(desc(tf_idf))

# To check it worked (hard to read, we address that in the next chunk)
tfidf_by_year
```

```{r}
#| fig-width: 8
#| fig-height: 10

# Top 10 most distinctive words per year (to make our tf-idf results readable)
tfidf_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 10) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, word, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive words per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

- A bote pronto no sé muy bien como intepretarlo (sobre todo los primeros años). Hay cosas muy llamativas e interesantes: covid, pandemic, etc. a partir de 2020 (), la envan vayendo en favor de otras palabras con el paso del tiempo), la entrada de "ai" en 2023 (lo cual parece confirmar nuestras hipótesis). Al menos nos deja ver dos momentos de ruptura (pandemia y chatgpt).
- Busco en  google "400th shakespeare" y resulta que ese año fue el 400 aniversario de la muerte de Shakespeare (eso explica también que haya otras palabras relacionadas con shakepeare). El fracking sé que era un temón del momento porque en 2019 (un poco tarde ya, yo hice un trabajo de relaciones internacionales sobre el fracking).
- Pregunta: ¿tendrán que ver "gun" con school shootings? ¿Habría mucho debate en torno a las armas en EEUU?


##### TF-IDF by period (before and after ChatGPT)

```{r}
# Create a period variable: pre-ChatGPT (2016–2022) vs post-ChatGPT (2023–2025)
words_by_period <- tidy_abstracts |>
  mutate(period = if_else(publication_year <= 2022, "2016–2022", "2023–2025")) |>
  count(period, word, sort = TRUE)

# Compute TF-IDF treating each period as a "document"
tfidf_by_period <- words_by_period |>
  bind_tf_idf(word, period, n) |>
  arrange(desc(tf_idf))

# To check it worked (hard to read, we address that in the next chunk)
tfidf_by_period
```
```{r}
# Top 15 most distinctive words per period (to make our tf-idf results readable)
tfidf_by_period |>
  group_by(period) |>
  slice_max(tf_idf, n = 15) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, period)) |>
  ggplot(aes(tf_idf, word, fill = period)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ period, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Most distinctive words: pre- vs. post-ChatGPT era",
    subtitle = "TF-IDF treating each period as a single document",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```
- Clara diferencia, many AI related terms appear in the top 15: "chagpt", "llms", "gpt". 
- Also cryptocurrency words: "blockchain", "cryptocurrency"-
- También metaverse ("vr", "metaverse").
- Deberíamos buscar las que no sabemos qué son.
- Sería interesante buscar también por qué está el número 10.1093

¿¿¿TIENE SENTIDO USAR TF-IDF ENTRE DOS PERIODOS ÚNICAMENTE???

**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?** **¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**
**¿HACEMOS TAMBIÉN UN TF-IDF DE BIGRAMS? PODRÍA ARROJAR RESULTADOS INTERESANTES. ¿Y OTROS NGRAMS?**

#### AI as an emerging topic using an AI-related lexicon

In this subsection we operationalise AI as a topic through a keyword lexicon built from the works of Liu, Shapira, and Yue (2021) and Ding, Lawson, and Shapira (2025). These works — which fully cover our study period — explore bibliometric search strategies for identifying AI-related publications, providing validated lexicons that are useful for our purposes. We built three sets from these sources: unigrams, bigrams, and trigrams.

##### Building the AI lexicon

We manually excluded a significant number of n-grams from these lexicons, prioritising precision over recall. That is, we accept that our lexicon will capture fewer AI-related abstracts, but we have higher confidence that each match is capturing a genuine AI-related article. We excluded terms like "random forest" or "decision tree" because they refer to statistical methods often used in social science and not necessarily related to AI; to keep our lexicon easily manageable, we also excluded terms that are too technical and unlikely to appear in social science journal abstracts; and we excluded ambiguous terms, such as "ai" which Liu et al. (2021, p. 3171) reject for being an abbreviation with up to 164 meanings — and a hit ratio of just 17.8% in their analysis. This approach is more than sufficient to capture broad trends and answer our research questions.

In addition, we included extra unigrams, bigrams, and trigrams to capture named generative AI models and providers absent from both sources, such as GPT-4, Gemini, Claude, LLaMA, and DeepSeek, which Ding et al. deliberately excluded due to noise concerns at large corpus scales but which are unambiguous at our scope of approximately 9,500 abstracts.

```{r}
# AI unigram lexicon
ai_lexicon_unigrams <- c(
  # Liu et al. (2021)
  "autoencoder",
  "perceptron",
  "neurocomputing",
  # Ding et al. (2025)
  "chatgpt",
  "llm",
  # Custom
  "deepseek",
  "openai"
)

# AI bigram lexicon
ai_lexicon_bigrams <- c(
  # Liu et al. (2021)
  "artificial intelligence",
  "neural network",
  "machine learning",
  "deep learning",
  "reinforcement learning",
  "learning algorithm",
  "backpropagation learning",
  "backpropagation algorithm",
  "machine intelligence",
  "computational intelligence",
  "neural model",
  "deep network",
  # Ding et al. (2025)
  "generative ai",
  "language model",
  "large language model",
  "pretrained transformer",
  "ai-generated content",
  "openai gpt",
  # Custom
  "claude sonnet",
  "claude opus",
  "claude haiku",
  "google gemini",
  "gemini pro",
  "gemini ultra",
  "microsoft copilot",
  "github copilot",
  "mistral ai",
  "meta llama",
  "llama 2",
  "llama 3",
  "stable diffusion",
  "gpt 4o",
  "gpt 4",
  "gpt 3",
  "gpt 2",
  "gpt 1",
  "dall e",
  # Custom
  "agentic ai"
)

# AI trigram lexicon
ai_lexicon_trigrams <- c(
  # Ding et al. (2025)
  "large language model",
  "generative language model",
  "generative large language",
  "generative pretrained transformer",
  "generative artificial intelligence",
  "artificial intelligence generated",
  "openai large language",
  # Custom
  "openai language model",
  "ai language model",
  "ai generated content"
)
```

##### Tokenizing abstracts at each n-gram level

Since a unigram, bigram, and trigram lexicon, we need to tokenize the abstracts separately for each, and then match against the corresponding lexicon. After that, we combine all matches in single tibble.

```{r}
# Source abstracts. We start from articles each time since n-gram tokenization
# needs the original text
abstracts_base <- articles |>
  select(id, publication_year, abstract)

# Unigram matches: tokenize at word level and match against unigram lexicon
ai_hits_unigrams <- abstracts_base |>
  unnest_tokens(word, abstract, token = "words") |>
  filter(word %in% ai_lexicon_unigrams) |>
  rename(term = word) |>
  mutate(ngram_level = "unigram")

# Bigram matches: tokenize at bigram level and match against bigram lexicon
ai_hits_bigrams <- abstracts_base |>
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  filter(bigram %in% ai_lexicon_bigrams) |>
  rename(term = bigram) |>
  mutate(ngram_level = "bigram")

# Trigram matches — tokenize at trigram level and match against trigram lexicon
ai_hits_trigrams <- abstracts_base |>
  unnest_tokens(trigram, abstract, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  filter(trigram %in% ai_lexicon_trigrams) |>
  rename(term = trigram) |>
  mutate(ngram_level = "trigram")

# Combine all matches into a single dataframe
ai_hits <- bind_rows(ai_hits_unigrams, ai_hits_bigrams, ai_hits_trigrams)
ai_hits
```

##### AI terms frequency over time

```{r}
# Count unique articles containing at least one AI related term per year
ai_articles_by_year <- ai_hits |>
  distinct(id, publication_year) |>
  count(publication_year, name = "articles_with_ai")

# Total number of abstracts per year
abstracts_by_year <- articles |>
  filter(!is.na(abstract), publication_year >= 2016, publication_year <= 2025) |>
  count(publication_year, name = "total_abstracts")

# Compute proportion
ai_proportion <- ai_articles_by_year |>
  left_join(abstracts_by_year, by = "publication_year") |>
  mutate(proportion_ai_articles = articles_with_ai / total_abstracts)



library(scales) # Needed for percent_format()

# Find the maximum proportion to place the annotation dynamically
max_prop <- max(ai_proportion$proportion_ai_articles, na.rm = TRUE)

ggplot(ai_proportion, aes(x = publication_year, y = proportion_ai_articles)) +
  # Use geom_col for a bar plot instead of geom_line
  geom_col(fill = "#2C3E6B", width = 0.7) +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  labs(
    title = "AI-related terminology in social science abstracts",
    subtitle = "Proportion of articles containing at least one AI term, by year",
    x = "Publication year",
    y = "Proportion of articles (%)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title         = element_text(size = 14, face = "bold", color = "#1a1a2e"),
    plot.subtitle      = element_text(size = 10, color = "#4a4a6a"),
    axis.text          = element_text(color = "#4a4a6a"),
    panel.grid.major.x = element_blank(), # Removes vertical grid lines for cleaner bars
    panel.grid.minor.x = element_blank()
  )
```
- Cambio muy claro, en 2017 y 2018 prácticamente nada. En 2019 sube al 1%, el año en que se hace público chatgpt pasa hasta el 4%, y tres años después un 9% de los abstracts continen términos relacionados con IA. (Recordemos que estamos empleando un léxico conservador, seguramente hay más menciones y no estamos capturando todo, pero el patrón es evidente). 
- En relación con nuestra RQ: influencia innegable de la IA.


#### Which terms are driving the trend?

We observe a clear trend, let's explore which terms are driving it.

```{r}
#| fig-width: 6
#| fig-height: 8
# Top AI terms per year by absolute article count

ai_hits |>
  # Keep only unique combinations of article, year, and term
  distinct(id, publication_year, term) |>
  # Count the absolute number of articles per term per year
  count(publication_year, term, name = "article_count") |>
  group_by(publication_year) |>
  slice_max(article_count, n = 5, with_ties = FALSE) |>
  # Assign a rank (1 to 5) to the top terms
  mutate(rank = row_number()) |> 
  ungroup() |>
  # Add 0 value to missing ranks so every year has exactly ranks 1 through 5
  # and all bars have the same width
  complete(publication_year, rank = 1:5, fill = list(article_count = 0)) |>
  # Create invisible unique names for the blank bars using spaces:
  mutate(
    term = ifelse(is.na(term), strrep(" ", rank), term),
    term = reorder_within(term, -rank, publication_year) 
  ) |>
  
  # Plot it:
  ggplot(aes(x = article_count, y = term, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE, fill = "#2C3E6B") +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Top AI-related terms per year",
    subtitle = "Absolute number of articles containing each term",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text         = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.y        = element_text(size = 9, color = "#4a4a6a"),
    axis.text.x        = element_text(color = "#4a4a6a"),
    panel.grid.major.y = element_blank() 
  )
```
- Before 2020 not even 5 AI related words from our lexicon.
- Artificial intelligence and machine learning the main drivers since they appear in 2020.
- Worth mentioning: the  appearance of chatgpt from 2023 onwards. Also "generative ai in 2024. May point to an increased (and likely) interest in those topics after the raise of generative llms.


*RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ*


## Section 4. WRITING STYLE AND THE RISE OF AI \[RQ3\]

## Section 5. SENTIMENT TOWARDS AI IN SOCIAL SCIENCE ABSTRACTS \[RQ4\]

## Section 6. IF WE CAN \[RQ5\]

## Section 7. Conclussions and limitations

## Section 8. References
