---
title: "Text Mining Final Project"
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 4
    toc-title: "Table of Contents"
    number-sections: true
    embed-resources: true
    df-print: paged
editor: visual
---

## Section 1. Introduction

**Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final**

## Section 2. Setup, and data loading

### Packages

We begin by loading all R packages required throughout this project.

```{r}
#| message: false
#| warning: false

# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS # REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS

library(tidyverse)
library(tidytext)
library(textdata)
library(sentimentr)
library(topicmodels)
library(quanteda)
library(quanteda.textplots)
library(textstem)
library(igraph)
library(ggraph)
library(widyr)
library(reshape2)
library(RColorBrewer)
library(wordcloud)
library(explore)
library(patchwork)
library(scales)
```

### Loading the data

Our corpus was downloaded from [OpenAlex](https://openalex.org/). We created a custom R script (`openalex_download.R`, included in this project folder, along with documentation) to bulk-download article data from our target journals.

Having run the script, we load the `.RData` file, containing all data structures exactly as they were returned by the OpenAlex API.

```{r}
# Load the RData file. This creates the object `articles_raw` in the environment
load("data/articles.RData")
```

### Initial Inspection

We begin by examining the structure of the raw dataset: variable types, missing values, and basic summary statistics.

```{r}
# Overview of all variables: types, missing values, and distributions
describe(articles_raw)
```

Not all records include abstracts (19.6% of the observations do not include it).Since our entire analysis is based on abstract text, we remove records without abstracts or publication year,and restrict the corpus to our study period (2016–2025). Another relevant column is `publication_year`, we see there are no missing values, so we do not need to filter any observations out.

```{r}
# Filter: keep only records with an abstract, within our study period
articles <- articles_raw |> 
  filter(!is.na(abstract)) |>
  filter(publication_year >= 2016, publication_year <= 2025)

# Let's observe our new filtered data:
describe(articles)
```

Removing missing abstracts we went from 12589 to 9484 articles. We have 44 columns from which we will later select only the relevant ones for our analysis.

We explore the distribution of articles by journal and year.

```{r}
#| fig-width: 7
#| fig-height: 5.5
# Merge Palgrave Communications into HSSC — same journal, renamed June 2020
# Full names with abbreviations in parentheses for clarity
articles_merged <- articles |>
  mutate(source_display_name = if_else(
    source_display_name == "Palgrave Communications",
    "Humanities and Social Sciences Communications",
    source_display_name
  )) |>
  mutate(journal_label = case_match(
    source_display_name,
    "American Political Science Review"             ~ "American Political Science Review (APSR)",
    "American Sociological Review"                  ~ "American Sociological Review (ASR)",
    "European Journal of Political Research"        ~ "European Journal of Political Research (EJPR)",
    "European Sociological Review"                  ~ "European Sociological Review (ESR)",
    "Nature Human Behaviour"                        ~ "Nature Human Behaviour (NHB)",
    "Humanities and Social Sciences Communications" ~ "Humanities and Social Sciences Communications (HSSC)"
  ))

# Color palette — keys must match journal_label values exactly
journal_colors <- c(
  "American Political Science Review (APSR)"             = "#2C3E6B",
  "American Sociological Review (ASR)"                   = "#C0392B",
  "European Journal of Political Research (EJPR)"        = "#8E44AD",
  "European Sociological Review (ESR)"                   = "#E67E22",
  "Nature Human Behaviour (NHB)"                         = "#27AE60",
  "Humanities and Social Sciences Communications (HSSC)" = "#2980B9"
)

# Plot 1: Total abstracts per year (Macro-level overview)
# Colored in a neutral gray to differentiate from journal-specific data
p_total <- articles_merged |>
  count(publication_year) |>
  ggplot(aes(x = publication_year, y = n)) +
  geom_col(fill = "#7F8C8D", width = 0.7) +
  scale_x_continuous(breaks = seq(2016, 2025, by = 3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "Total abstracts", title = "Total abstracts by year") +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title       = element_text(size = 10, face = "bold", color = "#1a1a2e"),
    axis.text.x      = element_text(size = 7, color = "#4a4a6a"),
    axis.text.y      = element_text(size = 7, color = "#4a4a6a"),
    axis.title       = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e8e8f0", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", color = NA)
  )

# Plot 2: one facet per journal, bars by year
# Each panel has its own y-axis scale (scales = "free_y") so within-journal
# trends are visible even for journals with very different volumes
p_facets <- articles_merged |>
  count(journal_label, publication_year) |>
  ggplot(aes(x = publication_year, y = n, fill = journal_label)) +
  geom_col(width = 0.7) +
  facet_wrap(~ journal_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(breaks = seq(2016, 2025, by = 3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "Number of abstracts") +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text       = element_text(size = 8, face = "bold", color = "#1a1a2e",
                                    hjust = 0), # hjust = 0 aligns to the left
    axis.text.x      = element_text(size = 7, color = "#4a4a6a"),
    axis.text.y      = element_text(size = 7, color = "#4a4a6a"),
    axis.title       = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e8e8f0", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", color = NA)
  )

# Plot 3: thin stacked horizontal bar showing corpus composition
# No labels — colors alone carry the journal identity, matched to the facets above
p_bar <- articles_merged |>
  count(journal_label) |>
  mutate(
    pct = n / sum(n),
    journal_label = fct_reorder(journal_label, n)
  ) |>
  ggplot(aes(x = pct, y = 1, fill = journal_label)) +
  geom_col(orientation = "y", width = 0.6) +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(x = "Share of total abstracts", y = NULL) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.y      = element_blank(),
    axis.text.x      = element_text(size = 7, color = "#4a4a6a"),
    axis.title.x     = element_text(size = 9, color = "#4a4a6a", margin = margin(t = 6)),
    panel.grid       = element_blank(),
    plot.background  = element_rect(fill = "white", color = NA)
  )

# Stack the three plots vertically with patchwork
# heights = c(3, 10, 1) structures the visual hierarchy: Total trend -> Journal specific -> Proportion breakdown
p_total / p_facets / p_bar +
  plot_layout(heights = c(3, 10, 1)) +
  plot_annotation(
    title    = "Number of abstracts per year and journal",
    subtitle = "Palgrave Communications merged with HSSC (same journal, renamed June 2020)",
    theme    = theme(
      plot.title      = element_text(size = 14, face = "bold", color = "#1a1a2e",
                                     family = "serif", margin = margin(b = 4)),
      plot.subtitle   = element_text(size = 9, color = "#4a4a6a",
                                     family = "serif", margin = margin(b = 12)),
      plot.caption    = element_text(size = 7, color = "#7a7a9a",
                                     family = "serif", hjust = 0,
                                     margin = margin(t = 10)),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )
```

The six journals in our corpus do not represent a balanced share of articles: HSSC accounts for more than 60% of the articles, while the other five journals are relatively balanced between them. We also observe a temporal imbalance: with publication greatly increasing after 2022. Both imbalances are driven by the same factor: the great increase of publications in Journals part of the Nature Portfolio, particularly HSSC. This imbalance should be kept in mind when interpreting the results, but we will not exclude any journal from the corpus just to improve balance. We are interested in trends in social science abstracts in leading journals ---not in inter-journal comparison---, this imbalance does not affect our research questions, and removing a journal would make the corpus less representative by arbitrarily excluding thousands of publications in leading journals.

## Section 3. Topic analysis \[RQ1 + RQ2\]

### 3.1 Preprocessing setup

The techniques in this section use different tokenization and filtering strategies. We will define all shared elements here ---stopwords and filtered unigram corpus---, and explain which subsections use which.

We work with two filtering tools:

1.  **Standard stopwords**: using `stop_words` from `tidytext`. This will be applied in all techinques.

2.  **Domain-specific stopwords**: a list of words that provide no relevant information in our context. This tool (`domain_stopwords`) is fully applied in unigram analyses (both TF-IDF and LDA), a subset of it (`ngram_stopwrods`) is used in n-gram analyses ---for more details on why, see section 3.1.2---.

#### Defining domain-specific stopwords

We used *Claude Sonnet 4.6* to create an initial list of domain-specific words without informative value (see full prompt in Appendix I), then manually reviewed it, removing words we considered did have informational value, and adding others we found needed filtering while working on the analysis.

```{r}
# Domain-specific stopwords, organised by category.
# The category column allows subsetting for different filtering strategies.
domain_stopwords <- tibble(
  word = c(
    
    # 1. Academic boilerplate verbs
    # Verbs used generically across all abstracts regardless of topic.
    # Includes base forms, 3rd-person singular, past/passive, and gerund inflections.
    "examine", "investigate", "explore", "suggest", "find", "show", "demonstrate", "analyze",
    "analyse", "assess", "evaluate", "discuss", "argue", "propose", "address", "consider",
    "focus", "aim", "seek", "test", "draw", "provide", "offer", "highlight",
    "reveal", "indicate", "identify", "contribute", "develop", "present", "describe", "explain",
    "understand", "improve", "enhance", "employ", "apply", "use", "utilize", "utilise",
    "adopt", "extend", "advance", "inform", "situate", "engage", "establish", "illustrate",
    "document", "note", "observe", "report", "review", "compare", "measure", "estimate",
    "predict", "capture", "synthesize", "synthesise", "integrate", "combine", "incorporate", "build",
    "construct", "formulate", "frame", "conceptualize", "conceptualise", "operationalize", "operationalise", "uncover",
    "shed", "map", "trace", "track", "model", "simulate", "validate", "verify",
    "confirm", "challenge", "question", "reflect", "emphasize", "emphasise", "stress", "acknowledge",
    "recognize", "recognise", "account", "imply", "examines", "investigates", "explores", "suggests",
    "finds", "shows", "demonstrates", "analyses", "analyzes", "assesses", "evaluates", "discusses",
    "argues", "proposes", "addresses", "considers", "focuses", "aims", "seeks", "draws",
    "provides", "offers", "highlights", "reveals", "indicates", "identifies", "contributes", "develops",
    "presents", "describes", "explains", "improves", "enhances", "employs", "applies", "adopts",
    "extends", "informs", "engages", "establishes", "illustrates", "documents", "notes", "observes",
    "reports", "compares", "measures", "estimates", "predicts", "captures", "builds", "constructs",
    "integrates", "combines", "incorporates", "uncovers", "sheds", "maps", "traces", "tracks",
    "models", "validates", "confirms", "challenges", "questions", "reflects", "emphasizes", "emphasises",
    "stresses", "acknowledges", "recognizes", "recognises", "accounts", "implies", "examined", "investigated",
    "explored", "suggested", "found", "shown", "demonstrated", "analyzed", "analysed", "assessed",
    "evaluated", "discussed", "argued", "proposed", "addressed", "considered", "focused", "aimed",
    "sought", "drawn", "provided", "offered", "highlighted", "revealed", "indicated", "identified",
    "contributed", "developed", "presented", "described", "explained", "improved", "enhanced", "employed",
    "applied", "adopted", "extended", "informed", "engaged", "established", "illustrated", "documented",
    "noted", "observed", "reported", "compared", "measured", "estimated", "predicted", "captured",
    "built", "constructed", "integrated", "combined", "incorporated", "uncovered", "mapped", "tracked",
    "modeled", "modelled", "validated", "confirmed", "challenged", "reflected", "emphasized", "emphasised",
    "stressed", "acknowledged", "recognized", "recognised", "examining", "investigating", "exploring", "suggesting",
    "finding", "showing", "demonstrating", "analyzing", "analysing", "assessing", "evaluating", "discussing",
    "arguing", "proposing", "addressing", "considering", "focusing", "aiming", "seeking", "drawing",
    "providing", "offering", "highlighting", "revealing", "indicating", "identifying", "contributing", "developing",
    "presenting", "describing", "explaining", "understanding", "improving", "enhancing", "employing", "applying",
    "adopting", "extending", "informing", "engaging", "establishing", "illustrating", "documenting", "noting",
    "observing", "reporting", "comparing", "measuring", "estimating", "predicting", "capturing", "building",
    "constructing", "integrating", "combining", "incorporating", "uncovering", "shedding", "mapping", "tracing",
    "tracking", "modeling", "modelling", "validating", "confirming", "challenging", "questioning", "reflecting",
    "emphasizing", "emphasising", "stressing", "acknowledging", "recognizing", "recognising",
    
    # 2. Generic methodological terms
    # Words describing research designs, data types, statistical procedures,
    # and analytical approaches ubiquitous across all social science.
    "data", "dataset", "datasets", "survey", "surveys", "questionnaire", "questionnaires", "interview",
    "interviews", "observation", "observations", "fieldwork", "experiment", "experiments", "experimental", "sample",
    "samples", "sampling", "subsample", "subsamples", "respondent", "respondents", "participant", "participants",
    "informant", "informants", "census", "panel", "panels", "cohort", "cohorts", "wave",
    "waves", "method", "methods", "methodology", "methodological", "approach", "approaches", "framework",
    "frameworks", "design", "designs", "procedure", "procedures", "protocol", "protocols", "technique",
    "techniques", "instrument", "instruments", "paradigm", "paradigms", "quantitative", "statistical", "regression",
    "regressions", "multivariate", "bivariate", "univariate", "logistic", "probit", "ols", "coefficient",
    "coefficients", "parameter", "parameters", "estimation", "estimator", "estimators", "specification", "specifications",
    "robustness", "validity", "reliability", "endogeneity", "heterogeneity", "heteroskedasticity", "autocorrelation", "collinearity",
    "multicollinearity", "moderation", "mediation", "mediating", "moderating", "confound", "confounding", "confounder",
    "instrumental", "iv", "equation", "equations", "variance", "covariance", "correlation", "correlations",
    "significance", "p-value", "confidence", "interval", "intervals", "standard", "deviation", "deviations",
    "mean", "median", "mode", "distribution", "distributions", "skewness", "kurtosis", "outlier",
    "outliers", "imputation", "weighting", "weights", "weight", "index", "indices", "indicator",
    "indicators", "score", "scores", "scale", "scales", "latent", "manifest", "loading",
    "loadings", "eigenvalue", "eigenvalues", "goodness", "aic", "bic", "rmsea", "cfi",
    "tli", "srmr", "anova", "manova", "ancova", "bootstrapping", "bootstrap", "simulation",
    "simulations", "bayesian", "frequentist", "likelihood", "maximum", "posterior", "credible", "hypothesis",
    "null", "alternative", "threshold", "lag", "lags", "lagged", "fixed", "random",
    "causal", "causality", "qualitative", "grounded", "thematic", "coding",
    "codes", "code", "theme", "themes", "category", "categories", "saturation", "triangulation",
    "reflexivity", "positionality", "intersubjectivity", "hermeneutic", "phenomenological", "phenomenology", "interpretive", "constructivist",
    "analysis", "analytic", "analytical", "longitudinal", "cross-sectional", "multilevel", "multivariable", "hierarchical",
    "meta-analytic", "scoping", "realist", "synthesis", "syntheses", "mixed", "measurement", "variable",
    "variables", "dependent", "independent", "control", "controls", "controlled", "controlling", "proxy",
    "proxies", "check", "checks", "sensitivity", "validation", "replication", "comparative",
    
    # 3. Abstract structure boilerplate
    # Organisational scaffolding often used in academic abstracts:
    # document labels, rhetorical-move markers, literature-positioning words,
    # linking tokens, and non-discriminative evaluative fillers.
    "study", "studies", "paper", "papers", "article", "articles", "research", "work",
    "works", "thesis", "dissertation", "chapter", "chapters", "book", "books", "manuscript",
    "manuscripts", "introduction", "background", "motivation", "objective", "objectives", "purpose", "purposes",
    "goal", "goals", "rationale", "overview", "summary", "abstract", "conclusion", "conclusions",
    "discussion", "implication", "implications", "contribution", "contributions", "limitation", "limitations", "recommendation",
    "recommendations", "direction", "directions", "findings", "result", "results", "outcome", "outcomes",
    "literature", "prior", "previous", "existing", "recent", "current", "past", "earlier",
    "contemporary", "traditional", "conventional", "growing", "emerging", "nascent", "based", "using",
    "used", "given", "among", "across", "within", "via", "through", "including",
    "particularly", "specifically", "notably", "especially", "primarily", "mainly", "largely", "generally",
    "broadly", "widely", "commonly", "typically", "often", "frequently", "usually", "rarely",
    "however", "therefore", "thus", "hence", "whereas", "whereby", "therein", "furthermore",
    "moreover", "additionally", "nevertheless", "nonetheless", "accordingly", "consequently", "subsequently", "previously",
    "finally", "overall", "together", "concluded", "concludes", "derives", "derived", "conducted",
    "collected", "published", "carried", "undertaken", "outlined", "important", "significant", "key",
    "main", "major", "primary", "central", "core", "fundamental", "essential", "critical",
    "crucial", "vital", "notable", "relevant", "valuable", "meaningful", "useful", "effective",
    "novel", "innovative", "original", "unique", "comprehensive", "systematic", "rigorous", "robust",
    "substantial", "strong", "potential", "possible", "likely", "apparent", "clear", "direct",
    "indirect", "further", "additional", "various", "diverse", "multiple", "different", "similar",
    "comparable", "specific", "particular", "general", "broad", "wide", "limited", "due",
    "despite", "although", "though", "while", "whether", "both", "either", "neither",
    
    # 4. Generic social science concepts
    # Terms so broadly used in social science that they provide no discriminative
    # topical signal: generic agents, process/mechanism words, structural terms,
    # effect/outcome nouns, generic descriptors, and temporal/spatial terms.
    "society", "societies", "social", "individual", "individuals", "group", "groups", "population",
    "populations", "people", "person", "persons", "human", "humans", "actor", "actors",
    "agent", "agents", "member", "members", "community", "communities", "public", "citizen",
    "citizens", "subject", "subjects", "user", "users", "stakeholder", "stakeholders", "researcher",
    "researchers", "author", "authors", "scholar", "scholars", "behavior", "behaviour", "behaviors",
    "behaviours", "attitude", "attitudes", "perception", "perceptions", "belief", "beliefs", "preference",
    "preferences", "value", "values", "norm", "norms", "practice", "practices", "experience",
    "experiences", "interaction", "interactions", "process", "processes", "mechanism", "mechanisms", "dynamic",
    "dynamics", "pattern", "patterns", "relationship", "relationships", "relation", "relations", "association",
    "associations", "link", "links", "connection", "connections", "nexus", "interplay", "structure",
    "structures", "system", "systems", "network", "networks", "institution", "institutions", "organization",
    "organizations", "organisation", "organisations", "context", "contexts", "condition", "conditions", "environment",
    "environments", "setting", "settings", "domain", "domains", "sphere", "field", "fields",
    "sector", "sectors", "dimension", "dimensions", "aspect", "aspects", "element", "elements",
    "component", "components", "feature", "features", "factor", "factors", "effect", "effects",
    "impact", "impacts", "influence", "influences", "consequence", "consequences", "change", "changes",
    "difference", "differences", "variation", "variations", "increase", "increases", "decrease", "decreases",
    "reduction", "reductions", "improvement", "improvements", "development", "developments", "growth", "decline",
    "shift", "shifts", "transition", "transitions", "transformation", "transformations", "role", "roles",
    "level", "levels", "degree", "degrees", "extent", "scope", "range", "type",
    "types", "form", "forms", "kind", "kinds", "way", "ways", "modes",
    "nature", "perspective", "perspectives", "view", "views", "lens", "time", "period",
    "periods", "stage", "stages", "phase", "phases", "era", "eras", "decade",
    "decades", "year", "years", "month", "months", "day", "days", "long",
    "short", "term", "terms", "run", "world", "global", "international", "national",
    "local", "regional", "urban", "rural", "cross", "domestic", "foreign", "western",
    "eastern", "northern", "southern",
    
    # 5. Journal specific and publication boilerplate
    # Words introduced by journal formatting, editorial conventions, copyright
    # notices, metadata fragments, and abstract-template boilerplate specific
    # to these six journals.
    "copyright", "rights", "reserved", "license", "licensed", "licence", "licenced", "attribution",
    "commons", "creative", "reproduced", "reproduction", "permission", "permissions", "prohibited", "doi",
    "http", "https", "www", "org", "com", "net", "edu", "vol",
    "volume", "issue", "number", "pp", "page", "pages", "journal", "springer",
    "oxford", "cambridge", "wiley", "elsevier", "taylor", "francis", "routledge", "sage",
    "plos", "arxiv", "preprint", "open", "access", "publication", "publications", "publish",
    "publisher", "publishing", "editor", "editors", "editorial", "peer", "reviewed", "accepted",
    "submission", "submitted", "revised", "revision", "correspondence", "email", "orcid", "affiliation",
    "affiliations", "department", "university", "universities", "keywords", "keyword", "jel",
    "classification", "palgrave", "collection", "published", "cambridge.org", "doi.org", "10.1093",
    "s41599", "10.1057",
    
    # 6. Encoding artefacts and noise
    # Tokens from PDF/HTML extraction, encoding issues, OCR, or tokenisation:
    # Latin abbreviations, single characters, URL fragments, Greek letter tokens,
    # ligature remnants, and other non-linguistic noise.
    "i.e", "e.g", "et", "al", "etc", "ibid", "cf", "vs",
    "viz", "nb", "fig", "figs", "tab", "tabs", "eq", "eqs",
    "pg", "n.d", "n.a", "na", "nd", "a", "b", "c",
    "d", "e", "f", "g", "h", "i", "j", "k",
    "l", "m", "n", "o", "p", "q", "r", "s",
    "t", "u", "v", "w", "x", "y", "z", "oe",
    "ae", "fi", "ff", "ffi", "fl", "ffl", "html", "pdf",
    "xml", "json", "url", "uri", "beta", "alpha", "gamma", "delta",
    "sigma", "omega", "lambda", "mu", "nu", "rho", "phi", "chi", "07",
    "eta", "kappa", "tau", "theta", "epsilon", "zeta", "table", "appendix",
    "supplementary", "supplement", "support", "supporting", "information", "available", "request",
    
    # 7. High-frequency function words
    # Discourse connectives, hedging adverbs, academic filler adjectives,
    # numbers-as-words, and filler nouns that standard lists (snowball, SMART,
    # onix) miss but which carry no thematic signal in this corpus.
    "alternatively", "conversely", "meanwhile", "simultaneously", "otherwise", "namely", "predominantly", "chiefly",
    "mostly", "jointly", "collectively", "respectively", "comparatively", "relatively", "apparently", "seemingly",
    "arguably", "potentially", "possibly", "unlikely", "presumably", "ostensibly", "admittedly", "importantly",
    "interestingly", "surprisingly", "crucially", "significantly", "substantially", "considerably", "markedly", "increasingly",
    "positively", "negatively", "directly", "indirectly", "highly", "strongly", "effectively", "consistently",
    "robustly", "empirically", "theoretically", "conceptually", "practically", "systematically", "comprehensively", "rigorously",
    "explicitly", "implicitly", "dominant", "prominent", "salient", "unprecedented", "thorough", "considerable",
    "efficient", "successful", "increasing", "rising", "expanding", "deepening", "widening", "ongoing",
    "continuing", "persistent", "enduring", "modest", "marginal", "partial", "complex", "complicated",
    "nuanced", "multifaceted", "multidimensional", "interdisciplinary", "heterogeneous", "homogeneous", "distinct", "varied",
    "consistent", "inconsistent", "contradictory", "complementary", "broader", "deeper", "wider", "stronger",
    "weaker", "higher", "lower", "greater", "lesser", "larger", "smaller", "one",
    "two", "three", "four", "five", "six", "seven", "eight", "nine",
    "ten", "first", "second", "third", "fourth", "fifth", "half", "quarter",
    "percent", "percentage", "proportion", "share", "ratio", "rate", "insight", "insights",
    "understandings", "knowledge", "advancement", "progress", "step", "steps", "effort", "efforts",
    "attempt", "attempts", "need", "needs", "opportunity", "opportunities", "gap", "gaps",
    "concern", "concerns", "issues", "debate", "debates", "discussions", "attention", "lack",
    "absence", "presence", "existence"
  ),
  
  # Defining a category column  allows us to easily subset later
  category = c(
    # 1. Academic boilerplate verbs
    rep("boilerplate_verb", 310),
    # 2. Generic methodological terms
    rep("methodological_term", 215),
    # 3. Abstract structure boilerplate
    rep("abstract_boilerplate", 176),
    # 4. Generic social science concepts
    rep("generic_social_science", 209),
    # 5. Journal specific and publication boilerplate
    rep("publication_boilerplate", 80),
    # 6. Encoding artefacts and noise
    rep("encoding_artefact", 88),
    # 7. High-frequency function words
    rep("domain_function_word", 147)
  )
)
```

#### N-gram stopwords (subset for bigram/trigram pipelines)

For bigram and trigram analyses, we apply just a subset of `domain_stopwords`. Single words like "social" or "people" do not provide relevant information in our current context, but form meaningful n-grams ("social media" or "Roma people"), we should thus keep them. But we do filter terms that just produce noise: publication metadata, abstract specific words, or journal self-references.

```{r}
# Subset of domain stopwords need for n-gram filtering. Since we defined
# categories, we can easily choose the relevant subset
ngram_stopwords <- domain_stopwords |>
  filter(category %in% c("boilerplate_verb",
                         "abstract_boilerplate",
                         "publication_boilerplate",
                         "encoding_artefact"))
```

In addition, we include a filter for recent years, which appear frequently in abstracts, with no relevant meaning (often pointing to recent events or publication years). For example:

```{r}
# Print an abstract with a recent year:
articles |>
     filter(str_detect(abstract, "2021")) |>
     slice(1) |>
     pull(abstract)
```

We do not filter all numbers out, since other numbers could be relevant ---referring to historic events, or other meaningful values---.

```{r}
# Recent years as noise tokens
year_noise <- tibble(word = as.character(2015:2025),
                     category = "year_noise")
ngram_stopwords <- bind_rows(ngram_stopwords, year_noise)
```

#### Select relevant columns

We select the relevant columns for this section: id, publication year, journal, and abstract.

```{r}
# Select relevant columns:
tidy_abstracts_raw <- articles |>
  select(id, publication_year, source_display_name, abstract)
```

#### Filtered unigram corpus (shared by unigram TF-IDF and LDA)

Since both the unigram TF-IDF (section 3.2.1) and the LDA (section 3.2.3) use the same tokenization and filtering, we perform it here, to avoid repetitions.

We start by tokenizing at the word level:

```{r}
# Tokenize at word level, keeping metadata columns
tidy_abstracts <- articles |>
  unnest_tokens(word, abstract, token = "words")
```

Then filter standard stopwords, the full set of domain-specific stopwords, and recent years.

```{r}
tidy_abstracts <- tidy_abstracts |> 
  # Standard stopwords
  anti_join(stop_words, by = "word") |>
  # Full domain stopwords
  anti_join(domain_stopwords, by = "word") |>
  # Recent years
  filter(!word %in% as.character(2015:2025))

tidy_abstracts
```


### 3.2 Analysis \[RQ1 + RQ2\]

In order to address RQ1 and RQ2, we will use three complementary techniques: TF-IDF to identify distinctive vocabulary by year (and also for the period before and after the public release of ChatGPT); word frequency analysis using a custom set of AI-related words to measure the growth of AI-related terminology by year; and LDA topic modelling.

#### 3.2.1 Distinctive vocabulary by year (TF-IDF)

We use TF-IDF to identify the most distinctive terms of each year, treating each year as a single document. We apply this at three levels: unigram, bigram, and trigram level.

##### Unigram TF-IDF by year

```{r}
# Count how many times each word appears in each year
words_by_year <- tidy_abstracts |>
  count(publication_year, word, sort = TRUE)

# Compute TF-IDF treating each year as a "document"
tfidf_by_year <- words_by_year |>
  bind_tf_idf(word, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_by_year
```

```{r}
#| fig-width: 8
#| fig-height: 12

# Top 8 most distinctive words per year
tfidf_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, word, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive words per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

Focusing on our research questions, TF-IDF reveals two main findings. First, terms related to the COVID-19 pandemic got some of the highest scores from 2020 onwards, reflecting the influence the pandemic had in social science research. Secondly, and in relation with RQ2, "AI" enters the top 8 in 2023 (coinciding with the rise of generative AI models), and has the highest value in 2025, overtaking "covid". Outside of these, TF-IDF values are notably low, and several high-ranking terms are driven by single abstracts repeating a term, or the publishing of a thematic collection, particularly in the first few years, where we have less articles ---see, for instance, the first two context windows in appendix III---. This is an expected limitation given the relatively small size of our corpus.


##### Bigram TF-IDF by year

For bigrams we tokenize from the raw abstracts, we separate bigrams into their component words, filter standard stopwords from each position, and reunite them before computing TF-IDF. As discussed above, we do not filter by the full domain-specific stopwords list here, we use the subset with just abstract structure terms and metadata noise.

```{r}
# Tokenize into bigrams from raw abstracts
bigrams_by_year <- articles |>
  select(id, publication_year, abstract) |>
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  # Separate into two columns to filter stopwords per position
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    !word1 %in% ngram_stopwords$word,
    !word2 %in% ngram_stopwords$word
  ) |>
  # Reunite into a single bigram column
  unite(bigram, word1, word2, sep = " ") |>
  # Count by year
  count(publication_year, bigram, sort = TRUE)

bigrams_by_year
```

```{r}
# Compute TF-IDF on bigrams, treating each year as a "document"
tfidf_bigrams_by_year <- bigrams_by_year |>
  bind_tf_idf(bigram, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_bigrams_by_year
```

```{r}
#| fig-width: 9
#| fig-height: 12

# Top 8 most distinctive bigrams per year
tfidf_bigrams_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(bigram = reorder_within(bigram, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, bigram, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive bigrams per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

The bigram TF-IDF provides more easily interpretable results, and confirms our main observations from the unigram TF-IDF results. COVID-19 pandemic related-terms show the highest values. Beyond the pandemic, affective polarization enters the top 8 values in 2019, probably due to the rise of far right parties in western democracies. From 2023 onwards climate and sustainability words (particularly "carbon emissions") break into the top 8, likely indicating renewed attention to the climate crisis after the pandemic years. Regarding our RQ2, no terms directly related to AI appear, but we see indirectly related terms such as digital economy and digital transformation from 2023 to 2025.

##### Trigram TF-IDF by year

We perform TF-IDF on trigrams using the same workflow as with bigrams.

```{r}
# Same workflow for trigrams: tokenize, separate, filter, reunite
trigrams_by_year <- articles |>
  select(id, publication_year, abstract) |>
  unnest_tokens(trigram, abstract, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    !word3 %in% stop_words$word,
    !word1 %in% ngram_stopwords$word,
    !word2 %in% ngram_stopwords$word,
    !word3 %in% ngram_stopwords$word
  ) |>
  unite(trigram, word1, word2, word3, sep = " ") |>
  count(publication_year, trigram, sort = TRUE)

trigrams_by_year
```

```{r}
# Compute TF-IDF on trigrams
tfidf_trigrams_by_year <- trigrams_by_year |>
  bind_tf_idf(trigram, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_trigrams_by_year
```

```{r}
#| fig-width: 10
#| fig-height: 12

# Top 8 most distinctive trigrams per year
tfidf_trigrams_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(trigram = reorder_within(trigram, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, trigram, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive trigrams per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 7)
  )
```

Trigram TF-IDF confirms the pattern in the previous results. The COVID-19 pandemic appears as the most distinctive topic after 2020 (we see new terms, like "sars cov 2" in 2022). Climate-related terms are still among the most distinctive in the years after the pandemic, suggesting a renewed attention to the topic. Focusing on RQ2, "artificial intelligence ai" enters the top 8 in 2024 and reaches the top 4 in 2025.


#### 3.2.2 AI as an emerging topic using an AI-related lexicon

In this subsection we operationalise AI as a topic through a keyword lexicon built from the works of Liu, Shapira, and Yue (2021) and Ding, Lawson, and Shapira (2025). These works ---which fully cover our study period--- explore bibliometric search strategies for identifying AI-related publications, the lexicons they build are well thought out and validated by their research, and can prove useful for our purposes. We built three lexicons from these sources: a unigram, a bigram, and a trigram lexicon.

##### Building the AI lexicon

We manually excluded a significant number of n-grams from these lexicons, prioritising precision over recall. That is, we accept that our lexicon will capture fewer AI-related abstracts, but we have higher confidence that each match is capturing a genuine AI-related article. We excluded terms like "random forest" or "decision tree" because they refer to statistical methods often used in social science and not necessarily related to AI; to keep our lexicon easily manageable, we also excluded terms that are too technical and unlikely to appear in social science journal abstracts; and we excluded ambiguous terms ---while Liu et al. (2021, p. 3171) exclude "AI" for being an abbreviation with up to 164 meanings, we will keep it. The authors report it being a problem because it stands for terms present in science and medicine, like "Adequate Intake, Adaptive Iteration, Aridity Index, Artificial Insemination, Active Ingredient, Avian Influenza, Aromatase Inhibitor and Associative Ionization", given the nature of our corpus, restricted to social science journals, this is not a problem for us---.

In addition, we included extra unigrams, bigrams, and trigrams to capture named generative AI models and providers absent from both sources, such as GPT-4, Gemini, Claude, or LLama ---while the names by themselves could lead to ambiguity (the zodiac sign, Claude Lévi-Strauss, the animal, etc.), including the version in bigrams solves this ambiguity---.

While not perfect, this approach is more than sufficient to capture broad trends necessary to answer our research questions.

```{r}
# AI unigram lexicon
ai_lexicon_unigrams <- c(
  # Liu et al. (2021)
  "autoencoder",
  "perceptron",
  "neurocomputing",
  # Ding et al. (2025)
  "chatgpt",
  "llm",
  # Custom
  "ai",
  "deepseek",
  "openai"
)

# AI bigram lexicon
ai_lexicon_bigrams <- c(
  # Liu et al. (2021)
  "artificial intelligence",
  "neural network",
  "machine learning",
  "deep learning",
  "reinforcement learning",
  "learning algorithm",
  "backpropagation learning",
  "backpropagation algorithm",
  "machine intelligence",
  "computational intelligence",
  "neural model",
  "deep network",
  # Ding et al. (2025)
  "generative ai",
  "language model",
  "pretrained transformer",
  "ai-generated content",
  "openai gpt",
  # Custom
  "claude sonnet",
  "claude opus",
  "claude haiku",
  "google gemini",
  "gemini pro",
  "gemini ultra",
  "microsoft copilot",
  "github copilot",
  "mistral ai",
  "meta llama",
  "llama 2",
  "llama 3",
  "stable diffusion",
  "gpt 5",
  "gpt 4",
  "gpt 3",
  "gpt 2",
  "gpt 1",
  "dall e",
  # Custom
  "agentic ai"
)

# AI trigram lexicon
ai_lexicon_trigrams <- c(
  # Ding et al. (2025)
  "large language model",
  "generative language model",
  "generative large language",
  "generative pretrained transformer",
  "generative artificial intelligence",
  "artificial intelligence generated",
  "openai large language",
  # Custom
  "openai language model",
  "ai language model",
  "ai generated content"
)
```

##### Tokenizing abstracts at each n-gram level

Since we have a unigram, bigram, and trigram AI lexicon, we need to tokenize the abstracts separately for each, and then match against the corresponding lexicon.

```{r}
# Source abstracts. We start from articles each time since n-gram tokenization
# needs the original text
abstracts_base <- articles |>
  select(id, publication_year, abstract)

# Unigram matches: tokenize at word level and match against unigram lexicon
ai_hits_unigrams <- abstracts_base |>
  unnest_tokens(word, abstract, token = "words") |>
  filter(word %in% ai_lexicon_unigrams) |>
  rename(term = word) |>
  mutate(ngram_level = "unigram")

# Bigram matches: tokenize at bigram level and match against bigram lexicon
ai_hits_bigrams <- abstracts_base |>
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  filter(bigram %in% ai_lexicon_bigrams) |>
  rename(term = bigram) |>
  mutate(ngram_level = "bigram")

# Trigram matches: tokenize at trigram level and match against trigram lexicon
ai_hits_trigrams <- abstracts_base |>
  unnest_tokens(trigram, abstract, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  filter(trigram %in% ai_lexicon_trigrams) |>
  rename(term = trigram) |>
  mutate(ngram_level = "trigram")

# Combine all matches into a single dataframe
ai_hits <- bind_rows(ai_hits_unigrams, ai_hits_bigrams, ai_hits_trigrams)
ai_hits
```

##### AI terms frequency over time

```{r}
# Count unique articles containing at least one AI related term per year
ai_articles_by_year <- ai_hits |>
  distinct(id, publication_year) |>
  count(publication_year, name = "articles_with_ai")

# Total number of abstracts per year
abstracts_by_year <- articles |>
  filter(!is.na(abstract), publication_year >= 2016, publication_year <= 2025) |>
  count(publication_year, name = "total_abstracts")

# Compute proportion
ai_proportion <- ai_articles_by_year |>
  left_join(abstracts_by_year, by = "publication_year") |>
  mutate(proportion_ai_articles = articles_with_ai / total_abstracts)

# Find the maximum proportion to place the annotation dynamically
max_prop <- max(ai_proportion$proportion_ai_articles, na.rm = TRUE)

ggplot(ai_proportion, aes(x = publication_year, y = proportion_ai_articles)) +
  # Use geom_col for a bar plot instead of geom_line
  geom_col(fill = "#2C3E6B", width = 0.7) +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  labs(
    title = "AI-related terminology in social science abstracts",
    subtitle = "Proportion of articles containing at least one AI term, by year",
    x = "Publication year",
    y = "Proportion of articles (%)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title         = element_text(size = 14, face = "bold", color = "#1a1a2e"),
    plot.subtitle      = element_text(size = 10, color = "#4a4a6a"),
    axis.text          = element_text(color = "#4a4a6a"),
    panel.grid.major.x = element_blank(), # Removes vertical grid lines for cleaner bars
    panel.grid.minor.x = element_blank()
  )
```

The presence of AI-related terms was insignificant before 2019, but it quickly grew, reaching 4% in 2022 ---the year of ChatGPT's public release---, and surpassing 10% in 2025. Since our lexicon prioritizes precision over recall, these are probably underestimates. But the trend is clear: attention to AI has greatly increased following the rise of generative AI tools.

Let's explore which terms are driving the trend.

```{r}
#| fig-width: 6
#| fig-height: 8
# Top AI terms per year by proportion of articles
ai_hits |>
  # Keep only unique combinations of article, year, and term
  distinct(id, publication_year, term) |>
  # Count the number of articles per term per year
  count(publication_year, term, name = "article_count") |>
  # Join total abstracts per year to compute proportion
  left_join(abstracts_by_year, by = "publication_year") |>
  mutate(proportion = article_count / total_abstracts) |>
  # Group by publication year and get most frequent terms
  group_by(publication_year) |>
  slice_max(proportion, n = 5, with_ties = FALSE) |>
  # Assign a rank (1 to 5) to the top terms
  mutate(rank = row_number()) |>
  ungroup() |>
  # Add 0 value to missing ranks so every year has exactly ranks 1 through 5
  # and all bars have the same width
  complete(publication_year = 2016:max(publication_year), rank = 1:5, fill = list(proportion = 0)) |>
  # Create invisible unique names for the blank bars using spaces
  mutate(
    term = ifelse(is.na(term), strrep(" ", rank), term),
    term = reorder_within(term, -rank, publication_year)
  ) |>
  
  # Plot it
  ggplot(aes(x = proportion, y = term, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE, fill = "#2C3E6B") +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_x_continuous(
    labels = percent_format(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.1)),
    n.breaks = 3
  ) +
  scale_y_reordered() +
  labs(
    title = "Top AI-related terms per year",
    subtitle = "Proportion of articles containing each term",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text         = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.y        = element_text(size = 9, color = "#4a4a6a"),
    axis.text.x        = element_text(color = "#4a4a6a"),
    panel.grid.major.y = element_blank()
  )
```

Before 2020 fewer than five abstracts match any lexicon term; from 2020 onwards, *artificial intelligence* and its abbreviation are the main drivers; from 2023, *chatgpt* and *generative ai* appear at the same time as mentions to AI greatly increase, suggesting the rise of generative AI influenced academic interests in the social sciences. Since the term *"ai"* is the main driver, we explore it by itself:

```{r}
# Filter and count for just the term "ai"
ai_hits |>
  filter(term == "ai") |>
  distinct(id, publication_year) |>
  count(publication_year, name = "article_count") |>
  complete(publication_year = 2016:max(publication_year), fill = list(article_count = 0)) |>
  # Join total abstracts per year to compute proportion
  left_join(abstracts_by_year, by = "publication_year") |>
  mutate(proportion = article_count / total_abstracts) |>
  
  # Plot it
  ggplot(aes(x = publication_year, y = proportion)) +
  geom_line(color = "#2C3E6B", linewidth = 1) +
  geom_point(color = "#2C3E6B", size = 2) +
  scale_x_continuous(breaks = function(x) seq(ceiling(min(x)), floor(max(x)), by = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Trend of the term 'ai' over time",
    subtitle = "Proportion of articles containing the term",
    x = "Publication Year",
    y = "Proportion of articles (%)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.y        = element_text(size = 9, color = "#4a4a6a"),
    axis.text.x        = element_text(color = "#4a4a6a"),
    panel.grid.minor.x = element_blank()
  )
```

When examining the term ai in isolation, the pattern becomes even clearer: its usage explodes from 2023 onwards, coinciding with the rise of generative AI.

#### 3.2.3 Topic modelling (LDA)

We now move from individual words to latent topics. We use LDA to discover the main topics in our corpus and then examine how the weight of these topics has shifted year by year across our full study period (2016–2025).

##### Preparing the Document-Term Matrix

```{r}
# We create the Document-Term Matrix (DTM) required by LDA.
# We cast our tidy data (already filtered by standard and domain stopwords)
# to DTM format using the article id as the document identifier.
abstracts_dtm <- tidy_abstracts |>
  count(id, word) |>
  cast_dtm(id, word, n)

abstracts_dtm

# Manually calculate sparsity to see the exact value:
cat("Precise sparsity   :", 1 - (464617/358239443), "%")
```

We get a DTM with 9482 documents and 38566 terms, sparsity close to 100% (99.87%), indicating very varied vocabulary in our abstracts.

##### LDA model with k=5

```{r}
#| fig-width: 9
#| fig-height: 8

# Fit the LDA model. We set a seed for reproducibility.
lda_model5 <- LDA(abstracts_dtm, k = 5, control = list(seed = 123))

# Extract per-topic-per-word probabilities (beta)
topics_beta5 <- tidy(lda_model5, matrix = "beta")

# Top 10 highest-probability words per topic
top_terms5 <- topics_beta5 |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, desc(beta))

# Visualize top terms per topic
top_terms5 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Top 10 terms per topic (LDA, k = 5)",
    x = expression(beta),
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

We see some patterns, but topics are broad: topic 1 groups economic and sustainability terms, topic 2 politics and elections, topic 3 gender and family, and topic 4 is about COVID-19 and health. Topic 5 mixes education, AI, media, and language. In all topics, but especially 5, it seems like distinct topics are being mixed, we thus tried a higher k.

##### LDA model with k=6

```{r}
#| fig-width: 9
#| fig-height: 8

# Fit the LDA model. We set a seed for reproducibility.
lda_model6 <- LDA(abstracts_dtm, k = 6, control = list(seed = 123))

# Extract per-topic-per-word probabilities (beta)
topics_beta6 <- tidy(lda_model6, matrix = "beta")

# Top 10 highest-probability words per topic
top_terms6 <- topics_beta6 |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, desc(beta))

# Visualize top terms per topic
top_terms6 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Top 10 terms per topic (LDA, k = 6)",
    x = expression(beta),
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```
A new media and mental health topic appears, improving interpretability. However, the remaining topics still show significant overlaps, particularly around education and AI, so we try a higher k.

##### LDA model with k=7

```{r}
#| fig-width: 9
#| fig-height: 8

# Fit the LDA model. We set a seed for reproducibility.
lda_model7 <- LDA(abstracts_dtm, k = 7, control = list(seed = 123))

# Extract per-topic-per-word probabilities (beta)
topics_beta7 <- tidy(lda_model7, matrix = "beta")

# Top 10 highest-probability words per topic
top_terms7 <- topics_beta7 |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, desc(beta))

# Visualize top terms per topic
top_terms7 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Top 10 terms per topic (LDA, k = 7)",
    x = expression(beta),
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

At k = 7 the topics reach a pretty clear configuration: economic sustainability (topic 1), politics and elections (2), cultural discourse and identity (3), health and pandemic (4), media and mental health (5), AI and education (6), and gender inequality (7). AI is grouped with education and learning, suggesting that AI is primarly discussed in this context.

##### LDA model with k=8

```{r}
#| fig-width: 9
#| fig-height: 8

# Fit the LDA model. We set a seed for reproducibility.
lda_model8 <- LDA(abstracts_dtm, k = 8, control = list(seed = 123))

# Extract per-topic-per-word probabilities (beta)
topics_beta8 <- tidy(lda_model8, matrix = "beta")

# Top 10 highest-probability words per topic
top_terms8 <- topics_beta8 |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, desc(beta))

# Visualize top terms per topic
top_terms8 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Top 10 terms per topic (LDA, k = 8)",
    x = expression(beta),
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 9)
  )
```

Interpretability decreases, with a new topic combining climate, policy, risk, and legal terms. We thus choose k = 7 to analyze the evolution of topics over time.

##### How do topics evolve year by year? 

```{r}
#| fig-width: 10
#| fig-height: 5

# Extract per-document-per-topic probabilities (gamma)
topics_gamma7 <- tidy(lda_model7, matrix = "gamma")

# Join gamma with article metadata and average by year
topics_by_year7 <- topics_gamma7 |>
  left_join(
    articles |>
      select(id, publication_year) |>
      mutate(id = as.character(id)),
    by = c("document" = "id")
  ) |>
  group_by(topic, publication_year) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop")

# First point of each topic for the label layer (all except topic 1)
first_points <- topics_by_year7 |>
  group_by(topic) |>
  slice_min(publication_year, n = 1) |>
  filter(topic != 1)

# Last point of topic 1 for the label layer
last_point_t1 <- topics_by_year7 |>
  filter(topic == 1) |>
  slice_max(publication_year, n = 1)

# Plot: topic prevalence by year (line plot)
ggplot(topics_by_year7,
       aes(x = publication_year, y = mean_gamma, color = factor(topic))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  # Larger point at first year for topics 2-7
  geom_point(data = first_points, size = 7) +
  # Topic number inside the first point (topics 2-7)
  geom_text(data = first_points, aes(label = topic),
            color = "white", size = 3, fontface = "bold", family = "serif") +
  # Larger point at last year for topic 1
  geom_point(data = last_point_t1, size = 7) +
  # Topic number inside the last point (topic 1)
  geom_text(data = last_point_t1, aes(label = topic),
            color = "white", size = 3, fontface = "bold", family = "serif") +
  # Reference lines at key events
  geom_vline(xintercept = 2019.5, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 2022.5, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  annotate("text", x = 2017.5, y = Inf, label = "Pre-pandemic",
           vjust = 1.5, size = 3, color = "grey40", family = "serif") +
  annotate("text", x = 2021,   y = Inf, label = "Pandemic",
           vjust = 1.5, size = 3, color = "grey40", family = "serif") +
  annotate("text", x = 2024,   y = Inf, label = "Rise of generative AI",
           vjust = 1.5, size = 3, color = "grey40", family = "serif") +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Topic prevalence over time (LDA, k = 7)",
    subtitle = "Average γ per topic per year — dashed lines mark key inflection points",
    x = "Publication year",
    y = "Average topic proportion (γ)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title      = element_text(size = 14, face = "bold"),
    plot.subtitle   = element_text(size = 10, color = "#4a4a6a"),
    legend.position = "none"
  )
```


We observe a clear pattern: most topics follow a relatively stable increasing trend throughout the decade, while others gradually decline. However, inflection points appear in relation to the main events identified in the previous subsections (the pandemic in 2020 and the rise of generative AI). Notably, while the AI-related topic (topic 6) does not seem to spike precisely at 2022, a general pattern shift is observable across most topics after that year, suggesting the rise of generative AI had an effect, at least as a catalyzer for existing trends. To check the robustness of the pattern, we visualize the same graph on all the models we have performed.


```{r}
# Named list to identify models in the plot
model_list <- list(
  "k = 5" = lda_model5,
  "k = 6" = lda_model6,
  "k = 7" = lda_model7,
  "k = 8" = lda_model8
)

# Extract gamma for all models and bind them into one dataframe
# .id = "k_val" creates a column with the names from the list above
all_gamma <- map_dfr(model_list, ~tidy(.x, matrix = "gamma"), .id = "k_val")

# Join with metadata and average by year and k_val
all_topics_by_year <- all_gamma |>
  left_join(
    articles |>
      select(id, publication_year) |>
      mutate(id = as.character(id)),
    by = c("document" = "id")
  ) |>
  group_by(k_val, topic, publication_year) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop")

# Create the faceted plot
ggplot(all_topics_by_year, 
       aes(x = publication_year, y = mean_gamma, color = factor(topic))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 1.2) +
  # Add reference lines (Annotations removed for clarity in facets)
  geom_vline(xintercept = c(2019.5, 2022.5), linetype = "dashed", 
             color = "grey70", linewidth = 0.3) +
  # Faceting: 2 columns, 2 rows. free_y is critical for comparing trends.
  facet_wrap(~ k_val, ncol = 2, scales = "free_y") + 
  scale_x_continuous(breaks = seq(2016, 2024, by = 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Topic Prevalence Stability across different values of k",
    x = "Publication Year",
    y = "Mean Topic Proportion (γ)",
    color = "Topic #"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1.5, "lines")
  )
```

The shift coinciding with the rise of generative AI is consistent across all values of k.

### 3.3 Summary of results (Igual en unas conclusiones junto todo lo demás al final del trabajo)

Regarding RQ1, topics addressed by social scientists do not suddenly change from one year to the next, but do show evolution over time. This relatively slow change is to be expected given the ... los científicos sociales necesitan tiempo para producir... especialización que no se consigue d eun día para otro... en fin los tiempos de la ciencia. Having said that, find that some events greatly influence what social scientist pay attention to. In particular, our analyses highlight the influence of two events:

First, the COVID-19 pandemic reclaimed the attention of social scientists. It dominates TF-IDF rankings at every n-gram level from 2020 onwards, and defines a topic by itself when topics stabilise at k=7 in the LDA.

Secondly ---and connecting with RQ2---, there is no doubt that AI has received greatly increased attention in recent year, particularly after the rise of generative AI. We observe this pattern in the TF-IDF anlysis, with AI and related terms appearing from 2023 onwards, and in the AI lexicon frequency analysis ---a conservative estimation showed that the proportion of abstracts containing at least one AI-related term went from almost 0% before 2019 to over 10% in 2025. When performing LDA we did not find a topic exclusively for AI; at k = 7 AI appears as an important term within a a topic focused in education. However, when analyzing the evolution of topics over time, we observe big shifts coinciding with the rise of generative AI; this suggests that, even if AI has not become a subject of research by itself ---which makes sense, given that we are talking about leading social science journals, AI is more likely being studied in relation with other topics---, the rise of generative AI seems to have influenced ... other topics.

Limitations:
- Unbalanced corpus across time (although this reflects a genuine phenomena in top social science journals).
- An abstract dataset sometimes proved too small (for example, in tf-idf section a single abstact could make some terms break into the top 8). Further research: using full articles (or adding more journals).
- Cannot establish causality.
- QUÉ más?


## Section 4. WRITING STYLE AND THE RISE OF AI \[RQ3\]

As we have seen until now, papers on AI have risen exponentially sincet t

```{r number of articles by year}
articles |>
  filter(!is.na(abstract)) |>
  count(publication_year) |>
  ggplot(aes(x = publication_year, y = n)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    title = "Number of Abstracts by Year",
    x = "Publication Year",
    y = "Number of Abstracts"
  )
```

```{r}
tidy_abstracts |>
  count(word, sort = TRUE) |>
  slice_max(n, n = 20) |>
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Words in the Corpus",
    x = "Word",
    y = "Count"
  )
```

```{r}

#Based in Kousha & Thelwall, 2025
llm_words <- tibble(word = c(
  "underscore", "underscores", "underscored", "underscoring",
  "delve", "delves", "delved", "delving",
  "showcase", "showcases", "showcased", "showcasing",
  "unveil", "unveils", "unveiled", "unveiling",
  "intricate", "intricates", "intricated", "intricating",
  "meticulous", "meticulously",
  "pivotal",
  "heighten", "heightens", "heightened", "heightening",
  "nuance", "nuances", "nuanced",
  "bolster", "bolsters", "bolstered", "bolstering",
  "foster", "fosters", "fostered", "fostering",
  "interplay", "interplays", "interplayed", "interplaying",
# Added to account for 2025 findings
  "core"
))

control_words <- tibble(word = c("research", "findings", "policy", "support"))
```

```{r}
pattern_1 <- paste(llm_words$word, collapse = "|")

llm_words_freq <- tidy_abstracts_raw |>
  filter(str_detect(word, regex(pattern_1, ignore_case = TRUE))) |> 
  count(word, sort = TRUE)

ggplot(llm_words_freq |> 
         filter(!word %in% c("analysis", "findings", "policy", "support", "policymakers")) |>
         head(20), aes(x = reorder(word, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Frequency of AI signature Words in Abstracts",
    x = "Word",
    y = "Count"
  )
```

```{r}
library(patchwork)

top10_words_clean <- llm_words_freq |>
  filter(!word %in% control_words$word) |>
  slice_max(n, n = 10) |>
  pull(word)

papers_per_year <- tidy_abstracts_raw |>
  distinct(id, publication_year) |>
  count(publication_year, name = "total_papers")

plot_data <- tidy_abstracts_raw |>
  filter(word %in% c(top10_words_clean, control_words$word)) |>
  count(word, publication_year) |>
  left_join(papers_per_year, by = "publication_year") |>
  mutate(pct = n / total_papers * 100)

p_llm <- plot_data |>
  filter(word %in% top10_words_clean) |>
  ggplot(aes(x = publication_year, y = pct)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ word, scales = "free_y", ncol = 2) +
  labs(title = "Top 10 LLM Words", x = "Year", y = "% of Papers")

p_control <- plot_data |>
  filter(word %in% control_words$word) |>
  ggplot(aes(x = publication_year, y = pct)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ word, scales = "free_y", ncol = 1) +
  labs(title = "Control Words", x = "Year", y = "% of Papers")

p_control | p_llm
```

Add CONTROL TERMS

What about delve?

```{r}
papers_per_year <- tidy_abstracts_raw |>
  distinct(id, publication_year) |>
  count(publication_year, name = "total_papers")

tidy_abstracts_raw |>
  filter(word == "delve") |>
  count(publication_year) |>
  left_join(papers_per_year, by = "publication_year") |>
  mutate(pct = n / total_papers * 100) |>
  ggplot(aes(x = publication_year, y = pct)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  labs(
    title = 'Distribution of "delve" by Year',
    x = "Year",
    y = "% of Papers"
  )
```

How many papers have these words after 2022?

```{r}
tidy_abstracts_raw |>
  filter(word %in% llm_words$word, publication_year > 2022) |>
  distinct(id) |>
  left_join(
    articles |> select(id, title, publication_year, source_display_name, doi),
    by = "id"
  ) |>
  arrange(publication_year, title)
```

```{r}
library(httr2)

# Store your key: Sys.setenv(SAPLING_KEY = "your_key_here")
api_key <- "HAPVZA5R2TKPBA5IX02QKCH2P799UI9Q"

# Sample 10 abstracts from papers after 2022 containing top 10 words
set.seed(4823)
sample_ids <- tidy_abstracts_raw |>
  filter(word %in% llm_words$word, publication_year > 2022) |>
  distinct(id) |>
  slice_sample(n = 10) |>
  pull(id)

sample_abstracts <- articles |>
  filter(id %in% sample_ids) |>
  select(id, title, abstract)

# Send each abstract to the Sapling AI detection API
sapling_results <- sample_abstracts |>
  rowwise() |>
  mutate(
    response = list(
      request("https://api.sapling.ai/api/v1/aidetect") |>
        req_body_json(list(key = api_key, text = abstract)) |>
        req_perform() |>
        resp_body_json()
    ),
    ai_score = response$score
  ) |>
  ungroup() |>
  select(id, title, ai_score)

sapling_results
```


## Section 5. SENTIMENT TOWARDS AI IN SOCIAL SCIENCE ABSTRACTS \[RQ4\]

Our fourth research question —How has sentiment towards AI in social science literature changed over the last decade?— looks at how attitudes towards AI have changed during the period studied.

The **limitations** in answering this research question are significant: by selecting the subcorpus of analysis as the abstracts that contain terms related to artificial intelligence, we cannot guarantee that the sentiment is generated by evaluations directed specifically at AI itself, but rather that it may be influenced by words that frame the research context. Therefore, this first analysis is far from answering the research question in depth as we have formulated it, and would rather capture the general evaluative tone of abstracts that contain mentions of AI-related terms.

For this reason, the **analysis has been designed sequentially to approach the object of study**. First, a sentiment analysis is conducted on the full abstracts that mention terms related to AI. In the second step, only sentences containing explicit AI terms are considered to calculate the general sentiment of each abstract. In the third analysis, we select only those sentences containing bigrams/trigrams/fourgrams that address AI and its effects or consequences, thus making AI the focus of interest in that sentence and not just a simple mention. This is how we attempt to capture the sentiment these abstracts express regarding artificial intelligence.

In addition, we acknowledge the limitation of analyzing sentiment in abstracts of scientific texts, which are characterized by the neutrality of the vocabulary used. This is reflected in the exclusion of some abstracts and sentences (across all three levels of analysis) because none of their words appear in the sentiment lexicons used, Bing and AFFIN.

### 5.0 Set-up for sentiment analysis

Before analyzing the sentiment, we need to define de abstracts that would be part of the subcorpus of analysis. In this case the object of the sentiment is the "artificial intelligence", therefore, we just select the abstracts that have explicit mentions to IA and related terms and n-grams, using the ai_hits created above.

```{r}
# Creating the sentiment analysis subcorpus, selecting only the abstracts that contain the n-grams in ai_hits 
ai_abstracts <- ai_hits |>
  distinct(id) |>
  inner_join(abstracts_base, by = "id")
ai_abstracts
```

As mentioned, abstracts with no matches in the sentiment lexicons are excluded from the sentiment computing step. These texts probably have a more neutral and technical language, although their omission may capture limitations of lexicon-based sentiment analysis rather than the real absence of evaluative words about IA. In both cases of Bing and AFINN the number of abstracts excluded are 2 (529 total AI-related abstracts and 527 abstracts that contain at least one word in both lexicons).

```{r, warning = FALSE}
# Total of AI-related abstracts
ai_abstracts |> 
  summarise(n_ai_abstracts = n_distinct(id))

# Abstracts that have at least one word in the lexicon Bing 
ai_abstracts |>
  unnest_tokens(word, abstract) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  summarise(n_scored_abstracts = n_distinct(id))

# Abstracts that have at least one word in the lexicon Bing AFINN
ai_abstracts |>
  unnest_tokens(word, abstract) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  summarise(n_scored_abstracts = n_distinct(id))
```

Here, we can observe that both lexicons are negatively biased, this is, that they contain more terms that are categorized as negative (or negative values). This could influence sentiment results, making them more likely to lean to a negative direction; therefore, the results are influenced by the choice of lexicon. For this reason, all of the analysis are performed using both lexicons, which serves as a robustness check.

```{r}
# Calculating the number of times that a sentiment or a value appears in each lexicon > negatively biased
get_sentiments("bing") %>% 
  count(sentiment)

get_sentiments("afinn") %>% 
  count(value)
```

### 5.1. General sentiment in AI-related abstracts over time

This is the first layer of approximation to the analysis, that aims to estimate the general evaluative tone or sentiment in abstracts that mention AI and related terms as well as its evolution over time.

The methodological choice was to calculate a sentiment score per abstract and, after that, obtain the annual value as the average sentiment score of the articles published in a specific year. This ensures that abstracts remain the unit of analysis counting only once. In this connection, we decided to use net_sentiment_norm for Bing and afinn_mean for AFINN because in both cases the aim is to get a sentiment measure that can be compared across abstracts and does not depend on how many words/sentiment words each abstract contains.

```{r, warning=FALSE}
# In this chunk we create two tibbles, one per lexicon, that contain the id, the publication year and the sentiment variables related to each lexicon.

# In the Bing one, we compute the number of positive and negative words, the net sentiment (positive - negative) and the total sentiment words (positive + negative). With this last two values we obtain the variable of interest net_sentiment_norm (net sentiment / total sentiment words). In the AFINN tibble we computed the afinn mean, afinn sum and the number of sentiment words. 

# Bing
bing_abstract_sentiment <- ai_abstracts |>
  unnest_tokens(word, abstract) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  count(id, publication_year, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(
    net_sentiment = positive - negative,
    total_sentiment_words = positive + negative,
    net_sentiment_norm = if_else(
      total_sentiment_words > 0,
      (positive - negative) / total_sentiment_words,
      NA_real_))

bing_abstract_sentiment

# AFINN
afinn_abstract_sentiment <- ai_abstracts |>
  unnest_tokens(word, abstract) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  group_by(id, publication_year) |>
  summarise(
    afinn_sum = sum(value, na.rm = TRUE),
    afinn_mean = mean(value, na.rm = TRUE),
    n_sentiment_words = n(),
    .groups = "drop")

afinn_abstract_sentiment
```

```{r}
# Combining both lexicons into one single tibble 
bing_scores <- bing_abstract_sentiment |>
  transmute(
    id,
    publication_year,
    lexicon = "Bing",
    sentiment_score = net_sentiment_norm)

afinn_scores <- afinn_abstract_sentiment |>
  transmute(
    id,
    publication_year,
    lexicon = "AFINN",
    sentiment_score = afinn_mean)

sentiment_scores <- bind_rows(bing_scores, afinn_scores)

# Calculating the annual average sentiment by lexicon 
sentiment_yearly <- sentiment_scores |>
  group_by(lexicon, publication_year) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop")
sentiment_yearly
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the evolution of the annual average sentiment score: line chart 
ggplot(sentiment_yearly, aes(x = publication_year, y = mean_sentiment, color = lexicon)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 2016:2025,
    labels = 2016:2025
  ) +
  scale_color_manual(values = c("Bing" = "#2C3E6B", "AFINN" = "#8E44AD")) +
  labs(
    title = "Sentiment in AI-related abstracts over time",
    subtitle = "Yearly mean sentiment per abstract using Bing and AFINN lexicons",
    x = NULL,
    y = "Mean sentiment score",
    color = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "top",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank())

# Plotting the comparison before-after the rise of AI of the annual avergae sentiment score: boxplot 

period_scores <- sentiment_scores |>
  mutate(
    period = if_else(
      publication_year < 2023,
      "Before rise of AI",
      "After rise of AI"))

period_summary <- period_scores |>
  group_by(lexicon, period) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    median_sentiment = median(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop")

ggplot(
  period_scores,
  aes(x = period, y = sentiment_score, fill = period)) +
  geom_boxplot(alpha = 0.85, width = 0.6, outlier.alpha = 0.2) +
  scale_fill_manual(values = c(
    "Before rise of AI" = "steelblue",
    "After rise of AI"  = "#2C3E6B"
  )) +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Sentiment in AI-related abstracts before and after generative AI",
    x = NULL,
    y = "Sentiment score") +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "none",
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank()
  )
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the number of positive, negative, and neutral AI-related abstracts by year 

# Bing
bing_year_polarity <- bing_abstract_sentiment |>
  mutate(
    polarity = case_when(
      net_sentiment_norm > 0 ~ "Positive",
      net_sentiment_norm < 0 ~ "Negative",
      TRUE ~ "Neutral")) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "Bing")

# AFINN
afinn_year_polarity <- afinn_abstract_sentiment |>
  mutate(
    polarity = case_when(
      afinn_mean > 0 ~ "Positive",
      afinn_mean < 0 ~ "Negative",
      TRUE ~ "Neutral")) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "AFINN")

# Combine both lexicons
year_polarity <- bind_rows(
  bing_year_polarity,
  afinn_year_polarity)

ggplot(
  year_polarity,
  aes(x = factor(publication_year), y = n, fill = polarity)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "Positive" = "#27AE60",
    "Negative" = "#C0392B",
    "Neutral"  = "#2C3E6B"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Number of positive, negative, and neutral AI-related abstracts by year",
    x = NULL,
    y = "Number of abstracts",
    fill = "Overall sentiment"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    legend.position      = "bottom",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor     = element_blank(),
    panel.grid.major.y   = element_blank())
```

Firstly, the annual average sentiment scores show that the abstracts containing AI-related terms tend to have a general moderate positive tone across years. This pattern is observed in both lexicons, especially in AFINN that estimates higher positive values than Bing. In both lexicons the sentiment is neutral in 2017, rises in 2018 and declines in 2019. After 2020 Bing follows a softer pattern, declining until 2022 and then recovering from 2023 and AFINN stays clearly more positive and stable across years, with small fluctuations. Comparing the tendencies before and after the rise of AI in 2023 we can observe that the central tendency becomes somewhat more positive after 2023. The last graph shows the absolute number of positive and negative abstracts over time, and tells us that this difference becomes even more visible in the recent years, when the number of AI-related abstracts increases.

### 5.2 General sentiment in AI-related abstracts over time: using sentences mentioning AI

With full abstracts the sentiment score may be driven by any part of the text, even when AI is not the topic being evaluated but mentioned briefly. In order to reduce this noise, we also analysed the sentiment at the sentence level, selecting only the sentences that explicitly mention at least one AI-related term from the unigram, bigram, or trigram lexicons. To ensure comparability with the previous abstract-level results, we computed the sentence-level sentiment (with the net_sentiment_norm and the afinn_mean) and then average it back to the abstract level. As before, each abstract contributes a single observation to the annual aggregate

```{r}
# Creating the subcorpus of sentences that explicitly mention AI-related terms 
# Base sentences from AI-related abstracts
sentences_base <- ai_abstracts |>
  unnest_tokens(sentence, abstract, token = "sentences", drop = FALSE) |>
  group_by(id) |>
  mutate(sentence_id = row_number()) |>
  ungroup() |>
  select(id, publication_year, sentence_id, sentence, abstract)

# Unigram in sentences
sentence_hits_unigrams <- sentences_base |>
  unnest_tokens(word, sentence, token = "words") |>
  filter(word %in% ai_lexicon_unigrams) |>
  transmute(id, publication_year, sentence_id)

# Bigram in sentences
sentence_hits_bigrams <- sentences_base |>
  unnest_tokens(bigram, sentence, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  filter(bigram %in% ai_lexicon_bigrams) |>
  transmute(id, publication_year, sentence_id)

# Trigram in sentences
sentence_hits_trigrams <- sentences_base |>
  unnest_tokens(trigram, sentence, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  filter(trigram %in% ai_lexicon_trigrams) |>
  transmute(id, publication_year, sentence_id)

# Combining all sentence-level AI n-grams 
sentence_ai_hits <- bind_rows(
  sentence_hits_unigrams,
  sentence_hits_bigrams,
  sentence_hits_trigrams) |>
  distinct(id, publication_year, sentence_id)

# Keeping the sentences that explicitly mention AI-related terms
ai_sentences <- sentences_base |>
  semi_join(sentence_ai_hits, by = c("id", "publication_year", "sentence_id"))

ai_sentences
```

Following the previous logic, sentences with no matches in the sentiment lexicons are excluded from the sentiment computing step. The total number of sentences with AI terms are 1544, while in Bing the analysis is carried out using 1210 sentences and in AFINN using 1023.

```{r, warning=FALSE}
# We create a sentence-level id 
ai_sentences_ids <- ai_sentences |>
  mutate(sentence_uid = paste(id, sentence_id, sep = "_"))

# Total sentences that mention AI-related terms 
ai_sentences_ids |>
  summarise(n_ai_sentences = n_distinct(sentence_uid))

# Sentences that have at least one word in the lexicon Bing 
ai_sentences_ids |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  summarise(n_scored_sentences = n_distinct(sentence_uid))

# Sentences that have at least one word in the lexicon AFINN  
ai_sentences_ids |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  summarise(n_scored_sentences = n_distinct(sentence_uid))
```

```{r}
# In this chunk we create two tibbles, one per lexicon, that contain the id of the abstract and sentence, the publication year and the sentiment variables related to each lexicon.

# Bing 
bing_sentence_sentiment <- ai_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  count(id, publication_year, sentence_id, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(
    net_sentiment = positive - negative,
    total_sentiment_words = positive + negative,
    net_sentiment_norm = if_else(
      total_sentiment_words > 0,
      (positive - negative) / total_sentiment_words,
      NA_real_))
bing_sentence_sentiment

# AFINN
afinn_sentence_sentiment <- ai_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  group_by(id, publication_year, sentence_id) |>
  summarise(
    afinn_sum = sum(value, na.rm = TRUE),
    afinn_mean = mean(value, na.rm = TRUE),
    n_sentiment_words = n(),
    .groups = "drop")
afinn_sentence_sentiment
```

```{r}
# Aggregating sentence-level sentiment to abstract level

# BING: average sentence sentiment within each abstract
bing_sentence_to_abstract <- bing_sentence_sentiment |>
  group_by(id, publication_year) |>
  summarise(
    sentiment_score = mean(net_sentiment_norm, na.rm = TRUE),
    n_ai_sentences = n(),
    .groups = "drop") |>
  mutate(lexicon = "Bing")

# AFINN: average sentence sentiment within each abstract
afinn_sentence_to_abstract <- afinn_sentence_sentiment |>
  group_by(id, publication_year) |>
  summarise(
    sentiment_score = mean(afinn_mean, na.rm = TRUE),
    n_ai_sentences = n(),
    .groups = "drop") |>
  mutate(lexicon = "AFINN")

# Combine both lexicons into one single tibble 
sentence_abstract_scores <- bind_rows(
  bing_sentence_to_abstract,
  afinn_sentence_to_abstract)

# Calculating the annual average sentiment by lexicon 
sentence_abstract_yearly <- sentence_abstract_scores |>
  group_by(lexicon, publication_year) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop")
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the evolution of the annual average sentiment score (using sentences): line chart 
ggplot(sentence_abstract_yearly, aes(x = publication_year, y = mean_sentiment, color = lexicon)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 2016:2025,
    labels = 2016:2025
  ) +
  scale_color_manual(values = c("Bing" = "#2C3E6B", "AFINN" = "#8E44AD")) +
  labs(
    title = "Sentiment in AI-related abstracts over time",
    subtitle = "Yearly mean sentiment per abstract based on AI-mentioning sentences only",
    x = NULL,
    y = "Mean sentiment score",
    color = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "top",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank())

# Plotting the comparison before-after the rise of AI of the annual avergae sentiment score (using sentences): boxplot 

sentence_period_scores <- sentence_abstract_scores |>
  mutate(
    period = if_else(
      publication_year < 2023,
      "Before rise of AI",
      "After rise of AI"))

sentence_period <- sentence_period_scores |>
  group_by(lexicon, period) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    median_sentiment = median(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop")

ggplot(
  sentence_period_scores,
  aes(x = period, y = sentiment_score, fill = period)
) +
  geom_boxplot(alpha = 0.85, width = 0.6, outlier.alpha = 0.2) +
  scale_fill_manual(values = c(
    "Before rise of AI" = "steelblue",
    "After rise of AI"  = "#2C3E6B"
  )) +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Sentiment in AI-related abstracts before and after generative AI",
    subtitle = "Based on AI-mentioning sentences only",
    x = NULL,
    y = "Sentiment score"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "none",
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank()
  )
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the number of positive, negative, and neutral AI-related abstracts by year (using sentences)

# Bing
bing_sentence_year_polarity <- bing_sentence_to_abstract |>
  mutate(
    polarity = case_when(
      sentiment_score > 0 ~ "Positive",
      sentiment_score < 0 ~ "Negative",
      TRUE ~ "Neutral")) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "Bing")

# AFINN
afinn_sentence_year_polarity <- afinn_sentence_to_abstract |>
  mutate(
    polarity = case_when(
      sentiment_score > 0 ~ "Positive",
      sentiment_score < 0 ~ "Negative",
      TRUE ~ "Neutral")) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "AFINN")

# Combine both lexicons
sentence_year_polarity <- bind_rows(
  bing_sentence_year_polarity,
  afinn_sentence_year_polarity)

ggplot(
  sentence_year_polarity,
  aes(x = factor(publication_year), y = n, fill = polarity)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "Positive" = "#27AE60",
    "Negative" = "#C0392B",
    "Neutral"  = "#2C3E6B"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Number of positive, negative, and neutral AI-related abstracts by year",
    subtitle = "Based on AI-mentioning sentences only",
    x = NULL,
    y = "Number of abstracts",
    fill = "Overall sentiment"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    legend.position      = "bottom",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor     = element_blank(),
    panel.grid.major.y   = element_blank()
  )
```

### 5.3. Towards a more target-specific measure of sentiment around AI

The final stage of the analysis moves towards a more target-specific measure of sentiment around AI, trying to identify sentences in which AI is explicitly linked to consequences, risks, benefits, implications, or evaluations. To do this, the code builds a new lexicon by combining AI N-grams with effect and outcome terms: we defined separate vectors for AI-side components and effect-side components, and these were combined into bigrams, trigrams, and fourgrams. After that, the same sentiment analysis procedure is applied: scoring target-specific sentences with Bing and AFINN, aggregating them back to the abstract level, calculating yearly means, comparing positive and negative counts, and examining before/after differences around 2023.

```{r}
# Building the AI-related N-grams lexicon 
# AI SIDE COMPONENT VECTORS

# 1-word AI terms (drawn from ai_lexicon_unigrams)
ai_side_1w <- c(
  # neural architecture terms
  "autoencoder", "perceptron", "neurocomputing",
  # specific AI systems and acronyms
  "chatgpt", "llm",
  # vendors and umbrella terms
  "deepseek", "openai", "ai"
)

# 2-word AI terms (drawn from ai_lexicon_bigrams)
ai_side_2w <- c(
  # core ML / DL paradigms
  "artificial intelligence", "machine learning", "deep learning",
  "reinforcement learning", "machine intelligence", "computational intelligence",
  # network and model architectures
  "neural network", "neural model", "deep network",
  "learning algorithm", "backpropagation learning", "backpropagation algorithm",
  "pretrained transformer",
  # generative and language model terms
  "generative ai", "language model", "agentic ai",
  # AI-generated content
  "ai-generated content",
  # specific systems and vendors
  "openai gpt",
  "claude sonnet", "claude opus", "claude haiku",
  "google gemini", "gemini pro", "gemini ultra",
  "microsoft copilot", "github copilot",
  "mistral ai", "meta llama", "llama 2", "llama 3",
  "stable diffusion",
  "gpt 1", "gpt 2", "gpt 3", "gpt 4", "gpt 5",
  "dall e"
)

# 3-word AI terms (drawn from ai_lexicon_trigrams)
ai_side_3w <- c(
  # large / generative language model formulations
  "large language model", "generative language model",
  "generative large language", "generative pretrained transformer",
  # generative AI formulations
  "generative artificial intelligence", "artificial intelligence generated",
  # OpenAI-specific phrases
  "openai large language", "openai language model",
  # AI + language model phrases
  "ai language model", "ai generated content"
)

# EFFECT / OUTCOME SIDE COMPONENT VECTORS

# 1-word effect and outcome terms
effect_side_1w <- c(
  # direct outcome terms (neutral)
  "effects", "impacts", "consequences", "implications", "outcomes",
  "influence", "affect", "repercussions", "ramifications", "externalities",
  "spillovers", "footprint",

  # valence terms — NEGATIVE (~37)
  "risks", "harms", "dangers", "threats", "hazards", "misuse", "abuse",
  "drawbacks", "disadvantages", "limitations", "failures", "errors",
  "vulnerabilities", "pitfalls", "toxicity", "hallucinations", "unreliability",
  "inaccuracy", "overreliance", "dependence", "dependency", "addiction",
  "misalignment", "deception", "manipulation", "exploitation", "disruption",
  "fragility", "brittleness", "opacity", "obfuscation", "misinformation",
  "disinformation", "propaganda", "plagiarism", "cheating", "displacement",

  # valence terms — POSITIVE (~37)
  "benefits", "opportunities", "advantages", "gains", "improvements",
  "efficiency", "productivity", "innovation", "augmentation", "empowerment",
  "democratization", "accessibility", "scalability", "sustainability",
  "breakthroughs", "capabilities", "accuracy", "reliability", "performance",
  "personalization", "optimization", "collaboration", "assistance",
  "enhancement", "acceleration", "advancement", "progress", "discovery",
  "creation", "support", "convenience", "precision", "explainability",
  "interpretability", "quality", "usefulness", "enablement",

  # neutral evaluative terms
  "challenges", "tradeoffs", "tensions", "paradoxes", "dilemmas",
  "uncertainties", "ambiguities",

  # societal and structural dimensions (neutral descriptors)
  "ethical", "social", "economic", "labor", "employment", "unemployment",
  "inequality", "bias", "fairness", "privacy", "security", "trust",
  "accountability", "transparency", "justice", "discrimination", "welfare",
  "autonomy", "dignity", "democracy", "sovereignty", "marginalization",
  "exclusion", "inclusion", "diversity", "racism", "sexism", "alienation",
  "dehumanization", "commodification", "precarity", "polarization",
  "radicalization", "deskilling", "upskilling", "reskilling", "cohesion",
  "fragmentation", "concentration", "monopolization",

  # environmental dimensions
  "emissions", "carbon", "consumption",

  # health dimensions
  "wellbeing", "health", "safety", "diagnosis", "treatment", "triage",

  # education dimensions
  "literacy", "education", "creativity", "learning",

  # governance terms
  "regulation", "policy", "governance", "legislation", "oversight",
  "compliance", "liability", "audit", "certification", "standards",
  "enforcement", "prohibition", "moratorium", "ban", "restriction",
  "responsibility", "stewardship",

  # human response terms
  "adoption", "acceptance", "resistance", "perception", "attitudes",
  "concerns", "fears", "anxiety", "skepticism", "awareness", "distrust",
  "mistrust", "optimism", "pessimism", "enthusiasm", "reliance",
  "reluctance", "readiness", "preparedness",

  # domain-specific effect nouns
  "automation", "surveillance", "deepfakes", "decision-making"
)

# 2-word effect phrases (for fourgram Pattern B: [AI 2w] + [effect 2w])
effect_side_2w <- c(
  # labour and economic effects — NEGATIVE
  "job displacement", "job loss", "labor disruption", "wage inequality",
  "income inequality", "wage stagnation", "skill displacement", "skill gap",
  "workforce disruption", "technological unemployment", "economic disruption",
  "market concentration", "power concentration",

  # labour and economic effects — POSITIVE
  "job creation", "economic growth", "productivity gains", "wage growth",
  "skill development", "workforce augmentation", "workforce transformation",
  "economic efficiency", "creative destruction", "market competition",
  "value creation",

  # societal and structural effects — NEGATIVE
  "social inequality", "social fragmentation", "digital divide",
  "political polarization", "information disorder", "democratic erosion",
  "civil rights violations",

  # societal and structural effects — POSITIVE
  "social cohesion", "social welfare", "human rights", "civil liberties",
  "democratic participation", "civic engagement", "social mobility",
  "community empowerment",

  # ethical and governance effects — NEGATIVE
  "ethical violations", "ethical concerns", "ethical challenges",
  "ethical dilemmas", "moral hazard", "regulatory failure", "governance gap",

  # ethical and governance effects — POSITIVE
  "ethical guidelines", "ethical frameworks", "policy innovation",
  "regulatory framework", "governance framework", "human oversight",
  "responsible innovation", "responsible development",

  # safety, trust, and reliability — NEGATIVE
  "privacy concerns", "security risks", "data breaches", "system failures",
  "algorithmic bias", "algorithmic discrimination", "automated discrimination",
  "predictive policing", "privacy erosion",

  # safety, trust, and reliability — POSITIVE
  "public trust", "data privacy", "data security", "information security",
  "system reliability", "algorithmic fairness", "algorithmic accountability",
  "algorithmic transparency",

  # health and wellbeing — NEGATIVE
  "mental health risks", "patient harm", "clinical errors",
  "psychological distress", "cognitive overload",

  # health and wellbeing — POSITIVE
  "mental health", "public health", "health outcomes", "patient safety",
  "clinical decision", "early detection", "disease prevention",
  "drug discovery", "personalized medicine",

  # information and media — NEGATIVE
  "fake news", "misinformation spread", "disinformation campaign",
  "information manipulation", "academic dishonesty",

  # information and media — POSITIVE
  "content moderation", "fact checking", "information access",
  "knowledge dissemination", "scientific communication",

  # environmental — NEGATIVE
  "energy consumption", "carbon footprint", "resource consumption",
  "environmental degradation",

  # environmental — POSITIVE
  "energy efficiency", "climate solutions", "resource efficiency",
  "environmental monitoring",

  # decision and justice — NEGATIVE
  "criminal justice bias", "unfair treatment", "rights violations",
  "mass surveillance", "social control",

  # decision and justice — POSITIVE
  "decision making", "criminal justice", "human dignity",
  "fundamental rights", "equal treatment", "evidence-based decision",

  # existential and systemic — NEGATIVE
  "existential risk", "catastrophic risk", "systemic risk",
  "long-term risk", "unintended consequences",

  # existential and systemic — POSITIVE
  "long-term benefits", "systemic improvements", "transformative potential",
  "future opportunities",

  # adoption and use — NEGATIVE
  "misuse potential", "unsafe deployment", "irresponsible use",

  # adoption and use — POSITIVE
  "responsible use", "safe use", "ethical deployment",
  "beneficial application"
)

# 2-word effect-leading prefixes (for trigram Pattern B & fourgram Pattern A)
effect_side_2w_pre <- c(
  # outcome-of constructions (neutral)
  "effects of", "impact of", "impacts of", "consequences of",
  "implications of", "outcomes of", "influence of", "repercussions of",
  "ramifications of", "externalities of",

  # valence-of constructions — NEGATIVE
  "risks of", "harms of", "dangers of", "threats of", "hazards of",
  "misuse of", "abuse of", "limitations of", "failures of", "pitfalls of",
  "disruption by", "disruption from", "downsides of", "costs of",

  # valence-of constructions — POSITIVE
  "benefits of", "opportunities of", "advantages of", "potential of",
  "promise of", "gains from", "strengths of", "value of",
  "utility of", "contributions of", "advances in", "progress in",

  # neutral evaluative constructions
  "challenges of", "tradeoffs of", "concerns about", "debates about",
  "discourse on",

  # governance and responsibility (neutral)
  "governance of", "regulation of", "oversight of", "ethics of",
  "accountability for", "responsibility for", "liability for",
  "standards for", "audit of",

  # adoption and use (neutral)
  "adoption of", "use of", "deployment of", "integration of",
  "reliance on", "dependence on",

  # trust and perception — NEGATIVE
  "distrust of", "mistrust of", "fears about", "anxiety about",
  "skepticism about", "resistance to",

  # trust and perception — POSITIVE
  "trust in", "optimism about", "enthusiasm about",
  "attitudes toward", "attitudes towards", "perceptions of", "acceptance of",
  "awareness of",

  # threat constructions — NEGATIVE
  "threats to", "dangers to", "harm from", "damage from", "risks to",

  # opportunity constructions — POSITIVE
  "hope for", "support for", "case for",

  # role constructions (neutral)
  "role of", "power of"
)

# 3-word effect-leading prefixes (for fourgram Pattern C: [effect prefix 3w] + [AI 1w])
effect_side_3w_pre <- c(
  # ethical / social / economic implications of — NEGATIVE
  "ethical implications of", "social implications of",
  "economic implications of", "societal implications of",
  "political implications of", "legal implications of",
  "ethical challenges of", "social challenges of", "economic challenges of",
  "ethical concerns about", "social concerns about", "safety concerns about",

  # ethical / social / economic implications of — POSITIVE
  "ethical opportunities of", "social benefits of",
  "economic benefits of", "societal benefits of",
  "transformative potential of", "positive implications of",
  "innovative potential of",

  # impact on domain constructions — NEGATIVE
  "impact on inequality", "impact on privacy", "impact on surveillance",
  "impact on misinformation",

  # impact on domain constructions — POSITIVE
  "impact on society", "impact on democracy", "impact on employment",
  "impact on healthcare", "impact on education", "impact on labor",
  "impact on security", "impact on creativity", "impact on wellbeing",
  "impact on journalism", "impact on science", "impact on research",
  "impact on productivity",

  # social / economic / societal impact of — NEGATIVE
  "negative impact of", "harmful effects of", "adverse impact of",

  # social / economic / societal impact of — POSITIVE
  "social impact of", "economic impact of", "societal impact of",
  "environmental impact of", "positive impact of", "beneficial impact of",

  # public and societal framing — NEGATIVE
  "societal risks of", "societal harms of",

  # public and societal framing — POSITIVE
  "public perception of", "societal effects of", "societal benefits of",
  "societal contributions of",

  # risk / benefit framing (inherently balanced pairs)
  "risks and benefits", "benefits and risks", "risks and opportunities",
  "threats and opportunities", "costs and benefits",
  "risks associated with", "effects associated with",
  "impact associated with", "challenges associated with",
  "opportunities associated with", "concerns associated with",
  "harms associated with", "benefits associated with",
  "dangers associated with", "implications associated with",

  # potential framing — NEGATIVE
  "potential risks of", "potential harms of", "potential dangers of",

  # potential framing — POSITIVE
  "potential benefits of", "potential impact of", "potential effects of",
  "potential implications of", "potential value of",

  # responsible / ethical use framing — NEGATIVE
  "misuse and abuse", "unsafe deployment of",

  # responsible / ethical use framing — POSITIVE
  "responsible use of", "ethical use of", "safe use of",
  "responsible deployment of", "beneficial use of",

  # governance and policy framing (neutral)
  "governance and regulation", "policy implications of",
  "regulation and governance", "ethical governance of",
  "regulatory challenges of", "policy challenges of",

  # bias / fairness / privacy framing (neutral)
  "bias and fairness", "bias and discrimination",
  "privacy and security", "privacy and surveillance",
  "fairness and accountability", "transparency and accountability",
  "trust and accountability", "safety and security",

  # threat and opportunity framing — NEGATIVE
  "threats posed by", "risks posed by", "dangers posed by",
  "harms caused by", "disruption caused by",

  # threat and opportunity framing — POSITIVE
  "opportunities presented by", "benefits offered by",
  "value created by", "progress enabled by",

  # labour and displacement framing — NEGATIVE
  "labor market disruption", "displacement caused by",

  # labour and displacement framing — POSITIVE
  "labor market effects", "labor market impact", "labor market consequences",
  "automation driven by", "growth driven by",

  # temporal framing — NEGATIVE
  "long-term risks of", "short-term risks of",

  # temporal framing — POSITIVE
  "long-term effects of", "short-term effects of",
  "long-term impact of", "short-term impact of",
  "long-term benefits of", "long-term implications of",

  # unintended / adverse vs. intended / positive framing
  "unintended consequences of", "adverse effects of",
  "negative consequences of", "negative effects of",
  "positive effects of", "positive impact of",
  "intended benefits of",

  # public debate and discourse framing (neutral)
  "public debate on", "public discourse on",
  "debate about the", "concerns regarding the",
  "questions about the"
)

# COMBINATION STEPS

# ~1,800 bigrams
# [AI 1w] + [effect 1w]  AND  [effect 1w] + [AI 1w]
ai_sentiment_bigrams <- unique(trimws(c(
  as.vector(outer(ai_side_1w,     effect_side_1w, paste)),
  as.vector(outer(effect_side_1w, ai_side_1w,     paste))
)))

# ~5,400 trigrams
# Pattern A: [AI 2w]            + [effect 1w]
# Pattern B: [effect prefix 2w] + [AI 1w]
trigrams_a <- as.vector(outer(ai_side_2w,         effect_side_1w, paste))
trigrams_b <- as.vector(outer(effect_side_2w_pre, ai_side_1w,     paste))
ai_sentiment_trigrams <- unique(trimws(c(trigrams_a, trigrams_b)))

# ~10,200 fourgrams
# Pattern A: [effect prefix 2w] + [AI 2w]    — e.g. "risks of machine learning"
# Pattern B: [AI 2w]            + [effect 2w] — e.g. "machine learning job displacement"
# Pattern C: [effect prefix 3w] + [AI 1w]    — e.g. "ethical implications of ai"
# Pattern D: [AI 3w]            + [effect 1w] — e.g. "large language model risks"
fourgrams_a <- as.vector(outer(effect_side_2w_pre, ai_side_2w,    paste))
fourgrams_b <- as.vector(outer(ai_side_2w,         effect_side_2w, paste))
fourgrams_c <- as.vector(outer(effect_side_3w_pre, ai_side_1w,    paste))
fourgrams_d <- as.vector(outer(ai_side_3w,         effect_side_1w, paste))
ai_sentiment_fourgrams <- unique(trimws(c(fourgrams_a, fourgrams_b, fourgrams_c, fourgrams_d)))
```

```{r}
# Creating the subcorpus of sentences that explicitly talk about AI  
# Base sentences from AI-related abstracts
target_sentences_base <- ai_abstracts |>
  unnest_tokens(sentence, abstract, token = "sentences", drop = FALSE) |>
  group_by(id) |>
  mutate(sentence_id = row_number()) |>
  ungroup() |>
  select(id, publication_year, sentence_id, sentence, abstract)

# Bigram hits in sentences
target_hits_bigrams <- target_sentences_base |>
  unnest_tokens(bigram, sentence, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  filter(bigram %in% ai_sentiment_bigrams) |>
  transmute(id, publication_year, sentence_id)

# Trigram hits in sentences
target_hits_trigrams <- target_sentences_base |>
  unnest_tokens(trigram, sentence, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  filter(trigram %in% ai_sentiment_trigrams) |>
  transmute(id, publication_year, sentence_id)

# Fourgram hits in sentences
target_hits_fourgrams <- target_sentences_base |>
  unnest_tokens(fourgram, sentence, token = "ngrams", n = 4) |>
  filter(!is.na(fourgram)) |>
  filter(fourgram %in% ai_sentiment_fourgrams) |>
  transmute(id, publication_year, sentence_id)

# Combining all target-specific hits
target_sentence_hits <- bind_rows(
  target_hits_bigrams,
  target_hits_trigrams,
  target_hits_fourgrams
) |>
  distinct(id, publication_year, sentence_id)

# Keeping only target-specific sentences
ai_target_sentences <- target_sentences_base |>
  semi_join(target_sentence_hits, by = c("id", "publication_year", "sentence_id"))

ai_target_sentences
```

This target-specific filtering step reduces drastically the number of selected sentences to 266. The Bing analysis is carried out using 212 sentences and in AFINN using 190, which may limit the generalization of the results.

```{r}
# We create a sentence-level id 
ai_target_sentences_ids <- ai_target_sentences |>
  mutate(sentence_uid = paste(id, sentence_id, sep = "_"))

# Total sentences that are target-specific
ai_target_sentences_ids |>
  summarise(n_ai_target_sentences = n_distinct(sentence_uid))

# Target-specific sentences that have at least one word in the lexicon Bing 
ai_target_sentences_ids |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  summarise(n_bing_scored_sentences = n_distinct(sentence_uid))

# Target-specific sentences that have at least one word in the lexicon AFINN 
ai_target_sentences_ids |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  summarise(n_afinn_scored_sentences = n_distinct(sentence_uid))
```

```{r}
# In this chunk we create two tibbles, one per lexicon, that contain the id of the abstract and sentence, the publication year and the sentiment variables related to each lexicon.

# Bing sentiment by target-specific sentence
bing_target_sentence_sentiment <- ai_target_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  count(id, publication_year, sentence_id, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(
    net_sentiment = positive - negative,
    total_sentiment_words = positive + negative,
    net_sentiment_norm = if_else(
      total_sentiment_words > 0,
      (positive - negative) / total_sentiment_words,
      NA_real_))

bing_target_sentence_sentiment

# AFINN sentiment by target-specific sentence
afinn_target_sentence_sentiment <- ai_target_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  group_by(id, publication_year, sentence_id) |>
  summarise(
    afinn_sum = sum(value, na.rm = TRUE),
    afinn_mean = mean(value, na.rm = TRUE),
    n_sentiment_words = n(),
    .groups = "drop")

afinn_target_sentence_sentiment
```

```{r}
# Target-specific sentence-level sentiment aggregated back to abstract level

# BING: average target-specific sentence sentiment within each abstract
bing_target_sentence_to_abstract <- bing_target_sentence_sentiment |>
  group_by(id, publication_year) |>
  summarise(
    sentiment_score = mean(net_sentiment_norm, na.rm = TRUE),
    n_target_sentences = n(),
    .groups = "drop"
  ) |>
  mutate(lexicon = "Bing")

# AFINN: average target-specific sentence sentiment within each abstract
afinn_target_sentence_to_abstract <- afinn_target_sentence_sentiment |>
  group_by(id, publication_year) |>
  summarise(
    sentiment_score = mean(afinn_mean, na.rm = TRUE),
    n_target_sentences = n(),
    .groups = "drop"
  ) |>
  mutate(lexicon = "AFINN")

# Combining both lexicons
target_sentence_abstract_scores <- bind_rows(
  bing_target_sentence_to_abstract,
  afinn_target_sentence_to_abstract)

# Calculating the annual average sentiment by lexicon 
target_sentence_abstract_yearly <- target_sentence_abstract_scores |>
  group_by(lexicon, publication_year) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop")
```

```{r}
# Plotting the evolution of the annual average sentiment score (using target-specific sentences): line chart 
#| fig-width: 7
#| fig-height: 5

ggplot(
  target_sentence_abstract_yearly,
  aes(x = publication_year, y = mean_sentiment, color = lexicon)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 2016:2025,
    labels = 2016:2025
  ) +
  scale_color_manual(values = c("Bing" = "#2C3E6B", "AFINN" = "#8E44AD")) +
  labs(
    title = "Target-specific sentiment in AI-related abstracts over time",
    subtitle = "Yearly mean sentiment per abstract based on target-specific AI-effect sentences",
    x = NULL,
    y = "Mean sentiment score",
    color = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "top",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank()
  )

# Plotting the comparison before-after the rise of AI of the annual avergae sentiment score (using target-specific sentences): boxplot 

target_sentence_period_scores <- target_sentence_abstract_scores |>
  mutate(
    period = if_else(
      publication_year < 2023,
      "Before rise of AI",
      "After rise of AI"))

target_sentence_period <- target_sentence_period_scores |>
  group_by(lexicon, period) |>
  summarise(
    mean_sentiment = mean(sentiment_score, na.rm = TRUE),
    median_sentiment = median(sentiment_score, na.rm = TRUE),
    n_abstracts = n(),
    .groups = "drop"
  )

ggplot(
  target_sentence_period_scores,
  aes(x = period, y = sentiment_score, fill = period)
) +
  geom_boxplot(alpha = 0.85, width = 0.6, outlier.alpha = 0.2) +
  scale_fill_manual(values = c(
    "Before rise of AI" = "steelblue",
    "After rise of AI"  = "#2C3E6B"
  )) +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Sentiment in AI-related abstracts before and after generative AI",
    subtitle = "Based on target-specific AI-effect sentences only",
    x = NULL,
    y = "Sentiment score"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position      = "none",
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    panel.grid.minor     = element_blank()
  )
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the number of positive, negative, and neutral AI-related abstracts by year (using target-specific sentences)

# Bing
bing_target_sentence_year_polarity <- bing_target_sentence_to_abstract |>
  mutate(
    polarity = case_when(
      sentiment_score > 0 ~ "Positive",
      sentiment_score < 0 ~ "Negative",
      TRUE ~ "Neutral"
    )
  ) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "Bing")

# AFINN
afinn_target_sentence_year_polarity <- afinn_target_sentence_to_abstract |>
  mutate(
    polarity = case_when(
      sentiment_score > 0 ~ "Positive",
      sentiment_score < 0 ~ "Negative",
      TRUE ~ "Neutral"
    )
  ) |>
  count(publication_year, polarity) |>
  mutate(lexicon = "AFINN")

# Combine both lexicons
target_sentence_year_polarity <- bind_rows(
  bing_target_sentence_year_polarity,
  afinn_target_sentence_year_polarity
)

ggplot(
  target_sentence_year_polarity,
  aes(x = factor(publication_year), y = n, fill = polarity)
) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "Positive" = "#27AE60",
    "Negative" = "#C0392B",
    "Neutral"  = "#2C3E6B"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_flip() +
  facet_wrap(~ lexicon, ncol = 2, scales = "free_y") +
  labs(
    title = "Number of positive, negative, and neutral AI-related abstracts by year",
    subtitle = "Based on target-specific AI-effect sentences only",
    x = NULL,
    y = "Number of abstracts",
    fill = "Overall sentiment"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title           = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle        = element_text(color = "#4a4a6a"),
    strip.text           = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x          = element_text(color = "#4a4a6a"),
    axis.text.y          = element_text(color = "#4a4a6a"),
    legend.position      = "bottom",
    legend.text          = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor     = element_blank(),
    panel.grid.major.y   = element_blank()
  )
```

### 5.4 Which words drive the sentiment in target-specific AI-effect sentences?

Finally, we explore which words drive the sentiment in these target-specific AI-effect sentences of scientific papers' abstracts. We count the frequency of positive and negative words by lexicon within the selected sentences and we can observe agian the pattern that, in both lexicons, positive words contribute more to the overall sentiment than negative ones. On the positive side, words such as trust, ethical and intelligence and verbs like support, enhance and promote appear very often. On the negative side, words like bias, risks, distrust, critical, issues or negative are the most common in both lexicons. This set of terms provide insights into how artificial intelligence is conceptualized in scientific papers and their abstracts, the social risks and methodological limitations it creates, as well as its beneficial role.

```{r}
# We create a data frame with word, sentiment, and number of mentions to see which words are more common

# Bing 
bing_target_word_counts <- ai_target_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("bing"), by = "word") |>
  count(word, sentiment, sort = TRUE)
bing_target_word_counts

# AFINN 
afinn_target_word_counts <- ai_target_sentences |>
  unnest_tokens(word, sentence) |>
  inner_join(get_sentiments("afinn"), by = "word") |>
  mutate(
    sentiment = case_when(
      value > 0 ~ "Positive",
      value < 0 ~ "Negative",
      TRUE ~ "Neutral")) |>
  filter(sentiment != "Neutral") |>
  count(word, sentiment, sort = TRUE)
afinn_target_word_counts
```

```{r}
#| fig-width: 7
#| fig-height: 5

# Plotting the top contributing words by sentiment by lexicon 

target_word_counts <- bind_rows(
  bing_target_word_counts |>
    mutate(
      lexicon = "Bing",
      sentiment = case_when(
        sentiment == "positive" ~ "Positive",
        sentiment == "negative" ~ "Negative",
        TRUE ~ sentiment)),
  afinn_target_word_counts |>
    mutate(lexicon = "AFINN"))

target_word_counts |>
  group_by(lexicon, sentiment) |>
  slice_max(n, n = 15, with_ties = FALSE) |>
  ungroup() |>
  mutate(panel = paste(lexicon, sentiment, sep = " - ")) |>
  ggplot(aes(x = n, y = reorder(word, n), fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_fill_manual(values = c(
    "Positive" = "#27AE60",
    "Negative" = "#C0392B")) +
  labs(
    title = "Words driving sentiment in target-specific AI-effect sentences",
    subtitle = "Top 15 positive and negative words matched by each lexicon",
    x = "Contribution to sentiment",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title       = element_text(face = "bold", color = "#1a1a2e"),
    plot.subtitle    = element_text(color = "#4a4a6a"),
    strip.text       = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.x      = element_text(color = "#4a4a6a"),
    axis.text.y      = element_text(size = 8, color = "#4a4a6a"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```



## Section 6. IF WE CAN \[RQ5\]

## Section 7. Conclussions and limitations

## Section 8. References

Ding, L., Lawson, C., & Shapira, P. (2025). Rise of generative artificial intelligence in science. *Scientometrics*, 130(9), 5093–5114. https://doi.org/10.1007/s11192-025-05413-z

Liu, N., Shapira, P., & Yue, X. (2021). Tracking developments in artificial intelligence research: Constructing and applying a new search strategy. *Scientometrics*, 126(4), 3153–3192. https://doi.org/10.1007/s11192-021-03868-4

## Appendix I - Domain specific stopword list prompt

The domain-specific stopword list used in Section 3 was generated with the assistance of a large language model (Claude Sonnet 4.6, Anthropic). The complete prompt is reproduced below.

::: {style="border: 1px solid #ccc; padding: 1em; border-radius: 4px;"}
```         
You are an expert computational linguist and text mining specialist with deep 
knowledge of social science research and academic writing conventions. You are 
working on a large-scale corpus analysis project described in the context below.

<project_context>
The project analyzes a corpus of journal abstracts drawn from six leading social 
science journals over the period 2016–2025:

  - Nature Human Behaviour (behavioral science, interdisciplinary)
  - Humanities and Social Sciences Communications (interdisciplinary)
  - American Sociological Review (sociology, North America)
  - American Political Science Review (political science, North America)
  - European Sociological Review (sociology, Europe)
  - European Journal of Political Research (political science, Europe)

The corpus spans the period before and after the public rise of generative AI 
(November 2022, release of ChatGPT). Section 3 of the project (for which these 
stopwords are being prepared) addresses two research questions:

  RQ1: What are the predominant topics in top social science journals, by journal 
       and by year?
  RQ2: Has AI (in the broad sense) become one of the predominant topics, and has 
       it influenced the broader topical landscape?

The unit of analysis is the abstract (short, dense, highly formulaic texts 
averaging 150–250 words), written in academic English.
</project_context>

<corpus_frequency_data>
The file corpus_word_freq.csv is attached. It contains the top words in the corpus 
after removing standard English stopwords  and  numeric tokens, generated directly
from thousands of abstracts using tidytext. Columns:
  - word: the token (lowercase)
  - n: raw frequency count across all abstracts

Use this frequency data as empirical grounding for your stopword list:
  - Words that appear here at high frequency AND carry no thematic/discriminative 
    value are strong candidates for inclusion in the stopword list
  - Do not limit yourself to words in this list — be exhaustive across all 7 
    categories — but treat high-frequency words as a priority checklist to 
    ensure nothing obvious is missed
  - High-frequency words that carry topical signal (per the preservation rules 
    below) must NOT be added to the stopword list even if they rank highly
    (e.g. "policy", "political", "education", "health")
</corpus_frequency_data>

<task>
Generate a comprehensive, categorised list of domain-specific stopwords to be 
removed from this corpus. These are words that:

  (a) carry no thematic or discriminative value for social science research AND
  (b) appear at high frequency across the corpus, AND

Before generating each category, cross-check it against the corpus_frequency_data 
above to ensure all high-frequency uninformative words are captured.

Return the list as a clean, categorised R tibble with a column for the stopword 
and a column for its category, preceded by commented section headers explaining 
the rationale for each category.
</task>

<categories_to_cover>
Generate stopwords across ALL of the following categories. For each category, think 
systematically about what words in that class appear in social science abstracts, words like:

1. ACADEMIC BOILERPLATE VERBS
   Verbs used generically across all abstracts regardless of topic: reporting verbs 
   ("examine", "investigate", "explore", "suggest", "find", "show", "demonstrate", 
   "analyze", "assess", "evaluate", "discuss", "argue", "propose", "address", 
   "consider", "focus", "aim", "seek", "test", "draw", "provide", "offer", 
   "highlight", "reveal", "indicate", "identify", "contribute", "develop"), 
   and their inflected forms where relevant.

2. GENERIC METHODOLOGICAL TERMS
   Words describing research designs, data types, and analytical approaches that 
   are ubiquitous across social science and carry no topical signal: "survey", 
   "questionnaire", "interview", "sample", "dataset",  "longitudinal", "cross-sectional",
   "regression", "model", "framework",  "approach", "method", "methodology",
   "analysis", "data", "measure", "variable", "coefficient", "estimate", "robustness",
   "validity", "reliability", etc.

3. ABSTRACT-STRUCTURAL BOILERPLATE
   Phrases and words used to signal the structure of an abstract itself rather than 
   its content: "paper", "study", "article", "research", "literature", "review", 
   "background", "objective", "purpose", "contribution", "implication", "finding", 
   "result", "conclusion", "evidence", "prior", "existing", "previous", "recent", 
   "current", "present", "based", "using", "used", "given", "among", "across".

4. GENERIC SOCIAL SCIENCE CONCEPTS
   Terms so broadly used in social science that they provide no discriminative 
   topical signal between documents: "society", "social", "individual", "group", 
   "population", "people", "person", "human", "behavior", "behaviour", "attitude", 
   "perception", "factor", "role", "relationship", "impact", "effect", "influence", 
   "context", "level", "process", "system", "structure", "pattern", "change", 
   "difference", "association", "correlation", etc.

5. JOURNAL-SPECIFIC AND PUBLICATION BOILERPLATE
   Words likely introduced by journal formatting, editorial conventions, or abstract 
   templates specific to these six journals: copyright notices, volume/issue markers, 
   DOI fragments, and any words that appear due to systematic metadata leakage.

6. ENCODING ARTEFACTS AND NOISE
   Tokens that result from text extraction, encoding issues, or tokenisation: 
   single characters (except meaningful ones), numeric tokens, garbled unicode 
   fragments, OCR noise, HTML remnants, and other non-linguistic tokens.

7. HIGH-FREQUENCY FUNCTION WORDS NOT CAUGHT BY STANDARD STOPWORD LISTS
   Domain-specific function-like words that standard English stopword lists 
   (e.g., the Snowball or SMART lists) miss but which are uninformative in this 
   corpus: discipline-specific connectives, hedging phrases reduced to single tokens 
   ("however", "furthermore", "moreover", "nevertheless", "thus", "therefore", 
   "hence", "whereas", "whereby", "therein"), and academic filler adjectives 
   ("important", "significant", "key", "major", "critical", "novel", "robust", 
   "comprehensive", "systematic", "rigorous", "substantial").
</categories_to_cover>

<preservation_rules>
CRITICAL: The following word classes must NOT appear in the stopword list, even if 
they are frequent. Removing them would destroy the thematic signal needed for RQ1 
and RQ2:

  - AI and technology terms: any term related to artificial intelligence, machine 
    learning, large language models, specific model names, or related technology. 
    These are central to RQ2 and must be preserved.
  - Substantive social science topic terms: words indexing real research topics 
    such as "inequality", "democracy", "migration", "climate", "health", "labor", 
    "race", "gender", "education", "voting", "polarization", "trust", "identity", 
    "discrimination", "poverty", "welfare", etc.
  - Sentiment-bearing terms: words with clear positive or negative valence that 
    will be needed for the sentiment analysis in Section 5.
  - Proper nouns and named entities: country names, institutional names, author 
    names — these may carry topical signal.
  - Temporally specific terms: words that may index pre/post ChatGPT content, 
    which are needed for the temporal analyses in Sections 3–6.

When in doubt about whether a word carries signal, err on the side of keeping it.
</preservation_rules>

<output_format>
Return only executable R code. Structure it as follows:

  library(tibble)

  domain_stopwords <- tibble(
    word = c(
      # 1. ACADEMIC BOILERPLATE VERBS
      # These verbs appear in virtually every abstract regardless of topic.
      # Removing them prevents them from dominating topic-word distributions
      # in LDA and collapsing IDF weights in TF-IDF.
      "examine", "investigate", ...

      # 2. GENERIC METHODOLOGICAL TERMS
      # ...

      # [continue for all 7 categories]
    ),
    category = c(
      # mirror the word vector exactly, one category label per word
      "boilerplate_verb", "boilerplate_verb", ...
    )
  )

Rules:
  - All stopwords in lowercase
  - One row per word (not per phrase, we operate on unigrams after tokenisation)
  - The word and category vectors must be exactly the same length
  - Category labels should be short snake_case strings matching the 7 categories
  - Add a final comment line reporting the total count: # Total: N stopwords
  - Be exhaustive, put lots of effort
</output_format>
```
:::

## Appendix II - AI effects lexicon prompt

::: {style="border: 1px solid #ccc; padding: 1em; border-radius: 4px;"}
```         
You are an expert computational linguist and text mining specialist working on a 
systematic literature review about AI and its effects. Your task is to create three 
R vectors of n-grams that will be used to identify academic journal abstracts addressing artificial intelligence (AI) or its effects, consequences, risks, 
benefits, or impacts.

<objective>
Precision over recall. We accept missing relevant abstracts, but every abstract 
matched by these n-grams must unambiguously be about AI or its effects.
</objective>

<existing_ai_lexicon>
We already have a lexicon that identifies AI-related vocabulary. Use it as your 
reference for how AI is referred to across our corpus. Your component vectors must 
be compatible with this lexicon —meaning the AI-side tokens should draw from these 
terms or their natural variants—.

# AI unigram lexicon
ai_lexicon_unigrams <- c(
  "autoencoder", "perceptron", "neurocomputing",
  "chatgpt", "llm",
  "deepseek", "openai", "ai"
)
# AI bigram lexicon
ai_lexicon_bigrams <- c(
  "artificial intelligence", "neural network", "machine learning",
  "deep learning", "reinforcement learning", "learning algorithm",
  "backpropagation learning", "backpropagation algorithm",
  "machine intelligence", "computational intelligence",
  "neural model", "deep network",
  "generative ai", "language model", "pretrained transformer",
  "ai-generated content", "openai gpt",
  "claude sonnet", "claude opus", "claude haiku",
  "google gemini", "gemini pro", "gemini ultra",
  "microsoft copilot", "github copilot", "mistral ai",
  "meta llama", "llama 2", "llama 3",
  "stable diffusion", "gpt 5", "gpt 4", "gpt 3", "gpt 2", "gpt 1",
  "dall e", "agentic ai"
)
# AI trigram lexicon
ai_lexicon_trigrams <- c(
  "large language model", "generative language model",
  "generative large language", "generative pretrained transformer",
  "generative artificial intelligence", "artificial intelligence generated",
  "openai large language", "openai language model",
  "ai language model", "ai generated content"
)
</existing_ai_lexicon>

<task>
Rather than writing every n-gram by hand, use a **combinatorial approach**: define 
separate component vectors for the AI side and the effects side, then generate the 
final n-gram vectors programmatically using outer() and paste(). This makes the 
lexicon scalable and easy to maintain.
The architecture for each n-gram size is as follows:
## BIGRAMS (2 words total)
Combine:
  - ai_side_1w: AI unigrams (1-word AI terms, e.g. "ai", "llm", "chatgpt")
  - effect_side_1w: effect unigrams (1-word effect/outcome terms, e.g. "risks", 
    "benefits")
Generate with:
  ai_sentiment_bigrams <- unique(trimws(as.vector(
    outer(ai_side_1w, effect_side_1w, paste)
  )))
Note: outer() by default produces [AI term] + [effect term] order. To also capture
[effect term] + [AI term] order (e.g. "risks ai", "benefits llm"), generate both 
directions and combine:
  ai_sentiment_bigrams <- unique(trimws(c(
    as.vector(outer(ai_side_1w, effect_side_1w, paste)),
    as.vector(outer(effect_side_1w, ai_side_1w, paste))
  )))
## TRIGRAMS (3 words total)
Trigrams can be formed from two structural patterns. Generate both and combine:
Pattern A — [2-word AI term] + [1-word effect term]:
  Uses: ai_side_2w (AI bigrams) × effect_side_1w (effect unigrams)
  Example: "machine learning" + "risks" → "machine learning risks"
Pattern B — [2-word effect prefix] + [1-word AI term]:
  Define effect_side_2w_pre as 2-word effect-leading phrases like 
  "effects of", "risks of", "impact of", etc., then combine with ai_side_1w:
  Example: "effects of" + "ai" → "effects of ai"
  effect_side_2w_pre <- c("effects of", "impact of", "risks of", ...)
  
  # Pattern A: [2-word AI term] + [1-word effect term]
  trigrams_a <- as.vector(outer(ai_side_2w, effect_side_1w, paste))
  
  # Pattern B: [2-word effect prefix] + [1-word AI term]
  trigrams_b <- as.vector(outer(effect_side_2w_pre, ai_side_1w, paste))
  
  ai_sentiment_trigrams <- unique(trimws(c(trigrams_a, trigrams_b)))
  
## FOURGRAMS (4 words total)
Fourgrams can be formed from three structural patterns. Generate all and combine:
Pattern A — [2-word effect prefix] + [2-word AI term]:
  effect_side_2w_pre × ai_side_2w
  Example: "effects of" + "machine learning" → "effects of machine learning"
Pattern B — [2-word AI term] + [2-word effect term]:
  ai_side_2w × effect_side_2w
  Example: "machine learning" + "job displacement" → 
           "machine learning job displacement"
Pattern C — [3-word effect prefix] + [1-word AI term]:
  effect_side_3w_pre × ai_side_1w
  Example: "ethical implications of" + "ai" → "ethical implications of ai"
  fourgrams_a <- as.vector(outer(effect_side_2w_pre, ai_side_2w, paste))
  fourgrams_b <- as.vector(outer(ai_side_2w, effect_side_2w, paste))
  fourgrams_c <- as.vector(outer(effect_side_3w_pre, ai_side_1w, paste))
  ai_sentiment_fourgrams <- unique(trimws(c(fourgrams_a, fourgrams_b, fourgrams_c)))
</task>

<component_vector_requirements>
Define all component vectors clearly before the combination step, with inline 
comments grouping terms by semantic category. All strings must be lowercase.
The effects/outcome side vectors must cover a broad semantic field, including:
- Direct outcome terms: effects, impacts, consequences, implications, outcomes, 
  influence
- Valence terms: benefits, risks, harms, dangers, threats, opportunities, 
  advantages, disadvantages, challenges (IMPORTANT: make sure that non-neutral terms are balanced, that is, make sure there is no bias towards "positive" or "negative" valence)
- Societal dimensions: ethical, social, economic, labor, employment, inequality, 
  bias, fairness, privacy, security, trust, accountability, transparency
- Governance terms: regulation, policy, governance, legislation, oversight
- Human response terms: adoption, acceptance, resistance, perception, attitudes, 
  concerns, fears
- Domain-specific effect nouns: automation, misinformation, discrimination, 
  creativity, decision-making, displacement, surveillance
Use ALL terms from the existing lexicon on the AI side —include every term across 
unigrams, bigrams, and trigrams without exception—.
</component_vector_requirements>

<precision_rules>
The only constraint on generated combinations is semantic unambiguity: if a 
combination appears in an abstract, it must mean that AI or its effects addressed. Whether a combination actually occurs in the corpus is irrelevant —unmatched n-grams are harmless—. Go for exhaustive coverage. Do not prune for 
plausibility. 
Apply unique() and trimws() to deduplicate and clean whitespace.
</precision_rules>

<output_format>
Return only executable R code. Structure it in this order:
1. All component vectors (with inline comments by semantic category)
2. The three combination steps producing ai_sentiment_bigrams, 
   ai_sentiment_trigrams, ai_sentiment_fourgrams
3. A brief comment above each final vector stating how many n-grams 
   it contains, e.g.: # ~120 bigrams
No prose before or after the code block.
</output_format>
```
:::

## Appendix III - Unigram TF-IDF context windows

Some of the terms in the TF-IDF results were unclear: they were abbreviations or unexpected terms with no immediately evident interpretation. In order to understand them better, we explored their context windows using the `kwic` function from the `quanteda` package. We include them here for transparency and to allow the reader to verify our interpretations.

```{r}
# Create the corpus, linking text to our ID column
abstracts_corpus <- corpus(articles,
                           text_field = "abstract",
                           docid_field = "id")
```


```{r}
# Year 2016
# kwic = 400th
# We specify `quanteda::tokens` because there are conflicts with other libraries
kwic(quanteda::tokens(abstracts_corpus), "400th", window = 8)
# kwic = geg
kwic(quanteda::tokens(abstracts_corpus), "geg", window = 8)
# kwic = mlg
kwic(quanteda::tokens(abstracts_corpus), "mlg", window = 8)

# Year 2017
# kwic = gothic
kwic(quanteda::tokens(abstracts_corpus), "gothic", window = 8)
# kwic = fia
kwic(quanteda::tokens(abstracts_corpus), "fia", window = 8)

# Year 2018
# kwic = graffiti
kwic(quanteda::tokens(abstracts_corpus), "graffiti", window = 8)
# kwic = revok
kwic(quanteda::tokens(abstracts_corpus), "revok", window = 8)
# kwic = tves
kwic(quanteda::tokens(abstracts_corpus), "tves", window = 8)
# kwic = gamh
kwic(quanteda::tokens(abstracts_corpus), "gamh", window = 8)
# kwic = nua
kwic(quanteda::tokens(abstracts_corpus), "nua", window = 8)

# Year 2019
# kwic = gun
kwic(quanteda::tokens(abstracts_corpus), "gun", window = 8)
# kwic = guns
kwic(quanteda::tokens(abstracts_corpus), "guns", window = 8)
# kwic = lgds
kwic(quanteda::tokens(abstracts_corpus), "lgds", window = 8)
# kwic = ccs
kwic(quanteda::tokens(abstracts_corpus), "ccs", window = 8)

# Year 2020
# kwic = pea
kwic(quanteda::tokens(abstracts_corpus), "pea", window = 6)
# kwic = mta
kwic(quanteda::tokens(abstracts_corpus), "mta", window = 6)
# kwic = silcrete
kwic(quanteda::tokens(abstracts_corpus), "silcrete", window = 8)

# Year 2021
# kwic = coconut
kwic(quanteda::tokens(abstracts_corpus), "coconut", window = 8)
# kwic = hcws
kwic(quanteda::tokens(abstracts_corpus), "hcws", window = 6)
# kwic = wc
kwic(quanteda::tokens(abstracts_corpus), "wc", window = 6)

# Year 2023
# kwic = csr
kwic(quanteda::tokens(abstracts_corpus), "csr", window = 6)
# kwic = dms
kwic(quanteda::tokens(abstracts_corpus), "dms", window = 6)
# kwic = smmes
kwic(quanteda::tokens(abstracts_corpus), "smmes", window = 6)

# Year 2024-2025
# kwic = esg
kwic(quanteda::tokens(abstracts_corpus), "esg", window = 6)
# kwic = efl
kwic(quanteda::tokens(abstracts_corpus), "efl", window = 6)
# kwic = fintech
kwic(quanteda::tokens(abstracts_corpus), "fintech", window = 6)
```

## Appendix IV - Bigram TF-IDF context windows

Some of the terms in the TF-IDF results were unclear: they were abbreviations or unexpected terms with no immediately evident interpretation. In order to understand them better, we explored their context windows using the `kwic` function from the `quanteda` package. We include them here for transparency and to allow the reader to verify our interpretations.

```{r}
# Year 2016
# kwic = european sociological (this is metadata that we cannot filter without
# potentially losing relevant information, since the bigram is composed by
# two terms that could be meaninful in other combinations)
kwic(quanteda::tokens(abstracts_corpus), phrase("european sociological"), window = 8)

# Year 2017
# kwic = social pact
kwic(quanteda::tokens(abstracts_corpus), phrase("social pact"), window = 8)
# kwic = veto player
kwic(quanteda::tokens(abstracts_corpus), phrase("veto player"), window = 8)

# Year 2018
# kwic = mind upload
kwic(quanteda::tokens(abstracts_corpus), phrase("mind upload"), window = 8)
# kwic = fairy tale
kwic(quanteda::tokens(abstracts_corpus), phrase("fairy tale"), window = 8)

# Year 2020
# kwic = science diplomacy
kwic(quanteda::tokens(abstracts_corpus), phrase("science diplomacy"), window = 8)
# kwic = grass pea
kwic(quanteda::tokens(abstracts_corpus), phrase("grass pea"), window = 8)
# kwic = gun culture
kwic(quanteda::tokens(abstracts_corpus), phrase("gun culture"), window = 8)
# kwic = agile sd
kwic(quanteda::tokens(abstracts_corpus), phrase("agile sd"), window = 8)
# kwic = uhi tools
kwic(quanteda::tokens(abstracts_corpus), phrase("uhi tools"), window = 8)
# kwic = sa mta
kwic(quanteda::tokens(abstracts_corpus), phrase("sa mta"), window = 8)

# Year 2021
# kwic = frontline hcws
kwic(quanteda::tokens(abstracts_corpus), phrase("frontline hcws"), window = 8)

# Year 2023
# kwic = burden reduction
kwic(quanteda::tokens(abstracts_corpus), phrase("burden reduction"), window = 8)

# Year 2025
kwic(quanteda::tokens(abstracts_corpus), phrase("carbon emissions"), window = 8)
```


## Appendix V - Trigram TF-IDF context windows

Some of the terms in the TF-IDF results were unclear: they were abbreviations or unexpected terms with no immediately evident interpretation. In order to understand them better, we explored their context windows using the `kwic` function from the `quanteda` package. We include them here for transparency and to allow the reader to verify our interpretations.

```{r}
# Year 2016
# kwic = balance lifestyle mental
kwic(quanteda::tokens(abstracts_corpus), phrase("balance lifestyle mental"), window = 8)

# Year 2017
# kwic = veto player theory
kwic(quanteda::tokens(abstracts_corpus), phrase("veto player theory"), window = 8)

# Year 2018
# kwic = search query data
kwic(quanteda::tokens(abstracts_corpus), phrase("search query data"), window = 8)
# kwic = axial age concept
kwic(quanteda::tokens(abstracts_corpus), phrase("axial age concept"), window = 8)

# Year 2019
# kwic = eclipse prediction scheme
kwic(quanteda::tokens(abstracts_corpus), phrase("eclipse prediction scheme"), window = 8)
# kwic = zika conspiracy theories
kwic(quanteda::tokens(abstracts_corpus), phrase("zika conspiracy theories"), window = 8)

# Year 2020
# kwic = grass pea production
kwic(quanteda::tokens(abstracts_corpus), phrase("grass pea production"), window = 8)
# kwic = agile sd implementation
kwic(quanteda::tokens(abstracts_corpus), phrase("agile sd implementation"), window = 8)
# kwic = probability ratio index
kwic(quanteda::tokens(abstracts_corpus), phrase("probability ratio index"), window = 8)

# Year 2021
# kwic = dravidian tooth word
kwic(quanteda::tokens(abstracts_corpus), phrase("dravidian tooth word"), window = 8)
# kwic = proto dravidian tooth
kwic(quanteda::tokens(abstracts_corpus), phrase("proto dravidian tooth"), window = 8)
# kwic = av pilot projects
kwic(quanteda::tokens(abstracts_corpus), phrase("av pilot projects"), window = 8)
# kwic = short run sustainability
kwic(quanteda::tokens(abstracts_corpus), phrase("short run sustainability"), window = 8)

# Year 2022
# kwic = golf simulation industry
kwic(quanteda::tokens(abstracts_corpus), phrase("golf simulation industry"), window = 8)
# kwic = significantly lower sle
kwic(quanteda::tokens(abstracts_corpus), phrase("significantly lower sle"), window = 8)

# Year 2024
# kwic = squares structural equation
kwic(quanteda::tokens(abstracts_corpus), phrase("squares structural equation"), window = 8)
```

