---
title: "Text Mining Final Project"
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 4
    toc-title: "Table of Contents"
    number-sections: true
    embed-resources: true
    df-print: paged
editor: visual
---

## Section 1. Introduction

**Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final** **Está en el google doc, incluirla al final**

## Section 2. Setup, data loading, and preprocessing [SI NO HACEMOS EL PREPROCESSING AQUÍ, QUITARLO DEL TÍTULO]

### Packages

We begin by loading all R packages required throughout this project.

```{r}
#| message: false
#| warning: false

# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS # REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS# REVISAR SI USAMOS MÁS O SI HAY ALGUNO QUE NO USAMOS

library(tidyverse)
library(tidytext)
library(textdata)
library(sentimentr)
library(topicmodels)
library(quanteda)
library(quanteda.textplots)
library(textstem)
library(igraph)
library(ggraph)
library(widyr)
library(reshape2)
library(RColorBrewer)
library(wordcloud)
library(explore)
library(patchwork)
```

### Loading the data

Our corpus was downloaded from [OpenAlex](https://openalex.org/). We created a custom R script (`openalex_download.R`, included in this project folder, along with documentation) to bulk-download article data from our target journals.

Having run the script, we load the `.RData` file, containing all data structures exactly as they were returned by the OpenAlex API.

```{r}
# Load the RData file. This creates the object `articles_raw` in the environment
load("data/articles.RData")
```

### Initial Inspection

We begin by examining the structure of the raw dataset: variable types, missing values, and basic summary statistics.

```{r}
# Overview of all variables: types, missing values, and distributions
describe(articles_raw)
```

Not all records include abstracts (19.6% of the observations do not include it).Since our entire analysis is based on abstract text, we remove records without abstracts and restrict the corpus to our study period (2016–2025).

```{r}
# Filter: keep only records with an abstract, within our study period
articles <- articles_raw |> 
  filter(!is.na(abstract)) |>
  filter(publication_year >= 2016, publication_year <= 2025)

# Let's observe our new filtered data:
describe(articles)
```

Our new data passed from 12589 to 9484 articles. We have 44 columns from which we will later select only the relevant ones for our analysis.

We explore the distribution of articles by journal and year.

```{r}
# Articles per journal and year
articles |>
  count(source_display_name, publication_year) |>
  arrange(publication_year) |>
  pivot_wider(names_from = source_display_name, values_from = n, values_fill = 0)
```

```{r}
# Merge Palgrave Communications into HSSC — same journal, renamed June 2020
# Full names with abbreviations in parentheses for clarity
articles_merged <- articles |>
  mutate(source_display_name = if_else(
    source_display_name == "Palgrave Communications",
    "Humanities and Social Sciences Communications",
    source_display_name
  )) |>
  mutate(journal_label = case_match(
    source_display_name,
    "American Political Science Review"             ~ "American Political Science Review (APSR)",
    "American Sociological Review"                  ~ "American Sociological Review (ASR)",
    "European Journal of Political Research"        ~ "European Journal of Political Research (EJPR)",
    "European Sociological Review"                  ~ "European Sociological Review (ESR)",
    "Nature Human Behaviour"                        ~ "Nature Human Behaviour (NHB)",
    "Humanities and Social Sciences Communications" ~ "Humanities and Social Sciences Communications (HSSC)"
  ))

# Color palette — keys must match journal_label values exactly
journal_colors <- c(
  "American Political Science Review (APSR)"      = "#2C3E6B",
  "American Sociological Review (ASR)"            = "#C0392B",
  "European Journal of Political Research (EJPR)" = "#8E44AD",
  "European Sociological Review (ESR)"            = "#E67E22",
  "Nature Human Behaviour (NHB)"                  = "#27AE60",
  "Humanities and Social Sciences Communications (HSSC)" = "#2980B9"
)

# Plot 1: one facet per journal, bars by year
# Each panel has its own y-axis scale (scales = "free_y") so within-journal
# trends are visible even for journals with very different volumes
p_facets <- articles_merged |>
  count(journal_label, publication_year) |>
  ggplot(aes(x = publication_year, y = n, fill = journal_label)) +
  geom_col(width = 0.7) +
  facet_wrap(~ journal_label, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(breaks = seq(2016, 2025, by = 3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Publication year", y = "Number of abstracts") +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text       = element_text(size = 9, face = "bold", color = "#1a1a2e"),
    axis.text.x      = element_text(size = 7, color = "#4a4a6a"),
    axis.text.y      = element_text(size = 7, color = "#4a4a6a"),
    axis.title       = element_text(size = 9, color = "#4a4a6a"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#e8e8f0", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", color = NA)
  )

# Plot 2: thin stacked horizontal bar showing corpus composition
# No labels — colors alone carry the journal identity, matched to the facets above
p_bar <- articles_merged |>
  count(journal_label) |>
  mutate(
    pct = n / sum(n),
    journal_label = fct_reorder(journal_label, n)
  ) |>
  ggplot(aes(x = pct, y = 1, fill = journal_label)) +
  geom_col(orientation = "y", width = 0.6) +
  scale_fill_manual(values = journal_colors, guide = "none") +
  scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(x = "Share of total abstracts", y = NULL) +
  theme_minimal(base_family = "serif") +
  theme(
    axis.text.y     = element_blank(),
    axis.text.x     = element_text(size = 7, color = "#4a4a6a"),
    axis.title.x    = element_text(size = 9, color = "#4a4a6a", margin = margin(t = 6)),
    panel.grid      = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  )

# Stack the two plots vertically with patchwork
# heights = c(10, 1) keeps the proportion bar slim relative to the facets
p_facets / p_bar +
  plot_layout(heights = c(10, 1)) +
  plot_annotation(
    title    = "Number of abstracts per year and journal",
    subtitle = "Palgrave Communications merged with HSSC (same journal, renamed June 2020)",
    theme    = theme(
      plot.title      = element_text(size = 14, face = "bold", color = "#1a1a2e",
                                     family = "serif", margin = margin(b = 4)),
      plot.subtitle   = element_text(size = 9, color = "#4a4a6a",
                                     family = "serif", margin = margin(b = 12)),
      plot.caption    = element_text(size = 7, color = "#7a7a9a",
                                     family = "serif", hjust = 0,
                                     margin = margin(t = 10)),
      plot.background = element_rect(fill = "white", color = NA)
    )
  )
```

The six journals in our corpus do not have a balanced share of articles: HSSC represents over 60% of the corpus, while there is relative balanced among the other five journals. Another trend to take into account is related to the temporal distribution: while the first four journals show a relatively balanced distribution of articles by year, the bottom two ---the journals in the *Nature portfolio*--- have exponentially increased the amount of publications in recent years. While this is something we must take into account, we will not exclude any journal from the corpus just for the sake of balance, since they are all reputed and influential journals, excluding one (for example HSSC), would affect the representativeness of the corpus. This is a trade off we are willing to accept, even if many articles are published in the same journal, they are still articles being published in top social science journals, and we should take them into account when answering our research questions. **\[REVISAR ESTE PÁRRAFO; ME PARECE REPETITVO\]**

## Section 3. Topic analysis \[RQ1 + RQ2\]

### Preprocessing

#### Tokenization

We start by keeping the relevant columns: id, publication year, journal, and abstract.

```{r}
# We tokenize at the word level, keeping key metadata columns as identifiers:
tidy_abstracts_raw <- articles |>
  select(id, publication_year, source_display_name, abstract)
```

Then tokenize at the word level:

```{r}
# Tokenize at the word level. Each row will be one token:
tidy_abstracts_raw <- tidy_abstracts_raw |>
  unnest_tokens(
    output = word,
    input  = abstract,
    token  = "words"   # default: lowercase, strips punctuation
  )

# To check that it worked:
slice_head(tidy_abstracts_raw, n = 10 )
```

After tokenizing words, we need to remove all those stopwords that are not useful for analysis, since they have very little informational value.

#### Filtering standard and domain-specific stopwords

We use `stop_words`, a dataframe that includes stopwords from three lexicons (SMART, Snowball, Onix), provided by library `tidy_text` as our standard stopword filter.

```{r}
# Filter stopwords using tidytext `stop_words`
tidy_abstracts <- tidy_abstracts_raw |>
  anti_join(stop_words, by = "word")

tidy_abstracts
```

Academic abstracts also contain high-frequency words that carry no topical meaning and would pollute our analysis, particularly when exploring absolute frequencies

TF-IDF rankings and LDA topics if left in.

Using Claude Sonnet 4.6 to identify and create a list of domain-specif words without informative value ---see full prompt in Appendix I---. We manually reviewed this list, removing words we thought had informational value and adding some other little informative words (like "palgrave" or "collection").

```{r}
# Create the custom domain stopwords
domain_stopwords <- tibble(
  word = c(
    
    # 1. Academic boilerplate verbs
    # Verbs used generically across all abstracts regardless of topic.
    # Includes base forms, 3rd-person singular, past/passive, and gerund inflections.
    "examine", "investigate", "explore", "suggest", "find", "show", "demonstrate", "analyze",
    "analyse", "assess", "evaluate", "discuss", "argue", "propose", "address", "consider",
    "focus", "aim", "seek", "test", "draw", "provide", "offer", "highlight",
    "reveal", "indicate", "identify", "contribute", "develop", "present", "describe", "explain",
    "understand", "improve", "enhance", "employ", "apply", "use", "utilize", "utilise",
    "adopt", "extend", "advance", "inform", "situate", "engage", "establish", "illustrate",
    "document", "note", "observe", "report", "review", "compare", "measure", "estimate",
    "predict", "capture", "synthesize", "synthesise", "integrate", "combine", "incorporate", "build",
    "construct", "formulate", "frame", "conceptualize", "conceptualise", "operationalize", "operationalise", "uncover",
    "shed", "map", "trace", "track", "model", "simulate", "validate", "verify",
    "confirm", "challenge", "question", "reflect", "emphasize", "emphasise", "stress", "acknowledge",
    "recognize", "recognise", "account", "imply", "examines", "investigates", "explores", "suggests",
    "finds", "shows", "demonstrates", "analyses", "analyzes", "assesses", "evaluates", "discusses",
    "argues", "proposes", "addresses", "considers", "focuses", "aims", "seeks", "draws",
    "provides", "offers", "highlights", "reveals", "indicates", "identifies", "contributes", "develops",
    "presents", "describes", "explains", "improves", "enhances", "employs", "applies", "adopts",
    "extends", "informs", "engages", "establishes", "illustrates", "documents", "notes", "observes",
    "reports", "compares", "measures", "estimates", "predicts", "captures", "builds", "constructs",
    "integrates", "combines", "incorporates", "uncovers", "sheds", "maps", "traces", "tracks",
    "models", "validates", "confirms", "challenges", "questions", "reflects", "emphasizes", "emphasises",
    "stresses", "acknowledges", "recognizes", "recognises", "accounts", "implies", "examined", "investigated",
    "explored", "suggested", "found", "shown", "demonstrated", "analyzed", "analysed", "assessed",
    "evaluated", "discussed", "argued", "proposed", "addressed", "considered", "focused", "aimed",
    "sought", "drawn", "provided", "offered", "highlighted", "revealed", "indicated", "identified",
    "contributed", "developed", "presented", "described", "explained", "improved", "enhanced", "employed",
    "applied", "adopted", "extended", "informed", "engaged", "established", "illustrated", "documented",
    "noted", "observed", "reported", "compared", "measured", "estimated", "predicted", "captured",
    "built", "constructed", "integrated", "combined", "incorporated", "uncovered", "mapped", "tracked",
    "modeled", "modelled", "validated", "confirmed", "challenged", "reflected", "emphasized", "emphasised",
    "stressed", "acknowledged", "recognized", "recognised", "examining", "investigating", "exploring", "suggesting",
    "finding", "showing", "demonstrating", "analyzing", "analysing", "assessing", "evaluating", "discussing",
    "arguing", "proposing", "addressing", "considering", "focusing", "aiming", "seeking", "drawing",
    "providing", "offering", "highlighting", "revealing", "indicating", "identifying", "contributing", "developing",
    "presenting", "describing", "explaining", "understanding", "improving", "enhancing", "employing", "applying",
    "adopting", "extending", "informing", "engaging", "establishing", "illustrating", "documenting", "noting",
    "observing", "reporting", "comparing", "measuring", "estimating", "predicting", "capturing", "building",
    "constructing", "integrating", "combining", "incorporating", "uncovering", "shedding", "mapping", "tracing",
    "tracking", "modeling", "modelling", "validating", "confirming", "challenging", "questioning", "reflecting",
    "emphasizing", "emphasising", "stressing", "acknowledging", "recognizing", "recognising",
    
    # 2. Generic methodological terms
    # Words describing research designs, data types, statistical procedures,
    # and analytical approaches ubiquitous across all social science.
    "data", "dataset", "datasets", "survey", "surveys", "questionnaire", "questionnaires", "interview",
    "interviews", "observation", "observations", "fieldwork", "experiment", "experiments", "experimental", "sample",
    "samples", "sampling", "subsample", "subsamples", "respondent", "respondents", "participant", "participants",
    "informant", "informants", "census", "panel", "panels", "cohort", "cohorts", "wave",
    "waves", "method", "methods", "methodology", "methodological", "approach", "approaches", "framework",
    "frameworks", "design", "designs", "procedure", "procedures", "protocol", "protocols", "technique",
    "techniques", "instrument", "instruments", "paradigm", "paradigms", "quantitative", "statistical", "regression",
    "regressions", "multivariate", "bivariate", "univariate", "logistic", "probit", "ols", "coefficient",
    "coefficients", "parameter", "parameters", "estimation", "estimator", "estimators", "specification", "specifications",
    "robustness", "validity", "reliability", "endogeneity", "heterogeneity", "heteroskedasticity", "autocorrelation", "collinearity",
    "multicollinearity", "moderation", "mediation", "mediating", "moderating", "confound", "confounding", "confounder",
    "instrumental", "iv", "equation", "equations", "variance", "covariance", "correlation", "correlations",
    "significance", "p-value", "confidence", "interval", "intervals", "standard", "deviation", "deviations",
    "mean", "median", "mode", "distribution", "distributions", "skewness", "kurtosis", "outlier",
    "outliers", "imputation", "weighting", "weights", "weight", "index", "indices", "indicator",
    "indicators", "score", "scores", "scale", "scales", "latent", "manifest", "loading",
    "loadings", "eigenvalue", "eigenvalues", "goodness", "aic", "bic", "rmsea", "cfi",
    "tli", "srmr", "anova", "manova", "ancova", "bootstrapping", "bootstrap", "simulation",
    "simulations", "bayesian", "frequentist", "likelihood", "maximum", "posterior", "credible", "hypothesis",
    "null", "alternative", "threshold", "lag", "lags", "lagged", "fixed", "random",
    "causal", "causality", "qualitative", "grounded", "thematic", "coding",
    "codes", "code", "theme", "themes", "category", "categories", "saturation", "triangulation",
    "reflexivity", "positionality", "intersubjectivity", "hermeneutic", "phenomenological", "phenomenology", "interpretive", "constructivist",
    "analysis", "analytic", "analytical", "longitudinal", "cross-sectional", "multilevel", "multivariable", "hierarchical",
    "meta-analytic", "scoping", "realist", "synthesis", "syntheses", "mixed", "measurement", "variable",
    "variables", "dependent", "independent", "control", "controls", "controlled", "controlling", "proxy",
    "proxies", "check", "checks", "sensitivity", "validation", "replication", "comparative",
    
    # 3. Abstract structure boilerplate
    # Organisational scaffolding recurring formulaically in academic abstracts:
    # document labels, rhetorical-move markers, literature-positioning words,
    # linking tokens, and non-discriminative evaluative fillers.
    "study", "studies", "paper", "papers", "article", "articles", "research", "work",
    "works", "thesis", "dissertation", "chapter", "chapters", "book", "books", "manuscript",
    "manuscripts", "introduction", "background", "motivation", "objective", "objectives", "purpose", "purposes",
    "goal", "goals", "rationale", "overview", "summary", "abstract", "conclusion", "conclusions",
    "discussion", "implication", "implications", "contribution", "contributions", "limitation", "limitations", "recommendation",
    "recommendations", "direction", "directions", "findings", "result", "results", "outcome", "outcomes",
    "literature", "prior", "previous", "existing", "recent", "current", "past", "earlier",
    "contemporary", "traditional", "conventional", "growing", "emerging", "nascent", "based", "using",
    "used", "given", "among", "across", "within", "via", "through", "including",
    "particularly", "specifically", "notably", "especially", "primarily", "mainly", "largely", "generally",
    "broadly", "widely", "commonly", "typically", "often", "frequently", "usually", "rarely",
    "however", "therefore", "thus", "hence", "whereas", "whereby", "therein", "furthermore",
    "moreover", "additionally", "nevertheless", "nonetheless", "accordingly", "consequently", "subsequently", "previously",
    "finally", "overall", "together", "concluded", "concludes", "derives", "derived", "conducted",
    "collected", "published", "carried", "undertaken", "outlined", "important", "significant", "key",
    "main", "major", "primary", "central", "core", "fundamental", "essential", "critical",
    "crucial", "vital", "notable", "relevant", "valuable", "meaningful", "useful", "effective",
    "novel", "innovative", "original", "unique", "comprehensive", "systematic", "rigorous", "robust",
    "substantial", "strong", "potential", "possible", "likely", "apparent", "clear", "direct",
    "indirect", "further", "additional", "various", "diverse", "multiple", "different", "similar",
    "comparable", "specific", "particular", "general", "broad", "wide", "limited", "due",
    "despite", "although", "though", "while", "whether", "both", "either", "neither",
    
    # 4. Generic social science concepts
    # Terms so broadly used in social science that they provide no discriminative
    # topical signal: generic agents, process/mechanism words, structural terms,
    # effect/outcome nouns, generic descriptors, and temporal/spatial terms.
    "society", "societies", "social", "individual", "individuals", "group", "groups", "population",
    "populations", "people", "person", "persons", "human", "humans", "actor", "actors",
    "agent", "agents", "member", "members", "community", "communities", "public", "citizen",
    "citizens", "subject", "subjects", "user", "users", "stakeholder", "stakeholders", "researcher",
    "researchers", "author", "authors", "scholar", "scholars", "behavior", "behaviour", "behaviors",
    "behaviours", "attitude", "attitudes", "perception", "perceptions", "belief", "beliefs", "preference",
    "preferences", "value", "values", "norm", "norms", "practice", "practices", "experience",
    "experiences", "interaction", "interactions", "process", "processes", "mechanism", "mechanisms", "dynamic",
    "dynamics", "pattern", "patterns", "relationship", "relationships", "relation", "relations", "association",
    "associations", "link", "links", "connection", "connections", "nexus", "interplay", "structure",
    "structures", "system", "systems", "network", "networks", "institution", "institutions", "organization",
    "organizations", "organisation", "organisations", "context", "contexts", "condition", "conditions", "environment",
    "environments", "setting", "settings", "domain", "domains", "sphere", "field", "fields",
    "sector", "sectors", "dimension", "dimensions", "aspect", "aspects", "element", "elements",
    "component", "components", "feature", "features", "factor", "factors", "effect", "effects",
    "impact", "impacts", "influence", "influences", "consequence", "consequences", "change", "changes",
    "difference", "differences", "variation", "variations", "increase", "increases", "decrease", "decreases",
    "reduction", "reductions", "improvement", "improvements", "development", "developments", "growth", "decline",
    "shift", "shifts", "transition", "transitions", "transformation", "transformations", "role", "roles",
    "level", "levels", "degree", "degrees", "extent", "scope", "range", "type",
    "types", "form", "forms", "kind", "kinds", "way", "ways", "modes",
    "nature", "perspective", "perspectives", "view", "views", "lens", "time", "period",
    "periods", "stage", "stages", "phase", "phases", "era", "eras", "decade",
    "decades", "year", "years", "month", "months", "day", "days", "long",
    "short", "term", "terms", "run", "world", "global", "international", "national",
    "local", "regional", "urban", "rural", "cross", "domestic", "foreign", "western",
    "eastern", "northern", "southern",
    
    # 5. Journal specific and publication boilerplate
    # Words introduced by journal formatting, editorial conventions, copyright
    # notices, metadata fragments, and abstract-template boilerplate specific
    # to these six journals.
    "copyright", "rights", "reserved", "license", "licensed", "licence", "licenced", "attribution",
    "commons", "creative", "reproduced", "reproduction", "permission", "permissions", "prohibited", "doi",
    "http", "https", "www", "org", "com", "net", "edu", "vol",
    "volume", "issue", "number", "pp", "page", "pages", "journal", "springer",
    "oxford", "cambridge", "wiley", "elsevier", "taylor", "francis", "routledge", "sage",
    "plos", "arxiv", "preprint", "open", "access", "publication", "publications", "publish",
    "publisher", "publishing", "editor", "editors", "editorial", "peer", "reviewed", "accepted",
    "submission", "submitted", "revised", "revision", "correspondence", "email", "orcid", "affiliation",
    "affiliations", "department", "university", "universities", "keywords", "keyword", "jel",
    "classification", "palgrave", "collection", "published", "cambridge.org", "doi.org", "10.1093",
    "s41599", "10.1057",
    
    # 6. Encoding artefacts and noise
    # Tokens from PDF/HTML extraction, encoding issues, OCR, or tokenisation:
    # Latin abbreviations, single characters, URL fragments, Greek letter tokens,
    # ligature remnants, and other non-linguistic noise.
    "i.e", "e.g", "et", "al", "etc", "ibid", "cf", "vs",
    "viz", "nb", "fig", "figs", "tab", "tabs", "eq", "eqs",
    "pg", "n.d", "n.a", "na", "nd", "a", "b", "c",
    "d", "e", "f", "g", "h", "i", "j", "k",
    "l", "m", "n", "o", "p", "q", "r", "s",
    "t", "u", "v", "w", "x", "y", "z", "oe",
    "ae", "fi", "ff", "ffi", "fl", "ffl", "html", "pdf",
    "xml", "json", "url", "uri", "beta", "alpha", "gamma", "delta",
    "sigma", "omega", "lambda", "mu", "nu", "rho", "phi", "chi",
    "eta", "kappa", "tau", "theta", "epsilon", "zeta", "table", "appendix",
    "supplementary", "supplement", "support", "supporting", "information", "available", "request",
    
    # 7. High-frequency functions
    # Discourse connectives, hedging adverbs, academic filler adjectives,
    # numbers-as-words, and filler nouns that standard lists (Snowball, SMART,
    # onix) miss but which carry no thematic signal in this corpus.
    "alternatively", "conversely", "meanwhile", "simultaneously", "otherwise", "namely", "predominantly", "chiefly",
    "mostly", "jointly", "collectively", "respectively", "comparatively", "relatively", "apparently", "seemingly",
    "arguably", "potentially", "possibly", "unlikely", "presumably", "ostensibly", "admittedly", "importantly",
    "interestingly", "surprisingly", "crucially", "significantly", "substantially", "considerably", "markedly", "increasingly",
    "positively", "negatively", "directly", "indirectly", "highly", "strongly", "effectively", "consistently",
    "robustly", "empirically", "theoretically", "conceptually", "practically", "systematically", "comprehensively", "rigorously",
    "explicitly", "implicitly", "dominant", "prominent", "salient", "unprecedented", "thorough", "considerable",
    "efficient", "successful", "increasing", "rising", "expanding", "deepening", "widening", "ongoing",
    "continuing", "persistent", "enduring", "modest", "marginal", "partial", "complex", "complicated",
    "nuanced", "multifaceted", "multidimensional", "interdisciplinary", "heterogeneous", "homogeneous", "distinct", "varied",
    "consistent", "inconsistent", "contradictory", "complementary", "broader", "deeper", "wider", "stronger",
    "weaker", "higher", "lower", "greater", "lesser", "larger", "smaller", "one",
    "two", "three", "four", "five", "six", "seven", "eight", "nine",
    "ten", "first", "second", "third", "fourth", "fifth", "half", "quarter",
    "percent", "percentage", "proportion", "share", "ratio", "rate", "insight", "insights",
    "understandings", "knowledge", "advancement", "progress", "step", "steps", "effort", "efforts",
    "attempt", "attempts", "need", "needs", "opportunity", "opportunities", "gap", "gaps",
    "concern", "concerns", "issues", "debate", "debates", "discussions", "attention", "lack",
    "absence", "presence", "existence"
    
    # 8. Other
    # Other words we found out needed to be filter out while working on the
    # project.
    
    
    
  ),
  category = c(
    
    # 1. Academic boilerplate verbs
    # Verbs used generically across all abstracts regardless of topic.
    # Includes base forms, 3rd-person singular, past/passive, and gerund inflections.
    rep("boilerplate_verb", 310),
    
    # 2. Generic methodological terms
    # Words describing research designs, data types, statistical procedures,
    # and analytical approaches ubiquitous across all social science.
    rep("methodological_term", 215),
    
    # 3. Abstract structure boilerplate
    # Organisational scaffolding recurring formulaically in academic abstracts:
    # document labels, rhetorical-move markers, literature-positioning words,
    # linking tokens, and non-discriminative evaluative fillers.
    rep("abstract_boilerplate", 176),
    
    # 4. Generic social science concepts
    # Terms so broadly used in social science that they provide no discriminative
    # topical signal: generic agents, process/mechanism words, structural terms,
    # effect/outcome nouns, generic descriptors, and temporal/spatial terms.
    rep("generic_social_science", 209),
    
    # 5. Journal specific and publication boilerplate
    # Words introduced by journal formatting, editorial conventions, copyright
    # notices, metadata fragments, and abstract-template boilerplate specific
    # to these six journals.
    rep("publication_boilerplate", 80),
    
    # 6. Encoding artefacts and noise
    # Tokens from PDF/HTML extraction, encoding issues, OCR, or tokenisation:
    # Latin abbreviations, single characters, URL fragments, Greek letter tokens,
    # ligature remnants, and other non-linguistic noise.
    rep("encoding_artefact", 87),
    
    # 7. High-frequency functions
    # Discourse connectives, hedging adverbs, academic filler adjectives,
    # numbers-as-words, and filler nouns that standard lists (Snowball, SMART,
    # onix) miss but which carry no thematic signal in this corpus.
    rep("domain_function_word", 147),
    
    # 8. Other
    # Other words we found out needed to be filter out while working on the
    # project.
    rep("other", 0)
  )
)

# Filter out domain specific stopwords:
tidy_abstracts <- tidy_abstracts |>
  anti_join(domain_stopwords, by = "word")

# To check that it worked:
tidy_abstracts
```

There are still some words we do not wish to include in the analysis: publication years or years of recent events are often mentioned in abstracts. For example:

```{r}
# Print an abstract with a recent year:
articles |>
     filter(str_detect(abstract, "2021")) |>
     slice(1) |>
     pull(abstract)
```

We filter these out. Other years and numbers could be relevant ---referring to historic events, or significative values---. 

```{r}
# Filter out recent years:
tidy_abstracts <- tidy_abstracts |>
  filter(!word %in% as.character(2015:2025))
```

#### Summary and visualizations of the clean corpus [¿IGUAL ESTO HABRÍA QUE QUITARLO?] ¿realmente es preprocessing?

```{r}
# Total tokens after preprocessing
nrow(tidy_abstracts)

# Unique vocabulary size
n_distinct(tidy_abstracts$word)

# Tokens per journal
tidy_abstracts |>
  count(source_display_name, name = "n_tokens")

# Top 20 most frequent words overall
tidy_abstracts |>
  count(word, sort = TRUE) |>
  slice_head(n = 20)
```

```{r}
# Visualize top words
tidy_abstracts |>
  count(word, sort = TRUE) |>
  slice_head(n = 25) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word)) +
  geom_col(fill = "#2C3E6B", width = 0.7) +
  geom_text(aes(label = scales::comma(n)), 
            hjust = -0.15, size = 3, color = "#2C3E6B", family = "serif") +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = scales::comma
  ) +
  labs(
    title = "Most frequent terms after preprocessing",
    subtitle = "Top 25 tokens across all abstracts (2016–2025)",
    x = "Token frequency",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title       = element_text(size = 14, face = "bold", color = "#1a1a2e",
                                    margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 10, color = "#4a4a6a",
                                    margin = margin(b = 12)),
    plot.caption     = element_text(size = 7, color = "#7a7a9a", hjust = 0,
                                    margin = margin(t = 10)),
    axis.text.y      = element_text(size = 10, color = "#1a1a2e"),
    axis.text.x      = element_text(size = 8,  color = "#4a4a6a"),
    axis.title.x     = element_text(size = 9,  color = "#4a4a6a",
                                    margin = margin(t = 8)),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#e8e8f0", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.background    = element_rect(fill = "white", color = NA),
    plot.margin        = margin(20, 25, 15, 15)
  )
```

The top 25 terms paint a coherent picture of the corpus. The dominance of *policy*, *political*, *economic*, *environmental*, and *education* reflects some recurrent topics in social science journals. The cluster of *health*, *covid*, and *19* points to the influence of the COVID-19 pandemic, one of the main events in our study period. The prominence of *China* and *Chinese* (both in the top 15), and likely reflects China's growing role as a major actor in global affairs, making it a central subject of inquiry across political science, sociology, and economics research during this period. The *education*–*learning*–*students* cluster suggests that educational research is one of the most represented subfields, while words like *digital* and *media*, signal a sustained interest in technology and communication topics throughout the decade.



### Analysis [RQ1 + RQ2]

In order to address RQ1 and RQ2, we will use three complementary techniques: TF-IDF to identify distinctive vocabulary by year (and also for the period before and after the public release of ChatGPT); word frequency analysis using a custom set of AI-related words to measure the growth of AI-related terminology by year; and LDA topic modelling.

#### Distinctive vocabulary by year (TF-IDF)

We use TF-IDF to identify the most distinctive terms of each year, treating each year as a single document. We apply this at three levels: unigram, bigram, and trigram level.

##### Unigram TF-IDF by year

```{r}
# Count how many times each word appears in each year
words_by_year <- tidy_abstracts |>
  count(publication_year, word, sort = TRUE)

# Compute TF-IDF treating each year as a "document"
tfidf_by_year <- words_by_year |>
  bind_tf_idf(word, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_by_year
```

```{r}
#| fig-width: 8
#| fig-height: 12

# Top 8 most distinctive words per year
tfidf_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, word, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive words per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR
- que aparezca gun y guns
- el aniversario de la muerte de shakespeare
- ¿por qué tanto gothic?
- lo más evidente: la ruptura de covid 19
- ai a partir de 2023
Lo suyo sería buscar  lo que significan cosas como: esg, efl, csr, mml, dms, geg, etc. (Y no en google, buscarlo en el contexto de nuestro corpus usando lo de la context window, pero qué perezote). ¿Metemos el filtro que proponía alicia aunque nos estemos cargando "gun", pea", "ai"?
*- ¿Tendríamos que decir algo sobre los valores? ¿Son muy bajos? ¿El hecho de que sean muy bajos tiene consecuencias? ¿Hace que el significado sea poco robusto?*
-> En realidad podemos disfrazarlo escribiendo un parrafillo que mencione un par de anécdotas rollo (vemos el aniversario de shakespeare.. guns... cosas de crisis climática... la influencia de la pandemia... PERO CENTRÁNDONOS EN LO QUE NOS OCUPA: parece claro el interés despertado por la ia a partir del public release de chatgpt... con la palabra ia, chatgpt, digitalization)

##### Bigram TF-IDF by year

For bigrams we tokenize from the raw abstracts, we separate bigrams into their component words, filter standard stopwords from each position, and reunite them before computing TF-IDF. We do not filter by the full domain-specific stopwords here: words like "social" or "people" do not have informative value by themselves, but can be very meaningful in bigrams like "social media" or "Roma people". We do filter out abstract struture terms and metadata noise.

```{r}
# We subset abstract category from our domain-specific stopwords tibble,
# we will need this to get substantive results in TF-IDF and not noise
# when using bigrams. We also exclude recent years, since those add noise.
ngram_stopwords <- domain_stopwords |>
  filter(category %in% c("abstract_boilerplate",
                         "boilerplate_verb",
                         "publication_boilerplate",
                         "encoding_artefact"))
year_noise <- tibble(word = as.character(2015:2025),
                     category = "year_noise")
ngram_stopwords <- bind_rows(ngram_stopwords, year_noise)


# Tokenize into bigrams from raw abstracts
bigrams_by_year <- articles |>
  select(id, publication_year, abstract) |>
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  # Separate into two columns to filter stopwords per position
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    !word1 %in% ngram_stopwords$word,
    !word2 %in% ngram_stopwords$word
  ) |>
  # Reunite into a single bigram column
  unite(bigram, word1, word2, sep = " ") |>
  # Count by year
  count(publication_year, bigram, sort = TRUE)

bigrams_by_year
```

```{r}
# Compute TF-IDF on bigrams, treating each year as a "document"
tfidf_bigrams_by_year <- bigrams_by_year |>
  bind_tf_idf(bigram, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_bigrams_by_year
```

```{r}
#| fig-width: 9
#| fig-height: 12

# Top 8 most distinctive bigrams per year
tfidf_bigrams_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(bigram = reorder_within(bigram, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, bigram, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive bigrams per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```
INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETARINTERPRETAR INTERPRETAR INTERPRETAR (algunas ideas)
- Algo evidente: la ruptura de "covid 19", que pasa a estar en el top 2 en los últimos cinco años;
- Algo relacionado con nuestro tema: digitial economy, digital transformation
- Climate change related: carbon emissions, green innovation, green consumption, crude oil, green finance
*- ¿Tendríamos que decir algo sobre los valores? ¿Son muy bajos? ¿El hecho de que sean muy bajos tiene consecuencias? ¿Hace que el significado sea poco robusto?*


##### Trigram TF-IDF by year

```{r}
# Same workflow for trigrams: tokenize, separate, filter, reunite
trigrams_by_year <- articles |>
  select(id, publication_year, abstract) |>
  unnest_tokens(trigram, abstract, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    !word3 %in% stop_words$word,      # <--- Check 1: Standard list
    !word1 %in% ngram_stopwords$word,
    !word2 %in% ngram_stopwords$word,
    !word3 %in% ngram_stopwords$word   # <--- Check 2: Your custom list
  ) |>
  unite(trigram, word1, word2, word3, sep = " ") |>
  count(publication_year, trigram, sort = TRUE)

trigrams_by_year
```

```{r}
# Compute TF-IDF on trigrams
tfidf_trigrams_by_year <- trigrams_by_year |>
  bind_tf_idf(trigram, publication_year, n) |>
  arrange(desc(tf_idf))

tfidf_trigrams_by_year
```

```{r}
#| fig-width: 10
#| fig-height: 12

# Top 8 most distinctive trigrams per year
tfidf_trigrams_by_year |>
  group_by(publication_year) |>
  slice_max(tf_idf, n = 8) |>
  ungroup() |>
  mutate(trigram = reorder_within(trigram, tf_idf, publication_year)) |>
  ggplot(aes(tf_idf, trigram, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Most distinctive trigrams per year (TF-IDF)",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text  = element_text(face = "bold"),
    axis.text.y = element_text(size = 7)
  )
```

INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR INTERPRETAR
- De nuevo: evidente covid 19 pandemic
- ecological crisis:  ecological civilization construction, regional carbon efficiency, extreme weather impacts, farmers sustainable likelihood, green development behavior...
- artificial intelligence aparece en  2024 (un valor super bajo)
- Otros relacionados con nuestra movida (enterprise digital transformation, )
*- ¿Tendríamos que decir algo sobre los valores? ¿Son muy bajos? ¿El hecho de que sean muy bajos tiene consecuencias? ¿Hace que el significado sea poco robusto?*
- NOTA DEL CLAUDE: Trigrams are sparser than bigrams, so some years may have very few distinctive entries — that itself is informative.

#### AI as an emerging topic using an AI-related lexicon

In this subsection we operationalise AI as a topic through a keyword lexicon built from the works of Liu, Shapira, and Yue (2021) and Ding, Lawson, and Shapira (2025). These works ---which fully cover our study period--- explore bibliometric search strategies for identifying AI-related publications, the lexicons they build well thought out and validated by their research, and can prove useful for our purposes. We built three lexicons from these sources: a unigrams, a bigrams, and a trigram lexicon.

##### Building the AI lexicon

We manually excluded a significant number of n-grams from these lexicons, prioritising precision over recall. That is, we accept that our lexicon will capture fewer AI-related abstracts, but we have higher confidence that each match is capturing a genuine AI-related article. We excluded terms like "random forest" or "decision tree" because they refer to statistical methods often used in social science and not necessarily related to AI; to keep our lexicon easily manageable, we also excluded terms that are too technical and unlikely to appear in social science journal abstracts; and we excluded ambiguous terms, such as "ai" which Liu et al. (2021, p. 3171) reject for being an abbreviation with up to 164 meanings ---and a hit ratio of just 17.8%---. 

In addition, we included extra unigrams, bigrams, and trigrams to capture named generative AI models and providers absent from both sources, such as GPT-4, Gemini, Claude, or LLama ---while the names by themselves could lead to ambiguity (the zodiac sign, Claude Lévi-Strauss, the animal, etc.), including the version in bigrams solves this ambiguity---.

While not perfect, this approach is more than sufficient to capture broad trends necessary to answer our research questions.

```{r}
# AI unigram lexicon
ai_lexicon_unigrams <- c(
  # Liu et al. (2021)
  "autoencoder",
  "perceptron",
  "neurocomputing",
  # Ding et al. (2025)
  "chatgpt",
  "llm",
  # Custom
  "deepseek",
  "openai"
)

# AI bigram lexicon
ai_lexicon_bigrams <- c(
  # Liu et al. (2021)
  "artificial intelligence",
  "neural network",
  "machine learning",
  "deep learning",
  "reinforcement learning",
  "learning algorithm",
  "backpropagation learning",
  "backpropagation algorithm",
  "machine intelligence",
  "computational intelligence",
  "neural model",
  "deep network",
  # Ding et al. (2025)
  "generative ai",
  "language model",
  "pretrained transformer",
  "ai-generated content",
  "openai gpt",
  # Custom
  "claude sonnet",
  "claude opus",
  "claude haiku",
  "google gemini",
  "gemini pro",
  "gemini ultra",
  "microsoft copilot",
  "github copilot",
  "mistral ai",
  "meta llama",
  "llama 2",
  "llama 3",
  "stable diffusion",
  "gpt 5",
  "gpt 4",
  "gpt 3",
  "gpt 2",
  "gpt 1",
  "dall e",
  # Custom
  "agentic ai"
)

# AI trigram lexicon
ai_lexicon_trigrams <- c(
  # Ding et al. (2025)
  "large language model",
  "generative language model",
  "generative large language",
  "generative pretrained transformer",
  "generative artificial intelligence",
  "artificial intelligence generated",
  "openai large language",
  # Custom
  "openai language model",
  "ai language model",
  "ai generated content"
)
```

##### Tokenizing abstracts at each n-gram level

Since a unigram, bigram, and trigram lexicon, we need to tokenize the abstracts separately for each, and then match against the corresponding lexicon. After that, we combine all matches in single tibble.

```{r}
# Source abstracts. We start from articles each time since n-gram tokenization
# needs the original text
abstracts_base <- articles |>
  select(id, publication_year, abstract)

# Unigram matches: tokenize at word level and match against unigram lexicon
ai_hits_unigrams <- abstracts_base |>
  unnest_tokens(word, abstract, token = "words") |>
  filter(word %in% ai_lexicon_unigrams) |>
  rename(term = word) |>
  mutate(ngram_level = "unigram")

# Bigram matches: tokenize at bigram level and match against bigram lexicon
ai_hits_bigrams <- abstracts_base |>
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2) |>
  filter(!is.na(bigram)) |>
  filter(bigram %in% ai_lexicon_bigrams) |>
  rename(term = bigram) |>
  mutate(ngram_level = "bigram")

# Trigram matches — tokenize at trigram level and match against trigram lexicon
ai_hits_trigrams <- abstracts_base |>
  unnest_tokens(trigram, abstract, token = "ngrams", n = 3) |>
  filter(!is.na(trigram)) |>
  filter(trigram %in% ai_lexicon_trigrams) |>
  rename(term = trigram) |>
  mutate(ngram_level = "trigram")

# Combine all matches into a single dataframe
ai_hits <- bind_rows(ai_hits_unigrams, ai_hits_bigrams, ai_hits_trigrams)
ai_hits
```

##### AI terms frequency over time

```{r}
# Count unique articles containing at least one AI related term per year
ai_articles_by_year <- ai_hits |>
  distinct(id, publication_year) |>
  count(publication_year, name = "articles_with_ai")

# Total number of abstracts per year
abstracts_by_year <- articles |>
  filter(!is.na(abstract), publication_year >= 2016, publication_year <= 2025) |>
  count(publication_year, name = "total_abstracts")

# Compute proportion
ai_proportion <- ai_articles_by_year |>
  left_join(abstracts_by_year, by = "publication_year") |>
  mutate(proportion_ai_articles = articles_with_ai / total_abstracts)



library(scales) # Needed for percent_format()

# Find the maximum proportion to place the annotation dynamically
max_prop <- max(ai_proportion$proportion_ai_articles, na.rm = TRUE)

ggplot(ai_proportion, aes(x = publication_year, y = proportion_ai_articles)) +
  # Use geom_col for a bar plot instead of geom_line
  geom_col(fill = "#2C3E6B", width = 0.7) +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + 
  labs(
    title = "AI-related terminology in social science abstracts",
    subtitle = "Proportion of articles containing at least one AI term, by year",
    x = "Publication year",
    y = "Proportion of articles (%)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    plot.title         = element_text(size = 14, face = "bold", color = "#1a1a2e"),
    plot.subtitle      = element_text(size = 10, color = "#4a4a6a"),
    axis.text          = element_text(color = "#4a4a6a"),
    panel.grid.major.x = element_blank(), # Removes vertical grid lines for cleaner bars
    panel.grid.minor.x = element_blank()
  )
```
- Cambio muy claro, en 2017 y 2018 prácticamente nada. En 2019 sube al 1%, el año en que se hace público chatgpt pasa hasta el 4%, y tres años después un 9% de los abstracts continen términos relacionados con IA. (Recordemos que estamos empleando un léxico conservador, seguramente hay más menciones y no estamos capturando todo, pero el patrón es evidente). 
- En relación con nuestra RQ: influencia innegable de la IA.


#### Which terms are driving the trend?

We observe a clear trend, let's explore which terms are driving it.

```{r}
#| fig-width: 6
#| fig-height: 8

# Top AI terms per year by absolute article count
ai_hits |>
  # Keep only unique combinations of article, year, and term
  distinct(id, publication_year, term) |>
  # Count the absolute number of articles per term per year
  count(publication_year, term, name = "article_count") |>
  # Group by publication year and get most frequent terms
  group_by(publication_year) |>
  slice_max(article_count, n = 5, with_ties = FALSE) |>
  # Assign a rank (1 to 5) to the top terms
  mutate(rank = row_number()) |> 
  ungroup() |>
  # Add 0 value to missing ranks so every year has exactly ranks 1 through 5
  # and all bars have the same width
  complete(publication_year, rank = 1:5, fill = list(article_count = 0)) |>
  # Create invisible unique names for the blank bars using spaces
  mutate(
    term = ifelse(is.na(term), strrep(" ", rank), term),
    term = reorder_within(term, -rank, publication_year) 
  ) |>
  
  # Plot it
  ggplot(aes(x = article_count, y = term, fill = factor(publication_year))) +
  geom_col(show.legend = FALSE, fill = "#2C3E6B") +
  facet_wrap(~ publication_year, scales = "free", ncol = 2) +
  scale_y_reordered() +
  labs(
    title = "Top AI-related terms per year",
    subtitle = "Absolute number of articles containing each term",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    strip.text         = element_text(face = "bold", size = 10, color = "#1a1a2e"),
    axis.text.y        = element_text(size = 9, color = "#4a4a6a"),
    axis.text.x        = element_text(color = "#4a4a6a"),
    panel.grid.major.y = element_blank() 
  )
```
- Before 2020 not even 5 AI related words from our lexicon.
- Artificial intelligence and machine learning the main drivers since they appear in 2020.
- Worth mentioning: the  appearance of chatgpt from 2023 onwards. Also "generative ai in 2024. May point to an increased (and likely) interest in those topics after the raise of generative llms.


*RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ**RETOMO DESDE AQUÍ*


## Section 4. WRITING STYLE AND THE RISE OF AI \[RQ3\]

## Section 5. SENTIMENT TOWARDS AI IN SOCIAL SCIENCE ABSTRACTS \[RQ4\]

## Section 6. IF WE CAN \[RQ5\]

## Section 7. Conclussions and limitations

## Section 8. References

Ding, L., Lawson, C., & Shapira, P. (2025). Rise of generative artificial intelligence in science. *Scientometrics*, 130(9), 5093–5114. https://doi.org/10.1007/s11192-025-05413-z

Liu, N., Shapira, P., & Yue, X. (2021). Tracking developments in artificial intelligence research: Constructing and applying a new search strategy. *Scientometrics*, 126(4), 3153–3192. https://doi.org/10.1007/s11192-021-03868-4


## Appendix I - Domain specific stopword list prompt

The domain-specific stopword list used in Section 3 was generated with the assistance of a large language model (Claude Sonnet 4.6, Anthropic). The complete prompt is reproduced below.

::: {style="border: 1px solid #ccc; padding: 1em; border-radius: 4px;"}
```
You are an expert computational linguist and text mining specialist with deep 
knowledge of social science research and academic writing conventions. You are 
working on a large-scale corpus analysis project described in the context below.

<project_context>
The project analyzes a corpus of journal abstracts drawn from six leading social 
science journals over the period 2016–2025:

  - Nature Human Behaviour (behavioral science, interdisciplinary)
  - Humanities and Social Sciences Communications (interdisciplinary)
  - American Sociological Review (sociology, North America)
  - American Political Science Review (political science, North America)
  - European Sociological Review (sociology, Europe)
  - European Journal of Political Research (political science, Europe)

The corpus spans the period before and after the public rise of generative AI 
(November 2022, release of ChatGPT). Section 3 of the project (for which these 
stopwords are being prepared) addresses two research questions:

  RQ1: What are the predominant topics in top social science journals, by journal 
       and by year?
  RQ2: Has AI (in the broad sense) become one of the predominant topics, and has 
       it influenced the broader topical landscape?

The unit of analysis is the abstract (short, dense, highly formulaic texts 
averaging 150–250 words), written in academic English.
</project_context>

<corpus_frequency_data>
The file corpus_word_freq.csv is attached. It contains the top words in the corpus 
after removing standard English stopwords  and  numeric tokens, generated directly
from thousands of abstracts using tidytext. Columns:
  - word: the token (lowercase)
  - n: raw frequency count across all abstracts

Use this frequency data as empirical grounding for your stopword list:
  - Words that appear here at high frequency AND carry no thematic/discriminative 
    value are strong candidates for inclusion in the stopword list
  - Do not limit yourself to words in this list — be exhaustive across all 7 
    categories — but treat high-frequency words as a priority checklist to 
    ensure nothing obvious is missed
  - High-frequency words that carry topical signal (per the preservation rules 
    below) must NOT be added to the stopword list even if they rank highly
    (e.g. "policy", "political", "education", "health")
</corpus_frequency_data>

<task>
Generate a comprehensive, categorised list of domain-specific stopwords to be 
removed from this corpus. These are words that:

  (a) carry no thematic or discriminative value for social science research AND
  (b) appear at high frequency across the corpus, AND

Before generating each category, cross-check it against the corpus_frequency_data 
above to ensure all high-frequency uninformative words are captured.

Return the list as a clean, categorised R tibble with a column for the stopword 
and a column for its category, preceded by commented section headers explaining 
the rationale for each category.
</task>

<categories_to_cover>
Generate stopwords across ALL of the following categories. For each category, think 
systematically about what words in that class appear in social science abstracts, words like:

1. ACADEMIC BOILERPLATE VERBS
   Verbs used generically across all abstracts regardless of topic: reporting verbs 
   ("examine", "investigate", "explore", "suggest", "find", "show", "demonstrate", 
   "analyze", "assess", "evaluate", "discuss", "argue", "propose", "address", 
   "consider", "focus", "aim", "seek", "test", "draw", "provide", "offer", 
   "highlight", "reveal", "indicate", "identify", "contribute", "develop"), 
   and their inflected forms where relevant.

2. GENERIC METHODOLOGICAL TERMS
   Words describing research designs, data types, and analytical approaches that 
   are ubiquitous across social science and carry no topical signal: "survey", 
   "questionnaire", "interview", "sample", "dataset",  "longitudinal", "cross-sectional",
   "regression", "model", "framework",  "approach", "method", "methodology",
   "analysis", "data", "measure", "variable", "coefficient", "estimate", "robustness",
   "validity", "reliability", etc.

3. ABSTRACT-STRUCTURAL BOILERPLATE
   Phrases and words used to signal the structure of an abstract itself rather than 
   its content: "paper", "study", "article", "research", "literature", "review", 
   "background", "objective", "purpose", "contribution", "implication", "finding", 
   "result", "conclusion", "evidence", "prior", "existing", "previous", "recent", 
   "current", "present", "based", "using", "used", "given", "among", "across".

4. GENERIC SOCIAL SCIENCE CONCEPTS
   Terms so broadly used in social science that they provide no discriminative 
   topical signal between documents: "society", "social", "individual", "group", 
   "population", "people", "person", "human", "behavior", "behaviour", "attitude", 
   "perception", "factor", "role", "relationship", "impact", "effect", "influence", 
   "context", "level", "process", "system", "structure", "pattern", "change", 
   "difference", "association", "correlation", etc.

5. JOURNAL-SPECIFIC AND PUBLICATION BOILERPLATE
   Words likely introduced by journal formatting, editorial conventions, or abstract 
   templates specific to these six journals: copyright notices, volume/issue markers, 
   DOI fragments, and any words that appear due to systematic metadata leakage.

6. ENCODING ARTEFACTS AND NOISE
   Tokens that result from text extraction, encoding issues, or tokenisation: 
   single characters (except meaningful ones), numeric tokens, garbled unicode 
   fragments, OCR noise, HTML remnants, and other non-linguistic tokens.

7. HIGH-FREQUENCY FUNCTION WORDS NOT CAUGHT BY STANDARD STOPWORD LISTS
   Domain-specific function-like words that standard English stopword lists 
   (e.g., the Snowball or SMART lists) miss but which are uninformative in this 
   corpus: discipline-specific connectives, hedging phrases reduced to single tokens 
   ("however", "furthermore", "moreover", "nevertheless", "thus", "therefore", 
   "hence", "whereas", "whereby", "therein"), and academic filler adjectives 
   ("important", "significant", "key", "major", "critical", "novel", "robust", 
   "comprehensive", "systematic", "rigorous", "substantial").
</categories_to_cover>

<preservation_rules>
CRITICAL: The following word classes must NOT appear in the stopword list, even if 
they are frequent. Removing them would destroy the thematic signal needed for RQ1 
and RQ2:

  - AI and technology terms: any term related to artificial intelligence, machine 
    learning, large language models, specific model names, or related technology. 
    These are central to RQ2 and must be preserved.
  - Substantive social science topic terms: words indexing real research topics 
    such as "inequality", "democracy", "migration", "climate", "health", "labor", 
    "race", "gender", "education", "voting", "polarization", "trust", "identity", 
    "discrimination", "poverty", "welfare", etc.
  - Sentiment-bearing terms: words with clear positive or negative valence that 
    will be needed for the sentiment analysis in Section 5.
  - Proper nouns and named entities: country names, institutional names, author 
    names — these may carry topical signal.
  - Temporally specific terms: words that may index pre/post ChatGPT content, 
    which are needed for the temporal analyses in Sections 3–6.

When in doubt about whether a word carries signal, err on the side of keeping it.
</preservation_rules>

<output_format>
Return only executable R code. Structure it as follows:

  library(tibble)

  domain_stopwords <- tibble(
    word = c(
      # 1. ACADEMIC BOILERPLATE VERBS
      # These verbs appear in virtually every abstract regardless of topic.
      # Removing them prevents them from dominating topic-word distributions
      # in LDA and collapsing IDF weights in TF-IDF.
      "examine", "investigate", ...

      # 2. GENERIC METHODOLOGICAL TERMS
      # ...

      # [continue for all 7 categories]
    ),
    category = c(
      # mirror the word vector exactly, one category label per word
      "boilerplate_verb", "boilerplate_verb", ...
    )
  )

Rules:
  - All stopwords in lowercase
  - One row per word (not per phrase, we operate on unigrams after tokenisation)
  - The word and category vectors must be exactly the same length
  - Category labels should be short snake_case strings matching the 7 categories
  - Add a final comment line reporting the total count: # Total: N stopwords
  - Be exhaustive, put lots of effort
</output_format>
```
:::

## Appendix II - AÑADIR EL PROMPT DE EVA

::: {style="border: 1px solid #ccc; padding: 1em; border-radius: 4px;"}
```
PONRE EL PROMPT DE EVA AQUÍ
```
:::
